### 2023-07-07 paper/proposal.tex
add proposal

\documentclass[11pt, reqno]{article}
\usepackage[legalpaper, margin=1in]{geometry}

\usepackage[colorlinks, citecolor=blue, urlcolor=blue]{hyperref}
\usepackage{graphicx} % Required for inserting images

\usepackage[backend=biber, style=authoryear]{biblatex}
\addbibresource{references.bib}

\title{Population PME Project Propsal}
\author{Robert Zielinski}
\date{July 12, 2023}

\begin{document}

\maketitle

\section{Motivation}

The PME algorithm is a valuable approach for estimating a low-dimensional parameterization of a manifold in high-dimensional space. The LPME approach extends this to estimate how the embedding map found using PME changes over time. However, both of these methods are limited in the sense that the PME algorithm can only be used for dimension reduction for a single object, and the LPME algorithm can be used for that single object over multiple time points.

There are many cases where it would be useful to allow an estimate from one object to inform an estimate of a similar object, or to compare between estimates of two similar objects. For example, we have focused on the use of the LPME algorithm to estimate the surface of the hippocampus or other subcortical structures from MRI images. The hippocampi of one person are very similar in form to the hippocampi of another person, so it could be useful to allow the parameterization of the hippocampus from one individual to inform the estimation of an embedding map for another person or a larger group of people. This could improve the accuracy of our estimates and allow for consistent parameterizations (meaning that similar parameter values would correspond to similar locations on the hippocampus) across people, which is difficult to achieve given the current approaches.

We may also be interested in using the estimates from the PME and LPME algorithms to compare between similar objects for multiple people. As an example, we can again consider the hippocampus: because people with Alzheimer's disease (AD) tend to experience hippocampal atrophy, we would expect that the hippocampi of someone with AD would have different characteristics than the hippocampi of a cognitively healthy individual who is similar in most other respects. This could be visible at one time point, but the comparison would be even more powerful when viewed over time. Currently, the coefficients of the higher-level spline function estimated by the LPME algorithm would be useful for this purpose, if we had a means of ensuring a consistent parameterization across individuals.

Ultimately, it would be valuable to extend this type of analysis to comparison of multiple groups. For instance, we may be interested in understanding the aggregate change in the hippocampi of cognitively healthy individuals versus that of a group of people with AD. One potentially powerful use case could be in the analysis of clinical trials, where we may be interested in comparing changes in structures between two study groups. A modeling approach that allows for principal manifold estimation over large, potentially heterogeneous groups or populations would address these use cases.

\section{Proposed Approach}

The modeling approach I am considering right now is influenced by \cite{schulamIntegrativeAnalysisUsing2016}.

\section{Toy Data}

We have discussed using the MNIST dataset as a possible example dataset, and I think this would be a good case for it. We could consider trying to find an aggregate estimate of a manifold for the figure "0", where differences in the shape of the character from the group-wide estimate would be handled by the random-effects type of structure described above.

\section{Application}

A clear use case that would extend closely from our work on the LPME algorithm would be to develop estimates of the hippocampi of those in ADNI's healthy control, mild cognitive impairment, and AD groups. This would provide a way to test whether the method is capable of producing meaningful results given a real-world dataset that would closely reflect a potential use case.

\section{Potential Challenges}

Likely the most significant challenge to be addressed in this project will be how computation time will intersect with the ease of estimating the model parameters. Right now, I have seen the most extensive discussion of fitting hierarchical models from a Bayesian perspective, and there are software packages that may allow us to fit the model with relatively minor adjustments. However, given the iterative nature of the algorithm and the size of the datasets under consideration, it is unlikely that Bayesian inference will be a viable option due to time constraints. This will instead require a different approach to fitting the model (likely EM-related), which will require more math and more custom code, with benefits in terms of time that are unclear at the moment.

Another challenge that may arise is the prospect of overparameterization. Because smoothing spline coefficients are considered the model output, we will need to use models that have a very high number of parameters. If we are able to use Bayesian computation to fit the model, then priors on those parameters may help to alleviate this concern. However, in the more likely scenario that we will need to use frequentist approaches for estimation, overparameterization may become a concern. It is unclear to me how we would resolve that issue should it arise. In any case, the high number of parameters in the model will likely make interpretation difficult, although it may be helpful to have a different set of spline coefficients assigned to each category of interest.

\nocite{*}
\printbibliography

\end{document}

### 2023-07-12 paper/proposal.tex
add partial proposed method details

@@
 
-The modeling approach I am considering right now is influenced by \cite{schulamIntegrativeAnalysisUsing2016}.
+The modeling approach I am considering right now is influenced by the Latent Trajectory Model in \cite{schulamIntegrativeAnalysisUsing2016}.
+Essentially, rather than using a single spline function to represent the embedding function, I consider using several spline functions, each representing a different level of the model in question.
+For example, if we were interested in using the ADNI data to model the hippocampus at one time point, we may consider representing the embedding as a sum of three spline functions.
+One spline would represent the population level, another would represent the subgroup level (for instance, whether someone belongs to the cognitively healthy group or the group with AD), and the final spline would represent individual-level variability.
+The number of splines used, and their structure, would ultimately be flexible depending on the situation being modeled.
+However, I believe the general structure could be flexible enough to adapt to a number of distinct scenarios.
+
+To fit the model, it will be necessary to consider two separate scenarios: one where we consider only a single time point, and one where we attempt to model change over time.
+Both cases will largely resemble the PME algorithm.
+We begin with the single time point case.
+
+\begin{itemize}
+  \item Data Reduction
+  \item Initialization: Run PME on data from one observation, then use the estimated manifold to find parameterizations for all other observations.
+  \item Fitting: Iterate between estimating spline coefficients in alignment with the model structure determined for the given problem and updating the parameterization for each observation
+  \item Tuning: For each spline function, select the most appropriate smoothing value
+\end{itemize}
+
+The specific approach for data reduction, as well as how the concentration of points across individuals will be calculated, is currently unclear.
+Additionally, the estimation mechanism for the spline coefficients is also unclear.
+In \cite{schulamIntegrativeAnalysisUsing2016}, the EM algorithm is used to estimate the parameters, so this may be a viable option in this case as well.
 

### 2023-07-12 paper/proposal.tex
Updates from Overleaf
update proposal

@@
 
-\title{Population PME Project Propsal}
+\title{Population PME Project Proposal}
 \author{Robert Zielinski}
@@
 For example, if we were interested in using the ADNI data to model the hippocampus at one time point, we may consider representing the embedding as a sum of three spline functions.
-One spline would represent the population level, another would represent the subgroup level (for instance, whether someone belongs to the cognitively healthy group or the group with AD), and the final spline would represent individual-level variability.
+One spline would represent the population level, another would represent the subgroup level (for instance, whether someone belongs to the cognitively healthy group or the group with AD), with the spline modeling any differences between the population and each subgroup, and the final spline would represent individual-level variability from the subgroup they belong to.
 The number of splines used, and their structure, would ultimately be flexible depending on the situation being modeled.
 However, I believe the general structure could be flexible enough to adapt to a number of distinct scenarios.
+Because the hierarchical structure is over a large number of spline coefficients that each interact with each other, the additive approach will be helpful for keeping the model flexible, and having coefficients clearly assigned to different groups of interest will improve interpretability.
 
@@
 \begin{itemize}
-  \item Data Reduction
   \item Initialization: Run PME on data from one observation, then use the estimated manifold to find parameterizations for all other observations.
+  \item Data Reduction: Run HDMDE on each observation individually.
   \item Fitting: Iterate between estimating spline coefficients in alignment with the model structure determined for the given problem and updating the parameterization for each observation
@@
 
-The specific approach for data reduction, as well as how the concentration of points across individuals will be calculated, is currently unclear.
-Additionally, the estimation mechanism for the spline coefficients is also unclear.
+The estimation mechanism for the spline coefficients is currently unclear.
 In \cite{schulamIntegrativeAnalysisUsing2016}, the EM algorithm is used to estimate the parameters, so this may be a viable option in this case as well.
+
+The situation where we account for longitudinal change becomes more complicated.
 
@@
 We have discussed using the MNIST dataset as a possible example dataset, and I think this would be a good case for it.
-We could consider trying to find an aggregate estimate of a manifold for the figure "0", where differences in the shape of the character from the group-wide estimate would be handled by the random-effects type of structure described above.
+We could consider trying to find an aggregate estimate of a manifold for the figures "0", "1", "3", "5", and "7", where differences in the shape of the character from the group-wide estimate would be handled by the random-effects type of structure described above.
+The model structure in this case would have two levels, one for the population level and one for the individual level variations.
+If we wanted to make the situation more complex, we could also divide images of the figures "2", "4", and possibly "9" into the various methods people use to write them.
+These different writing methods could then represent subgroups, which would be represented by another level in the model structure.
 

### 2023-08-09 paper/main.tex
begin introduction

%\documentclass[12pt]{amsart}
\documentclass[11pt,reqno]{article}
%\usepackage{cases}

%\RequirePackage[numbers]{natbib}
%\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
\RequirePackage{graphicx}%% uncomment this for including figures

\usepackage[
backend=biber,
style=authoryear
]{biblatex}

\addbibresource{references.bib}

\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amscd}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage[perpage,symbol*]{footmisc}
\usepackage{float}
\usepackage{hyperref}
\usepackage{color}  
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}

\usepackage{algorithm2e}

%\usepackage{natbib}
%\bibliographystyle{abbrvnat}
%\setcitestyle{authoryear,open={(},close={)}}
\usepackage{authblk}

\usepackage{csquotes}
\usepackage[english]{babel}

\renewcommand{\baselinestretch}{1.0}
\setlength{\oddsidemargin}{-0.5cm}
\setlength{\evensidemargin}{-0.5cm}
\renewcommand{\topmargin}{-2cm}
\renewcommand{\oddsidemargin}{0mm}
\renewcommand{\evensidemargin}{0mm}
\renewcommand{\textwidth}{180mm}
\renewcommand{\textheight}{240mm}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\T}{\intercal}
\newcommand{\commentout}[1]{}

\newcommand{\kmedit}[1]{{\color{purple}  #1}}
\newcommand{\meng}[1]{{\color{purple} \sf $\clubsuit\clubsuit\clubsuit$ Kun Meng: [#1]}}
\newcommand{\Meng}[1]{\margMa{(Kun Meng) #1}}

\newcommand{\zielinski}[1]{{\color{blue} \sf $\spadesuit\spadesuit\spadesuit$ Rob Zielinski: [#1]}}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

\begin{document}

\title{Hierarchical Principal Manifold Estimation}
\author[1]{Robert Zielinski}
\author[2]{Kun Meng}
\author[1]{Ani Eloyan}
\affil[1]{Department of Biostatistics, Brown University}
\affil[2]{Division of Applied Mathematics, Brown University}

\maketitle

\doublespacing

\section*{Abstract}

\section{Introduction}

Neurodegenerative diseases such as Alzheimer's disease (AD) and Parkinson's disease are highly complex, chronic conditions that exhibit substantial heterogeneity between patients. To enable researchers to identify potential biomarkers of progression for these conditions, large multi-site longitudinal observational studies have been conducted that collect information from a wide range of participants over long durations across several modalities, including neuroimaging [Cite ADNI and PPMI studies]. Neuroimaging data in particular have been shown to be sensitive to developments in the early stages of disease progression, often before diagnosis occurs [Find citation from reading].

While observations from neuroimaging data are typically high-dimensional, it is commonly assumed that the data lie along a low-dimensional manifold [Relevant citations demonstrating this]. Under this assumption, manifold learning algorithms can be used to recover this underlying manifold and parameterize this high-dimensional input data in the learned low-dimensional space. Most commonly-used manifold learning algorithms, including Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), Locally Linear Embedding (\cite{roweisNonlinearDimensionalityReduction2000}), Laplacian Eigenmaps (\cite{belkinLaplacianEigenmapsDimensionality2003}), and Diffusion Maps (\cite{coifmanDiffusionMaps2006}) construct a graph of similarities between observations, which is then used to determine the structure of the manifold. \cite{mengPrincipalManifoldEstimation2021} presents an alternative principal manifold framework for manifold estimation which extends to higher dimensions the principal curve framework introduced in \cite{hastiePrincipalCurves1989}, in which principal curves are defined as one-dimensional curves that pass through the middle of a high-dimensional dataset.

In neuroimaging settings, manifold learning algorithms are most often used to estimate the underlying manifold at the population level, with vectorized image intensities being used as the input data [Citations]. \cite{yueParameterizationWhiteMatter2016} used manifold learning to estimate a low-dimensional parameterization of the corpus callosum at the individual level. However, in many cases, the outcome of interest is framed as a comparison between different groups (e.g. treatment groups in a clinical trial).

\section{Methods}

\section{Simulations}

\section{Applications}

\section{Discussion}

\newpage

\nocite{*}
%\bibliographystyle{plain}
%\bibliography{references}
\printbibliography

\end{document}

### 2023-08-09 paper/main.tex
introduction draft

@@
 \cite{yueParameterizationWhiteMatter2016} used manifold learning to estimate a low-dimensional parameterization of the corpus callosum at the individual level.
-However, in many cases, the outcome of interest is framed as a comparison between different groups (e.g.
+However, in many cases, the outcome of interest is framed as a comparison between different groups of individuals (e.g.
 treatment groups in a clinical trial).
+This goal is particularly relevant in the case of the complex neurodegenerative conditions mentioned above, where the high degree of heterogeneity in patient experiences have led to attempts to identify subgroups of patients based on a range of criteria such as symptom expression, onset, or rate of progression [Citations].
+Fundamentally, if a low dimensional manifold is thought to effectively represent data related to the patient experience of a neurodegenerative condition, and the experience of that condition differs by group, then the low-dimensional representation of that experience should also differ by group.
+
+A method to accurately and efficiently estimate the manifolds underlying data at the group level must take advantage of the grouped structure of observations found in the dataset to which it is to be applied.
+To develop an estimate of the manifold underlying the hippocampi of participants in the ADNI study, it is necessary to recognize the basic nesting structure of the data, in which brain images are nested within participants, which may belong to a latent disease subgroup for which patients tend to have more related symptom experiences or underlying physiological developments, which are in turn nested within study groups based on Alzheimer's disease status, which are finally nested within the greater study population.
+The goal of this article is to develop a manifold learning approach that appropriately accounts for this hierarchical structure, estimates the underlying manifold at each level of the hierarchy, and allows for statistical testing of differences between those low-dimensional manifolds.
+
+There have been several approaches taken to address hierarchical structure in manifold learning problems.
+\cite{bhatiaHierarchicalManifoldLearning2012} introduced a modification of Laplacian eigenmaps to estimate manifolds from recursively subdivided regions of medical images while enforcing similar alignment of the manifolds estimated from related regions.
+\cite{freicheCharacterizingMyocardialIschemia2022} builds on this approach to incorporate multiple imaging modalities and automate the tuning of the algorithm's hyperparameters.
+
+\cite{gaoHierarchicalManifoldLearning2019} proposed a two-step diffusion maps method for hierarchical data with variation over several domains, where diffusion maps is applied first to reduce the images for each individual to $d_1$ dimensions, resulting in individual-specific manifold estimates, then diffusion maps is applied again on the concatenated parameterizations to reduce over the individuals to $d_2$ dimensions, thus estimating a group-level manifold.
+This approach allows the first step of dimension reduction to capture the most important qualities of each image separately before reducing across subjects, but the two step procedure ensures that the manifold estimated at the individual level has no relation to the manifold estimated at the group level, posing problems for the interpretability of the embedding mappings and comparisons of manifolds between levels.
+
+\cite{guerreroGroupconstrainedManifoldLearning2017} noted that the grouping behavior of observations recorded longitudinally from the same individual adversely impacted the creation of the neighborhood graph between data points when estimating a manifold for the entire population.
+The authors proposed a constraint specifying that, a given observation's collection of nearest neighbors may include at most one point from each individual or other specified grouping.
+This encourages a more even representation of the entire space of observations in the neighborhood graph, and may be applied to any manifold learning algorithm that features a neighborhood graph to specify proximities between observations.
+However, it simply mitigates the effects of the nesting structure on a population-level manifold estimate rather than enabling a detailed comparison of manifold estimates for each group.
+
+This article builds on the principal manifold estimation (PME) algorithm introduced in \cite{mengPrincipalManifoldEstimation2021}, which uses an iterative procedure to specify an embedding mapping from the low-dimensional manifold to a high-dimensional space that takes the form of a smoothing spline.
+Drawing from literature in hierarchical and additive models, we replace the smoothing spline in the PME algorithm with an additive smoothing spline model (\cite{gelmanDataAnalysisUsing2007}, \cite{hastieGeneralizedAdditiveModels1990}, \cite{gelmanBayesianDataAnalysis2014}).
+While the specific model structure may change to meet the needs of the application, \cite{brumbackSmoothingSplineModels1998} and \cite{schulamFrameworkIndividualizingPredictions2015a} provide examples of the general model framework used.
+This ultimately yields a nonlinear dimension reduction algorithm that takes advantage of a statistically principled model to address nesting structures present in the input data.
+As it is based on the PME framework, the algorithm outputs an explicit representation of the embedding function, making it possible to both generate new data in high-dimensional space using low-dimensional parameters and map new input data to the estimated manifold.
+Smoothing spline analysis of variance allows the statistical comparison of the estimated manifolds between groups or individuals (\cite{wangMixedEffectsSmoothing1998}).
+
+The following sections are organized as follows: Section 2 reviews additive models, specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper.
+Section 3 applies this method to several simulated datasets using a variety of underlying manifold structures and tests performance on the MNIST dataset.
+Section 4 demonstrates how the model specified in Section 2 may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset, while Section 5 provides a concluding discussion of the proposed algorithm and opportunities for further study.
 

### 2023-08-10 paper/main.tex
additive model review

@@
 
-The following sections are organized as follows: Section 2 reviews additive models, specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper.
-Section 3 applies this method to several simulated datasets using a variety of underlying manifold structures and tests performance on the MNIST dataset.
+The following sections are organized as follows: Section 2 specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper.
+Section 3 applies this method to several simulated datasets using a variety of underlying manifold structures and tests performance on the MNIST dataset of handwritten digits.
 Section 4 demonstrates how the model specified in Section 2 may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset, while Section 5 provides a concluding discussion of the proposed algorithm and opportunities for further study.
@@
 
-\section{Simulations}
+\subsection{Additive Smoothing Spline Model}
 
-\section{Applications}
+We begin this section with a brief review of additive models.
+Consider a regression scenario where output variable $\mathbf{y} = \left(y_1, \dots, y_n\right)^{T}$ are thought to be influenced by covariate vectors $\mathbf{x}_i = \left(x_{i1}, \dots, x_{ip}\right)$.
+In this situation, the standard multiple linear regression model assumes that the dependence of $y_i$ on $\mathbf{x}_i$ takes the linear form
+\[%
+  y_i = \alpha + x_{i1}\beta_1 + \dots + x_{ip}\beta_p + \epsilon_i
+,\]%
+where $\epsilon$ has mean 0 and variance $\sigma^2$.
+Additive models are a direct extension of this model, where the additive structure of the model is retained, but the assumption of linearity is discarded.
+A simple additive model takes the form
+\[%
+  y_i = \alpha + \sum_{j=1}^{p}f_j(X_{ij}) + \epsilon_i
+,\]%
+where again $\epsilon_i$ has mean 0 and variance $\sigma^2$, but the $f_j$s represent arbitrary functions which are commonly smooth.
+While the functions in the model above only take as parameters the value of one covariate, with each covariate being represented in the model once, this does not need to be the case.
+The structure of additive models seeks to balance benefits of interpretability found in additive models with the flexibility of nonlinear, and frequently, nonparametric functional forms.
+Further information can be found in \cite{hastieGeneralizedAdditiveModels1990}.
+
+With this background in place, we turn to the motivating situation of analyzing the subcortical surfaces of ADNI participants.
+
+\subsection{Hierarchical Principal Manifold Estimation}
+
+\section{Application to Simulated and MNIST Data}
+
+\subsection{Application to Simulated Data}
+
+\subsection{Application to MNIST Data}
+
+\section{Application to ADNI Data}
 

### 2023-08-11 paper/main.tex
add model specification

@@
 
-With this background in place, we turn to the motivating situation of analyzing the subcortical surfaces of ADNI participants.
+With this background in place, we turn to the motivating situation of analyzing the subcortical surfaces of ADNI participants using MR images, with a focus on the hippocampus.
+ADNI is a longitudinal observational study, where several imaging modalities are used to record neuroimages from participants over time.
+Participants can generally be considered to belong to three groups based on cognitive state: elderly healthy controls, those displaying symptoms of mild cognitive impairment, and those diagnosed with AD.
+In practice, participants may transition between these categorizations as their symptoms progress.
+However, because patients may be diagnosed with AD at different stages, accurately modeling these transitions is challenging and is beyond the scope of this article.
+For simplicity, we use the groupings recorded at the time each participant was recruited to the study.
+
+Individual participants are typically scheduled for imaging at six month intervals, though variability exists in terms of the recorded intervals.
+Also of note, the numbers of observations contributed by each participant varies, so groups may be unbalanced.
+We denote the total number of participants as $N$, and use $n_i$ to represent the number of images present for each individual.
+Meanwhile, $G$ denotes the number of observed groups under consideration.
+
+Within a segmented structural MRI, the voxels estimated to mark the surface of the hippocampus can be assumed to lie along a two-dimensional manifold that has been wrapped into the three-dimensional image space.
+More generally, we let $d$ denote the intrinsic dimension of the low-dimensional manifold, and $D$ represent the dimension of the high-dimensional input space.
+For the $j$th image from the $i$th individual belonging to group $g$ with $i = 1, \dots, N$, $j = 1, \dots, n_i$, and $g = 1, \dots, G$, we denote the $k$th voxel in the high-dimensional space as $\mathbf{x}_{ijk} \in \mathbb{R}^{3}$, and its associated low-dimensional parameterization as $\mathbf{y}_{ijk} \in \mathbb{R}^2$.
+We let $t_{ij}$ denote the time at which the $j$th image for individual $i$ is recorded.
+The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{ijk})$, which can be modeled as \[%
+  \mathbf{x}_{ijk} = \psi(\mathbf{y}_{ijk}) = f_p(\mathbf{y}_{ijk}, t_{ij}) + f_{g(i)}(\mathbf{y}_{ijk}, t_{ij}) + f_i(\mathbf{y}_{ijk}, t_{ij}) + f_{ij}(\mathbf{y}_{ijk}) + \epsilon
+.\]%
+
+Each function in this additive model represents a different level of nesting in the data, with $f_p(\mathbf{y}, t)$ estimating the embedding at the population level, and $f_g(\mathbf{y}, t)$, $f_i(\mathbf{y}, t)$, and $f_{ij}(\mathbf{y})$ describing deviations from the population estimates specific to the group, individual, and image in question, respectively.
+As estimating longitudinal changes in the underlying manifold may be of interest, $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$ and $f_i(\mathbf{y}, t)$ are each denoted as varying-coefficient models (\cite{hastieVaryingCoefficientModels1993}).
+Thus, the model is capable of estimating separate longitudinal changes for population, group, and individual estimates.
+In cases where longitudinal changes are not of interest, each varying-coefficient model may be constrained to be held constant with respect to time.
+As the $ij$th image is recorded at a single time point, $f_{ij}(\mathbf{y}, t)$ may not be dependent on time.
+We specify $f_{ij}(\mathbf{y})$ as a smoothing spline, while the functions $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$, and $f_i(\mathbf{y}, t)$ are each varying-coefficient smoothing spline models, estimated by using a smoothing spline to smooth over smoothing spline coefficients calculated at each time point available in the data.
+
+As mentioned previously, this proposed additive spline model structure has several benefits.
 

### 2023-08-11 paper/main.tex
bibliography test

@@
   backend=biber,
-style=authoryear
+  style=numeric
 ]{biblatex}
@@
 As mentioned previously, this proposed additive spline model structure has several benefits.
+First, it maintains interpretability at each level of the model.
+It is possible to visually inspect the embedding maps of the population and each group and individual to better understand how they interact.
+Additionally, basing our handling of the data's hierarchical structure on a statistical model allows us to assess statistical properties of the estimated manifold, where less statistically principled approaches make this option difficult.
 

### 2023-08-13 paper/main.tex
add weighted smoothing spline fitting details

@@
   backend=biber,
-  style=numeric
+  style=authoryear
 ]{biblatex}
@@
 More generally, we let $d$ denote the intrinsic dimension of the low-dimensional manifold, and $D$ represent the dimension of the high-dimensional input space.
-For the $j$th image from the $i$th individual belonging to group $g$ with $i = 1, \dots, N$, $j = 1, \dots, n_i$, and $g = 1, \dots, G$, we denote the $k$th voxel in the high-dimensional space as $\mathbf{x}_{ijk} \in \mathbb{R}^{3}$, and its associated low-dimensional parameterization as $\mathbf{y}_{ijk} \in \mathbb{R}^2$.
+For the $j$th image from the $i$th individual belonging to group $g$ with $i = 1, \dots, N$, $j = 1, \dots, n_i$, and $g = 1, \dots, G$, we denote the $k$th voxel in the high-dimensional space as $\mathbf{x}_{ijk} \in \mathbb{R}^{3}$, and its associated low-dimensional parameterization as $\mathbf{y}_{ijk} \in \mathbb{R}^2$ and weight $w_{ijk}$.
 We let $t_{ij}$ denote the time at which the $j$th image for individual $i$ is recorded.
@@
 
+The smoothing spline functions $f_{ij}$ minimize 
+\[%
+  \mathcal{K}_{\lambda}(f_{ij}) = \sum_{l=1}^{D}\left\{\sum_{k=1}^{m_{ij}}w_{ijk}|\mathbf{x}_{ijk, l} - f_{ij, l}(\mathbf{y}_{ijk})|^2 + \lambda\|\nabla^{\bigotimes 2}f_l\|^2\right\}
+,\]%
+where $\nabla^{\bigotimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is given. Thus, $f_{ij}$ takes the form
+\[%
+  f_{ij, l}(\mathbf{y}) = \sum_{k=1}^{m_{ij}} s_{k, l} \times \eta_{4-d}\left(\mathbf{y} - \mathbf{y}_{ijk}\right) + \sum_{p=1}^{d+1} \alpha_{p, l} \times \phi_p(\mathbf{y}), \quad l = 1, 2, \dots, D
+,\]%
+with $\phi_p(\mathbf{y})$ representing the basis of the linear space of polynomials on $\mathbb{R}^{d}$ with degree $\leq 1$ and $\eta_{\nu}(\mathbf{y}) = \|\mathbf{y}\|_{\mathbb{R}^{d}}^{\nu}\log\left(\|\mathbf{y}\|_{\mathbb{R}^{d}}\right)$ when $\|\mathbf{y}\|_{\mathbb{R}^{d}} \neq 0$ and $\nu$ is even, $\eta_{\nu}(\mathbf{y}) = 0$ when $\|\mathbf{y}\|_{\mathbb{R}^{d}} = 0$ and $\nu$ is even, and $\eta_{\nu}(\mathbf{y}) = \|\mathbf{y}\|_{\mathbb{R}^{d}}^{\nu}$ when $\nu$ is odd.
+
+Without accounting for the additive structure of the model, and given $m_{ij}$ input values $\mathbf{y}_{ij}$, weights $\mathbf{w}_{ij}$, and outputs $x_{ij}$, we can estimate $f_{ij}$ by obtaining the values of $s_l \in \mathbb{R}^{m_{ij}}$ and $\alpha_l \in \mathbb{R}^{d + 1}$ for $l = 1, \dots, D$ that minimize $\|\mathbf{W}^{\frac{1}{2}}\left(\mathbf{x}_{ij, l} - \mathbf{E}s_l - \mathbf{R}\alpha_l\right)\|^2_{\mathbb{R}^{m_{ij}}} + \lambda\|\mathbf{E}^{\frac{1}{2}}s_l\|^2_{\mathbb{R}^{m_{ij}}}$ under the constraint that $\mathbf{R}^{T}s_l = 0$, where
+\begin{itemize}
+  \item $\mathbf{E} = (E_{qr})_{1 \leq q, r, \leq m_{ij}}$ is defined by $E_{qr} = \eta_{4-d}(\mathbf{y}_{ijq} - \mathbf{y}_{ijr})$.
+  \item $\mathbf{R} = \left(R_{kp}\right)_{1 \leq k \leq m_{ij}, \ 1 \leq p \leq d + 1}$ is defined by $R_{kp} = \phi_p(\mathbf{y}_{ijk})$.
+  \item $\mathbf{W} = diag(w_{ij1}, \dots, w_{ijm_{ij}})$.
+\end{itemize}
+
+With $m_l$ representing Lagrange multipliers, the solutions for $s_l$ and $\alpha_l$ can be found by solving
+
+\[%
+  \left(\begin{array}{ccc}
+    2\mathbf{EWE} + 2\lambda\mathbf{E} & 2\mathbf{EWR} & \mathbf{R} \\
+    2\mathbf{R}^{T}\mathbf{WE} & 2\mathbf{R}^{T}\mathbf{WR} & \mathbf{0} \\
+    \mathbf{R}^{T} & \mathbf{0} & \mathbf{0}
+  \end{array}\right)\left(
+  \begin{array}{c}
+    s_l \\
+    \alpha_l \\
+    m_l
+  \end{array}
+  \right) = \left(
+  \begin{array}{c}
+    2\mathbf{EW}\mathbf{x}_l \\
+    2\mathbf{R}^{T}\mathbf{W}\mathbf{x}_l \\
+    \mathbf{0}
+  \end{array}
+  \right)
+.\]%
+
 As mentioned previously, this proposed additive spline model structure has several benefits.
@@
 It is possible to visually inspect the embedding maps of the population and each group and individual to better understand how they interact.
+While the model described here is intended to be applicable to the specific situation involving the modeling of data from the ADNI study, this general modeling approach can be conveniently adapted to a variety of scenarios.
 Additionally, basing our handling of the data's hierarchical structure on a statistical model allows us to assess statistical properties of the estimated manifold, where less statistically principled approaches make this option difficult.
+Of particular interest, \cite{wangMixedEffectsSmoothing1998} and \cite{brumbackSmoothingSplineModels1998} explore connections between smoothing spline models and mixed-effects models.
+The smoothing spline analysis of variance proposed by \cite{wangMixedEffectsSmoothing1998} may be applicable to the model structure proposed here, which would enable a detailed decomposition of the variation in manifold structure by level.
+For the time being, this extension is left to future work.
+
+As discussed in \cite{hastieGeneralizedAdditiveModels1990}, the additive model $\psi(\mathbf{y})$ can be estimated using a backfitting procedure, which uses the model's additive nature to iteratively fit each distinct smoothing spline function until convergence.
+Specifically, we use the fact that, under the model,
+
+\[%
+f_p(\mathbf{y}_{ijk}, t_{ij}) = E\left[\mathbf{x}_{ijk} - f_{g(i)}(\mathbf{y}_{ijk}, t_{ij}) - f_{i}(\mathbf{y}_{ijk}, t_{ij}) - f_{ij}(\mathbf{y}_{ijk})\right]
+,\]%
+with a similar result holding for each of the smooth functions making up the additive model. Thus, the backfitting algorithm to fit the proposed model is given below.
+
+\RestyleAlgo{ruled}
+\LinesNumbered
+
+\SetKwComment{Comment}{/* }{ */}
+
+\RestyleAlgo{ruled}
+\LinesNumbered
+
+\zielinski{The following algorithm is currently used as a placeholder while I work out the specifics.}
+
+\zielinski{Try using a ratio of the norm of the difference in model coefficients and the norm of the old model to measure convergence.
+This will set the change in the model coefficients (as measured by the norm of the difference vector/matrix) relative to the magnitude of the model coefficients.}
+
+\begin{algorithm}
+\caption{Backfitting Algorithm for Additive Spline Model}\label{alg:backfitting}
+  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{ijk}\right\}_{i=1, j=1, k=1}^{N, n_i, m_{ij}}$, $d$-dimensional input values $\left\{\mathbf{y}_{ijk}\right\}_{i=1, j=1, k=1}^{N, n_i, m_{ij}}$, and observation times $\left\{t_{ij}\right\}_{i=1, j=1}^{N, n_i}$.}
+  \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{N}$, and $\left\{\hat{f}_{ij}\right\}_{i=1, j=1}^{N, n_i}$, and optimal tuning parameters $\hat{\lambda}_p$, $\hat{\gamma}_p$, $\left\{\hat{\lambda}_g\right\}_{g=1}^{G}$, $\left\{\hat{\gamma}\right\}_{g=1}^{G}$, $\left\{\hat{\lambda}_{i}\right\}_{i=1}^{N}$, $\left\{\hat{\gamma}_{i}\right\}_{i=1}^{N}$, $\left\{\hat{\lambda}_{ij}\right\}_{i=1, j=1}^{N, n_i}$.}
+
+\For{$t = 1, 2, \dots, T$} {
+  Apply HDMDE algorithm with input $\left(\left\{Y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
+}
+
+  Apply Isomap to parameterize $\left\{\mu_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$.
+Formally set $\pi_{f_t(0)}(\mu_{j, t}) \gets r_{j, t}$ for $j = 1, 2, \dots, N_t$, $t = 1, 2, \dots, T$\;
+
+\For{t = 1, 2, \dots, T} {
+  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t$, $\lambda_t^*$, and $\tau_t$\;
+  $\left\{r_{j, t}\right\} \gets \pi_{f_t}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $X_{j, t} \gets f_t(r_{j, t})$ for $j = 1, \dots, N_t$\;
+}
+
+  Let $N = \max(N_t)$, $\left\{r_i^*\right\}_{i=1}^{N}$ be a grid spanning the range of estimated values of $\left\{r_{i, j}\right\}_{i=1, t=1}^{N_t, T}$\;
+
+\For{t = 1, 2, \dots, T} {
+  Set $X_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$\;
+  Compute $f_t^*(r)$ by solving \eqref{eq:13}\;
+  Set $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$\;
+}
+
+  Define $\boldsymbol{\omega}$ by setting $\boldsymbol{\omega}_t = \left[\boldsymbol{s}_t, \boldsymbol{\alpha}_t\right]$\;
+  \For{l = 1, 2, \dots, L} {
+    Compute $g_{\gamma_l}(t)$ by solving \eqref{eq:16}\;
+    \For{t = 1, 2, \dots, T} {
+      Compute $g_{\gamma_l}^{(t)}$ by solving \eqref{eq:16}\;
+    }
+    Estimate $MSD(\gamma_l)$ using \eqref{eq:18}\;
+  }
+  $\gamma^* = \arg\min_{\gamma}MSD(\gamma)$\;
+  $f^*(t, \boldsymbol{r}) = f_{\gamma^*}(t, \boldsymbol{r})$, where the form of $f^*(t, \boldsymbol{r})$ is given in \eqref{eq:17}
+\end{algorithm}
+
+A clear concern with the use of this procedure in this case is the number of functions to be fit, which can quickly increase the computational burden of fitting the models.
+However, the nesting of the different levels of the data helps to limit the number of function combinations that must be fit, helping to keep computational requirements manageable.
+\cite{brumbackSmoothingSplineModels1998} used restricted maximum likelihood estimation to reach an estimate of their proposed additive spline model, which may potentially provide computational improvements.
+However, in the case that the model should be modified to meet the needs of a new dataset, the backfitting algorithm will be more easily adaptable.
 
 \subsection{Hierarchical Principal Manifold Estimation}
+
+With the structure of the additive spline model established, we now incorporate it into the larger manifold learning procedure.
+This procedure is based on the Principal Manifold Estimation (PME) algorithm introduced in \cite{mengPrincipalManifoldEstimation2021}, which iterates between estimating a function to embed observations in $D$-dimensional space and reparameterizing those observations in the $d$-dimensional space to estimate the underlying manifold.
 

### 2023-08-13 paper/main.tex
Updates from Overleaf
add start of varying-coefficient model explanation

@@
 
+The optimal value of $\lambda$ can be found using cross-validation.
+The solutions to this minimization problem suffice for estimating $f_{ij}(\mathbf{y})$, in which the coefficients are not time-varying.
+To reach estimates for the varying-coefficient models, the above minimization problem is solved for the data available within the relevant grouping at each time point.
+These estimates are then used to generate a new set of outcome values $\left\{\mathbf{x}'\right\}$ that correspond to a new set of parameters, $\left\{\mathbf{y}'\right\}$, which are evenly spaced throughout the estimated parameter space.
+The spline function is reestimated $\left\{\mathbf{x}'\right\}$ and $\left\{\mathbf{y}'\right\}$, which serve as common knots that allow the spline coefficients to be compared across time points.
+
+The coefficients found at each time point $t$, $\mathbf{s}_t$ and $\mathbf{\alpha}_t$ are concatenated in matrix $\mathbf{\omega}_t = \left(\mathbf{s}_t^T, \mathbf{\alpha}_t^T\right)$.
+This coefficient matrix is then used as the output value in a cubic smoothing spline function using the time points as input values.
+
 As mentioned previously, this proposed additive spline model structure has several benefits.

### 2023-08-15 paper/main.tex
add beginning of varying-coefficient estimation algorithm

--- 2023-08-13 paper/main.tex
@@
 A simple additive model takes the form
-\[%
-  y_i = \alpha + \sum_{j=1}^{p}f_j(X_{ij}) + \epsilon_i
-,\]%
+\begin{equation}
+  y_i = \alpha + \sum_{j=1}^{p}f_j(X_{ij}) + \epsilon_i, \label{eq:1}
+\end{equation}
 where again $\epsilon_i$ has mean 0 and variance $\sigma^2$, but the $f_j$s represent arbitrary functions which are commonly smooth.
-While the functions in the model above only take as parameters the value of one covariate, with each covariate being represented in the model once, this does not need to be the case.
+While the functions in model \ref{eq:1} only take as parameters the value of one covariate, with each covariate being represented in the model once, this does not need to be the case.
 The structure of additive models seeks to balance benefits of interpretability found in additive models with the flexibility of nonlinear, and frequently, nonparametric functional forms.
@@
 Also of note, the numbers of observations contributed by each participant varies, so groups may be unbalanced.
-We denote the total number of participants as $N$, and use $n_i$ to represent the number of images present for each individual.
-Meanwhile, $G$ denotes the number of observed groups under consideration.
+We use $G$ to denote the number of observed groups under consideration, then denote the number of individuals as $N$, the number of images belonging to the $i$th individual as $n_i$, and the number of voxels from the $j$th image of individual $i$ as $m_{ij}$.
 
@@
 More generally, we let $d$ denote the intrinsic dimension of the low-dimensional manifold, and $D$ represent the dimension of the high-dimensional input space.
-For the $j$th image from the $i$th individual belonging to group $g$ with $i = 1, \dots, N$, $j = 1, \dots, n_i$, and $g = 1, \dots, G$, we denote the $k$th voxel in the high-dimensional space as $\mathbf{x}_{ijk} \in \mathbb{R}^{3}$, and its associated low-dimensional parameterization as $\mathbf{y}_{ijk} \in \mathbb{R}^2$ and weight $w_{ijk}$.
-We let $t_{ij}$ denote the time at which the $j$th image for individual $i$ is recorded.
-The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{ijk})$, which can be modeled as \[%
-  \mathbf{x}_{ijk} = \psi(\mathbf{y}_{ijk}) = f_p(\mathbf{y}_{ijk}, t_{ij}) + f_{g(i)}(\mathbf{y}_{ijk}, t_{ij}) + f_i(\mathbf{y}_{ijk}, t_{ij}) + f_{ij}(\mathbf{y}_{ijk}) + \epsilon
-.\]%
+For the $j$th image from the $i$th individual belonging to group $g$ with $i = 1, \dots, N_g$, $j = 1, \dots, n_i$, and $g = 1, \dots, G$, we denote the $k$th voxel in the $D$-dimensional space as $\mathbf{x}_{ijk} \in \mathbb{R}^{D}$, its associated $d$-dimensional parameterization as $\mathbf{y}_{ijk} \in \mathbb{R}^{d}$, and its weight as $w_{ijk}$.
+We let $t_{ij}$ denote the time at which the $j$th image for individual $i$ is recorded, with all times measured as the difference between the date the image was taken and the date of the participant's baseline visit.
+We use $\mathbf{x}_g$, $\mathbf{x}_i$, $\mathbf{x}_{ij}$, and $\mathbf{x}_t$ to denote the input voxels from all images in the $g$th group, $i$th individual, $ij$th image, and $t$th time point, respectively, and use similar notation for all other relevant terms.
 
+The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{ijk})$, which can be modeled as 
+\begin{equation}
+  \mathbf{x}_{ijk} = \psi(\mathbf{y}_{ijk}) = f_p(\mathbf{y}_{ijk}, t_{ij}) + f_{g(i)}(\mathbf{y}_{ijk}, t_{ij}) + f_i(\mathbf{y}_{ijk}, t_{ij}) + f_{ij}(\mathbf{y}_{ijk}) + \epsilon. \label{eq:2}
+\end{equation}
 Each function in this additive model represents a different level of nesting in the data, with $f_p(\mathbf{y}, t)$ estimating the embedding at the population level, and $f_g(\mathbf{y}, t)$, $f_i(\mathbf{y}, t)$, and $f_{ij}(\mathbf{y})$ describing deviations from the population estimates specific to the group, individual, and image in question, respectively.
-As estimating longitudinal changes in the underlying manifold may be of interest, $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$ and $f_i(\mathbf{y}, t)$ are each denoted as varying-coefficient models (\cite{hastieVaryingCoefficientModels1993}).
+As estimating longitudinal changes in the underlying manifold may be of interest, $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$ and $f_i(\mathbf{y}, t)$ are each denoted as varying-coefficient models, where the coefficients of each function are dependent on time (\cite{hastieVaryingCoefficientModels1993}).
 Thus, the model is capable of estimating separate longitudinal changes for population, group, and individual estimates.
@@
 The smoothing spline functions $f_{ij}$ minimize 
-\[%
-  \mathcal{K}_{\lambda}(f_{ij}) = \sum_{l=1}^{D}\left\{\sum_{k=1}^{m_{ij}}w_{ijk}|\mathbf{x}_{ijk, l} - f_{ij, l}(\mathbf{y}_{ijk})|^2 + \lambda\|\nabla^{\bigotimes 2}f_l\|^2\right\}
-,\]%
-where $\nabla^{\bigotimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is given. Thus, $f_{ij}$ takes the form
-\[%
-  f_{ij, l}(\mathbf{y}) = \sum_{k=1}^{m_{ij}} s_{k, l} \times \eta_{4-d}\left(\mathbf{y} - \mathbf{y}_{ijk}\right) + \sum_{p=1}^{d+1} \alpha_{p, l} \times \phi_p(\mathbf{y}), \quad l = 1, 2, \dots, D
-,\]%
+\begin{equation}
+  \mathcal{K}_{\lambda}(f_{ij}) = \sum_{l=1}^{D}\left\{\sum_{k=1}^{m_{ij}}w_{ijk}|\mathbf{x}_{ijk, l} - f_{ij, l}(\mathbf{y}_{ijk})|^2 + \lambda\|\nabla^{\otimes 2}f_l\|^2\right\}, \label{eq:3}
+\end{equation}
+where $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is given. Thus, $f_{ij}$ takes the form
+\begin{equation}
+  f_{ij, l}(\mathbf{y}) = \sum_{k=1}^{m_{ij}} s_{k, l} \times \eta_{4-d}\left(\mathbf{y} - \mathbf{y}_{ijk}\right) + \sum_{p=1}^{d+1} \alpha_{p, l} \times \phi_p(\mathbf{y}), \quad l = 1, 2, \dots, D, \label{eq:4}
+\end{equation}
 with $\phi_p(\mathbf{y})$ representing the basis of the linear space of polynomials on $\mathbb{R}^{d}$ with degree $\leq 1$ and $\eta_{\nu}(\mathbf{y}) = \|\mathbf{y}\|_{\mathbb{R}^{d}}^{\nu}\log\left(\|\mathbf{y}\|_{\mathbb{R}^{d}}\right)$ when $\|\mathbf{y}\|_{\mathbb{R}^{d}} \neq 0$ and $\nu$ is even, $\eta_{\nu}(\mathbf{y}) = 0$ when $\|\mathbf{y}\|_{\mathbb{R}^{d}} = 0$ and $\nu$ is even, and $\eta_{\nu}(\mathbf{y}) = \|\mathbf{y}\|_{\mathbb{R}^{d}}^{\nu}$ when $\nu$ is odd.
@@
 
-\[%
+\begin{equation}
    \left(\begin{array}{ccc}
@@
   \end{array}
-  \right)
-.\]%
+  \right). \label{eq:5}
+\end{equation}
+The optimal value of $\lambda$ can be found using cross-validation, with the functions being evaluated by the mean squared error, written as
+\begin{equation}
+  MSE(\lambda) = \frac{1}{m_{ij}}\sum_{k=1}^{m_{ij}}\left(\mathbf{x}_{ijk} - f_{ij}(\mathbf{y}_{ijk})\right)^2. \label{eq:6}
+\end{equation}
+
+The solutions to this minimization problem suffice for estimating $f_{ij}(\mathbf{y})$, in which the coefficients are not time-varying.
+To reach estimates for the varying-coefficient models, equation \ref{eq:5} is solved using the data available at each time point within the relevant group.
+To estimate $f_p(\mathbf{y}, t)$, we use $\left\{\mathbf{x}_t\right\}_{t=1}^T$ and $\left\{\mathbf{y}_t\right\}_{t=1}^T$, yielding fitted function $\left\{\hat{f}_t(\mathbf{y})\right\}_{t=1}^T$.
+Using the maximum and minimum values of $\mathbf{y}_t$ in each dimension, a new $d$-dimensional grid of $m' = \text{max}\left(\left\{m_{ij}\right\}_{i=1, j = 1}^{N, N_i}\right)$ knots, $\left\{\mathbf{y}'\right\}$, is generated with the knots being evenly spread throughout the parameter space.
+These parameters serve as common knots that allow the spline coefficients to be compared across time points.
+They are then used to compute a new set of corresponding outcome values $\left\{\mathbf{x}'_t\right\}_{t=1}^{T}$.
+The spline functions are reestimated as $\left\{\hat{f}'_t\right\}_{t=1}^{T}$ using $\left\{\mathbf{x}'_t\right\}_{t=1}^{T}$ and $\mathbf{y}'$.
+These newly generated knots are weighted equally, so an unweighted smoothing spline model is used.
+Following \cite{greenSilverman1994}, with
+
+\begin{itemize}
+  \item $\mathbf{E}' = \left(E'_{qr}\right)_{1 \leq q, r, \leq m'}$ defined by $E'_{qr} = \eta_{4-d}\left(\mathbf{y}'_{q} - \mathbf{y}'_{r}\right)$.
+  \item $\mathbf{R}' = \left(R'_{qp}\right)_{1 \leq q \leq m', \ 1 \leq p \leq d + 1}$ is defined by $R'_{qp} = \phi_p(\mathbf{y}'_{q})$
+\end{itemize}
+the coefficients that minimize $\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right)^{T}\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right) + \gamma\mathbf{s}_t^{T}\mathbf{E}'\mathbf{s}_t$ can be found by solving
+\begin{equation}
+  \left(
+  \begin{array}{cc}
+    \mathbf{E}' + \lambda_{t}\mathbf{I} & \mathbf{R}'^{T} \\
+    \mathbf{R}' & \mathbf{0}
+  \end{array}
+  \right)\left(
+  \begin{array}{c}
+    \mathbf{s}'_t \\
+    \mathbf{\alpha}'_t
+  \end{array}
+  \right) = \left(
+  \begin{array}{c}
+    \mathbf{x}'_t \\
+    \mathbf{0}
+  \end{array}
+  \right). \label{eq:7}
+\end{equation}
+
+The coefficients found at each time point $t$, $\mathbf{s'}_{t}$ and $\mathbf{\alpha'}_{t}$ are concatenated in matrix $\mathbf{\omega}_{t} = \left(\mathbf{s'}_{t}^T, \mathbf{\alpha'}_{t}^T\right)$.
+This coefficient matrix is then used as the output value in a cubic smoothing spline function using the time points as input values.
+This function takes the form
+\begin{equation}
+  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i\|t - t_i\|^{3} + \sum_{p=1}^{2}\tau_p \phi(t)_p. \label{eq:7}
+\end{equation}
+We define matrix $\mathbf{A}$ as $\mathbf{A}_{qr} = \|t_q - t_r\|^{3}$, matrix $\mathbf{T}$ as $\mathbf{T}_{pr} = \phi_p(t_r)$, $\mathbf{\delta} = \left(\delta_1, \dots, \delta_T\right)^{T}$, and $\mathbf{\tau} = \left(\tau_1, \tau_2\right)^{T}$.
+Then, the coefficients of $g_{\gamma}(t)$, given a value of $\gamma$, can be found by solving
+\begin{equation}
+  \left(
+  \begin{array}{cc}
+    \mathbf{A} + \gamma\mathbf{I} & \mathbf{T}^{T} \\
+    \mathbf{T} & \mathbf{0}
+  \end{array}
+  \right)\left(
+  \begin{array}{c}
+    \mathbf{\delta} \\
+    \mathbf{\tau}
+  \end{array}
+  \right) = \left(
+  \begin{array}{c}
+    \mathbf{\omega} \\
+    \mathbf{0}
+  \end{array}
+  \right). \label{eq:8}
+\end{equation}
+
+As with the tuning process for $\lambda$, the optimal value of $\gamma$ can be identified using cross-validation, with either leave-one-out or $k$-folds cross-validation being appropriate depending on the number of time points available.
+With a proper estimate of the optimum $\gamma$, we may denote the coefficients estimated by $g(t)$ as $g(t) = \Omega(t) = \left(\mathbf{s}_{\gamma}(t)^{T}, \mathbf{\alpha}_{\gamma}(t)^{T}\right)^{T}$.
+Therefore, the varying-coefficient models have the following outputs:
+\begin{equation}
+  f(\mathbf{y}, t) = \sum_{k=1}^{m'}\mathbf{s}_{\gamma}(t)_l \eta_{4-d}\left(\|\mathbf{y} - \mathbf{y'}_k\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\gamma}(t)_p \phi_p(\mathbf{y}). \label{eq:9}
+\end{equation}
+The fitting process for the varying-coefficient smoothing spline model is given formally as Algorithm \ref{alg:varying_spline}.
+
+\SetKwComment{Comment}{/* }{ */}
+
+\RestyleAlgo{ruled}
+\LinesNumbered
+
+\begin{algorithm}
+\caption{Varying-Coefficient Smoothing Spline Model}\label{alg:varying_spline}
+  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{ijk}\right\}_{i=1, j = 1, k = 1}^{N, n_i, m_{ij}}$, $d$-dimensional input values $\left\{\mathbf{y}_{ijk}\right\}_{i=1, j=1, k=1}^{N, n_i, m_{ij}}$, weights $\left\{w_{ijk}\right\}_{i=1, j=1, k=1}^{N, n_i, m_{ij}}$, observation times $\left\{t_{ij}\right\}_{i=1, j=1}^{N, n_i}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$.}
+  \KwResult{Analytic formula of $\hat{f}(\mathbf{y}, t)$ and optimal tuning parameter $\hat{\gamma}$.}
+
+\For{$t = 1, 2, \dots, T$} {
+  Set $U_t$ equal to the number of images included in the data at time $t$\;
+  \For{$l = 1, 2, \dots, L$} {
+    Solve equation \ref{eq:5} to estimate $f_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}_t$, $\mathbf{y}_t$, and $w_t$\;
+    \For{$u = 1, 2, \dots, U_t$} {
+      Solve equation \ref{eq:5} to estimate $f^{(-u)}_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}^{(-u)}_t$, $\mathbf{y}^{(-u)}_t$, $w^{(-u)}_t$\;
+      Estimate $MSE(\lambda_{l})_u$
+    }
+  }
+}
+\end{algorithm}
 
@@
 with a similar result holding for each of the smooth functions making up the additive model. Thus, the backfitting algorithm to fit the proposed model is given below.
-
-\RestyleAlgo{ruled}
-\LinesNumbered
-
-\SetKwComment{Comment}{/* }{ */}
-
-\RestyleAlgo{ruled}
-\LinesNumbered
 

### 2023-08-15 paper/main.tex
complete varying-coefficient model algorithm

@@
 \begin{equation}
-  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i\|t - t_i\|^{3} + \sum_{p=1}^{2}\tau_p \phi(t)_p. \label{eq:7}
+  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i\|t - t_i\|^{3} + \sum_{p=1}^{2}\tau_p \phi(t)_p. \label{eq:8}
 \end{equation}
@@
   \end{array}
-  \right). \label{eq:8}
+  \right). \label{eq:9}
 \end{equation}
@@
 \begin{equation}
-  f(\mathbf{y}, t) = \sum_{k=1}^{m'}\mathbf{s}_{\gamma}(t)_l \eta_{4-d}\left(\|\mathbf{y} - \mathbf{y'}_k\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\gamma}(t)_p \phi_p(\mathbf{y}). \label{eq:9}
+  f(\mathbf{y}, t) = \sum_{k=1}^{m'}\mathbf{s}_{\gamma}(t)_l \eta_{4-d}\left(\|\mathbf{y} - \mathbf{y'}_k\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\gamma}(t)_p \phi_p(\mathbf{y}). \label{eq:10}
 \end{equation}
@@
       Solve equation \ref{eq:5} to estimate $f^{(-u)}_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}^{(-u)}_t$, $\mathbf{y}^{(-u)}_t$, $w^{(-u)}_t$\;
-      Estimate $MSE(\lambda_{l})_u$
+      Estimate $MSE(\lambda_{l})_u$ using equation \ref{eq:6}\;
     }
+    Set $MSE(\lambda_l) = \frac{1}{U}\sum_{u=1}^{U}MSE(\lambda_l)_u$\;
   }
+  $\lambda_t = \arg \min_{\lambda} MSE(\lambda)$\;
+  $f_t(\mathbf{y}) = f_{\lambda_t}(\mathbf{y})$\;
 }
+
+  Let $m' = \max(m_{ij})$, $\left\{\mathbf{y}'_{k}\right\}_{k=1}^{m'}$ be a grid spanning the range of values of $\left\{\mathbf{y}\right\}$\;
+  \For{$t = 1, 2, \dots, T$} {
+    Compute $\left\{\mathbf{x'}_{k, t}\right\}_{k=1}^{m'} = \left\{f_t(\mathbf{y'}_k)\right\}$\;
+    Solve equation \ref{eq:7} to compute $\mathbf{s'}_t$ and $\mathbf{\alpha'}_t$\;
+    Set $\mathbf{\omega}_t = \left[\mathbf{s'}_t^{T}, \mathbf{\alpha'}_t^{T}\right]$\;
+  }
+  \For{$p = 1, 2, \dots, P$} {
+    Compute $g_{\gamma_l}(t)$ by solving equation \ref{eq:9}\;
+    \For{$t = 1, 2, \dots, T$} {
+      Compute $g_{\gamma_l}^{(t)}$ by solving \ref{eq:9}\;
+    }
+    Estimate $MSE(\gamma_l)$ using \ref{eq:10}\;
+  }
+  $\hat{\gamma} = \arg \min_{\gamma} MSE(\gamma)$\;
+  $\hat{f}(\mathbf{y}, t) = f_{\hat{\gamma}}(\mathbf{y}, t)$, where the form of $\hat{f}(\mathbf{y}, t)$ is given in \ref{eq:11}
 \end{algorithm}

### 2023-08-15 paper/main.tex
update notation and correct varying-coefficient spline model

--- 2023-08-15 paper/main.tex
@@
 Also of note, the numbers of observations contributed by each participant varies, so groups may be unbalanced.
-We use $G$ to denote the number of observed groups under consideration, then denote the number of individuals as $N$, the number of images belonging to the $i$th individual as $n_i$, and the number of voxels from the $j$th image of individual $i$ as $m_{ij}$.
+We denote the number of groups under consideration using $G$, the number of individuals included in the study as $I$, the number of images available as $N$, and the number of observations available for the $j$th image, with $j = 1, 2, \dots, N$, as $M_j$.
+The individual to which the $j$th image belongs is denoted as $i(j)$, the group of the $i(j)$th individual is represented by $g(j)$, and we use $t_j$ to denote the time at which the $j$th image was recorded, with all times being relative to the $i(j)$th individual's baseline visit.
+Finally, we use $N_t$, $N_i$, and $N_g$ to denote the number of images available for time point $t$, individual $i$, and group $g$, respectively.
 
@@
 More generally, we let $d$ denote the intrinsic dimension of the low-dimensional manifold, and $D$ represent the dimension of the high-dimensional input space.
-For the $j$th image from the $i$th individual belonging to group $g$ with $i = 1, \dots, N_g$, $j = 1, \dots, n_i$, and $g = 1, \dots, G$, we denote the $k$th voxel in the $D$-dimensional space as $\mathbf{x}_{ijk} \in \mathbb{R}^{D}$, its associated $d$-dimensional parameterization as $\mathbf{y}_{ijk} \in \mathbb{R}^{d}$, and its weight as $w_{ijk}$.
-We let $t_{ij}$ denote the time at which the $j$th image for individual $i$ is recorded, with all times measured as the difference between the date the image was taken and the date of the participant's baseline visit.
-We use $\mathbf{x}_g$, $\mathbf{x}_i$, $\mathbf{x}_{ij}$, and $\mathbf{x}_t$ to denote the input voxels from all images in the $g$th group, $i$th individual, $ij$th image, and $t$th time point, respectively, and use similar notation for all other relevant terms.
+For the $j$th image with $j = 1, \dots, N$, we denote the $m$th voxel in the $D$-dimensional space as $\mathbf{x}_{jm} \in \mathbb{R}^{D}$, its associated $d$-dimensional parameterization as $\mathbf{y}_{jm} \in \mathbb{R}^{d}$, and its weight as $w_{jm}$.
+We use $\mathbf{x}_g$, $\mathbf{x}_i$, $\mathbf{x}_{j}$, and $\mathbf{x}_t$ to denote the input voxels from all images in the $g$th group, $i$th individual, $j$th image, and $t$th time point, respectively, and use similar notation for all other relevant terms.
 
 The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{ijk})$, which can be modeled as
-\begin{equation}
-  \mathbf{x}_{ijk} = \psi(\mathbf{y}_{ijk}) = f_p(\mathbf{y}_{ijk}, t_{ij}) + f_{g(i)}(\mathbf{y}_{ijk}, t_{ij}) + f_i(\mathbf{y}_{ijk}, t_{ij}) + f_{ij}(\mathbf{y}_{ijk}) + \epsilon. \label{eq:2}
-\end{equation}
-Each function in this additive model represents a different level of nesting in the data, with $f_p(\mathbf{y}, t)$ estimating the embedding at the population level, and $f_g(\mathbf{y}, t)$, $f_i(\mathbf{y}, t)$, and $f_{ij}(\mathbf{y})$ describing deviations from the population estimates specific to the group, individual, and image in question, respectively.
+
+\[%
+  \mathbf{x}_{jm} = \psi(\mathbf{y}_{jm}) = f_p(\mathbf{y}_{jm}, t_{j}) + f_{g(j)}(\mathbf{y}_{jm}, t_{j}) + f_{i(j)}(\mathbf{y}_{jm}, t_{j}) + f_j(\mathbf{y}_{jm}) + \epsilon_{jm}
+.\]%
+
+Each function in this additive model represents a different level of nesting in the data, with $f_p(\mathbf{y}, t)$ estimating the embedding at the population level, and $f_g(\mathbf{y}, t)$, $f_i(\mathbf{y}, t)$, and $f_{j}(\mathbf{y})$ describing deviations from the population estimates specific to the group, individual, and image in question, respectively.
 As estimating longitudinal changes in the underlying manifold may be of interest, $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$ and $f_i(\mathbf{y}, t)$ are each denoted as varying-coefficient models, where the coefficients of each function are dependent on time (\cite{hastieVaryingCoefficientModels1993}).
@@
 In cases where longitudinal changes are not of interest, each varying-coefficient model may be constrained to be held constant with respect to time.
-As the $ij$th image is recorded at a single time point, $f_{ij}(\mathbf{y}, t)$ may not be dependent on time.
-We specify $f_{ij}(\mathbf{y})$ as a smoothing spline, while the functions $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$, and $f_i(\mathbf{y}, t)$ are each varying-coefficient smoothing spline models, estimated by using a smoothing spline to smooth over smoothing spline coefficients calculated at each time point available in the data.
+As the $j$th image is recorded at a single time point, $f_{j}(\mathbf{y}, t)$ may not be dependent on time.
+We specify $f_{j}(\mathbf{y})$ as a smoothing spline, while the functions $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$, and $f_i(\mathbf{y}, t)$ are each varying-coefficient smoothing spline models, estimated by using a smoothing spline to smooth over smoothing spline coefficients calculated at each time point available in the data.
 
-The smoothing spline functions $f_{ij}$ minimize 
+The smoothing spline functions $f(\mathbf{y})$ minimize 
+\[%
+  \mathcal{K}_{\lambda}(f) = \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - f(\mathbf{y}_{jm})\|^2 + \lambda\|\nabla^{\otimes 2}f\|^2
+,\]%
+where $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is given. Thus, $f(\mathbf{y})$ takes the form
 \begin{equation}
-  \mathcal{K}_{\lambda}(f_{ij}) = \sum_{l=1}^{D}\left\{\sum_{k=1}^{m_{ij}}w_{ijk}|\mathbf{x}_{ijk, l} - f_{ij, l}(\mathbf{y}_{ijk})|^2 + \lambda\|\nabla^{\otimes 2}f_l\|^2\right\}, \label{eq:3}
+  f_l(\mathbf{y}) = \sum_{j=1}^{N}\sum_{m=1}^{M_j}s_{jm, l} \times \eta_{d}\left(\|\mathbf{y} - \mathbf{y}_{jm}\|\right) + \sum_{p = 1}^{d + 1}\alpha_{p, l} \times \phi_{p}(\mathbf{y}), \quad l = 1, 2, \dots, D, \label{eq:2}
 \end{equation}
-where $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is given. Thus, $f_{ij}$ takes the form
-\begin{equation}
-  f_{ij, l}(\mathbf{y}) = \sum_{k=1}^{m_{ij}} s_{k, l} \times \eta_{4-d}\left(\mathbf{y} - \mathbf{y}_{ijk}\right) + \sum_{p=1}^{d+1} \alpha_{p, l} \times \phi_p(\mathbf{y}), \quad l = 1, 2, \dots, D, \label{eq:4}
-\end{equation}
-with $\phi_p(\mathbf{y})$ representing the basis of the linear space of polynomials on $\mathbb{R}^{d}$ with degree $\leq 1$ and $\eta_{\nu}(\mathbf{y}) = \|\mathbf{y}\|_{\mathbb{R}^{d}}^{\nu}\log\left(\|\mathbf{y}\|_{\mathbb{R}^{d}}\right)$ when $\|\mathbf{y}\|_{\mathbb{R}^{d}} \neq 0$ and $\nu$ is even, $\eta_{\nu}(\mathbf{y}) = 0$ when $\|\mathbf{y}\|_{\mathbb{R}^{d}} = 0$ and $\nu$ is even, and $\eta_{\nu}(\mathbf{y}) = \|\mathbf{y}\|_{\mathbb{R}^{d}}^{\nu}$ when $\nu$ is odd.
+with $\phi_p(\mathbf{y})$ representing the basis of the linear space of polynomials on $\mathbb{R}^{d}$ with degree $\leq 1$ and $\eta_{d}(y) = y^{4-d}\log y$ when $\log y \neq 0$ and $d$ is even, $\eta_{d}(y) = 0$ when $\log y = 0$ and $d$ is even, and $\eta_{d}(y) = y^{4-d}$ when $d$ is odd.
 
-Without accounting for the additive structure of the model, and given $m_{ij}$ input values $\mathbf{y}_{ij}$, weights $\mathbf{w}_{ij}$, and outputs $x_{ij}$, we can estimate $f_{ij}$ by obtaining the values of $s_l \in \mathbb{R}^{m_{ij}}$ and $\alpha_l \in \mathbb{R}^{d + 1}$ for $l = 1, \dots, D$ that minimize $\|\mathbf{W}^{\frac{1}{2}}\left(\mathbf{x}_{ij, l} - \mathbf{E}s_l - \mathbf{R}\alpha_l\right)\|^2_{\mathbb{R}^{m_{ij}}} + \lambda\|\mathbf{E}^{\frac{1}{2}}s_l\|^2_{\mathbb{R}^{m_{ij}}}$ under the constraint that $\mathbf{R}^{T}s_l = 0$, where
+Without accounting for the additive structure of the model, and after concatenating all relevant values into $M = \sum_{j=1}^{N}M_{j}$-dimensional vectors of input values $\mathbf{y}$, weights $\mathbf{w}$, and outputs $\mathbf{x}$, we can estimate $f$ by obtaining the values of $s_l \in \mathbb{R}^{M}$ and $\alpha_l \in \mathbb{R}^{d + 1}$ for $l = 1, \dots, D$ that minimize $\|\mathbf{W}^{\frac{1}{2}}\left(\mathbf{x}_{l} - \mathbf{E}s_l - \mathbf{R}\alpha_l\right)\|^2_{\mathbb{R}^{M}} + \lambda\|\mathbf{E}^{\frac{1}{2}}s_l\|^2_{\mathbb{R}^{M}}$ under the constraint that $\mathbf{R}^{T}s_l = 0$, where
 \begin{itemize}
-  \item $\mathbf{E} = (E_{qr})_{1 \leq q, r, \leq m_{ij}}$ is defined by $E_{qr} = \eta_{4-d}(\mathbf{y}_{ijq} - \mathbf{y}_{ijr})$.
-  \item $\mathbf{R} = \left(R_{kp}\right)_{1 \leq k \leq m_{ij}, \ 1 \leq p \leq d + 1}$ is defined by $R_{kp} = \phi_p(\mathbf{y}_{ijk})$.
-  \item $\mathbf{W} = diag(w_{ij1}, \dots, w_{ijm_{ij}})$.
+  \item $\mathbf{E} = (E_{qr})_{1 \leq q, r, \leq M}$ is defined by $E_{qr} = \eta_{d}\left(\|\mathbf{y}_{q} - \mathbf{y}_{r}\|\right))$.
+  \item $\mathbf{R} = \left(R_{mp}\right)_{1 \leq m \leq M, \ 1 \leq p \leq d + 1}$ is defined by $R_{mp} = \phi_p(\mathbf{y}_{m})$.
+  \item $\mathbf{W} = diag(w_{1}, \dots, w_{M})$.
 \end{itemize}
 
-With $m_l$ representing Lagrange multipliers, the solutions for $s_l$ and $\alpha_l$ can be found by solving
+With $\upsilon_l$ representing Lagrange multipliers, the solutions for $s_l$ and $\alpha_l$ can be found by solving
 
@@
     \alpha_l \\
-    m_l
+    \upsilon_l
   \end{array}
@@
   \end{array}
-  \right). \label{eq:5}
+  \right). \label{eq:3}
 \end{equation}
-The optimal value of $\lambda$ can be found using cross-validation, with the functions being evaluated by the mean squared error, written as
+The optimal value of $\lambda$ can be found using cross-validation, where with images being randomly split into $K$ folds, functions may be evaluated by the mean squared error, written as
 \begin{equation}
-  MSE(\lambda) = \frac{1}{m_{ij}}\sum_{k=1}^{m_{ij}}\left(\mathbf{x}_{ijk} - f_{ij}(\mathbf{y}_{ijk})\right)^2. \label{eq:6}
+  MSE(\lambda) = \frac{1}{K}\sum_{k=1}^{K}\frac{1}{N_u}\sum_{j=1}^{N_u}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\lambda}^{(k)}(\mathbf{y}_{jm})\|^2. \label{eq:4}
 \end{equation}
+Here, $f_{\lambda}^{(k)}(\mathbf{y})$ denotes the estimated spline function given tuning parameter $\lambda$ and with the data for the $k$th fold excluded.
 
-The solutions to this minimization problem suffice for estimating $f_{ij}(\mathbf{y})$, in which the coefficients are not time-varying.
-To reach estimates for the varying-coefficient models, equation \ref{eq:5} is solved using the data available at each time point within the relevant group.
+The coefficients and tuning parameter that minimize $\mathcal{K}_{\lambda}(f)$ suffice for estimating $f_{j}(\mathbf{y})$, in which the coefficients are not time-varying.
+To reach estimates for the varying-coefficient models, equation \ref{eq:3} is solved using the data available at each time point within the relevant study level.
 To estimate $f_p(\mathbf{y}, t)$, we use $\left\{\mathbf{x}_t\right\}_{t=1}^T$ and $\left\{\mathbf{y}_t\right\}_{t=1}^T$, yielding fitted function $\left\{\hat{f}_t(\mathbf{y})\right\}_{t=1}^T$.
-Using the maximum and minimum values of $\mathbf{y}_t$ in each dimension, a new $d$-dimensional grid of $m' = \text{max}\left(\left\{m_{ij}\right\}_{i=1, j = 1}^{N, N_i}\right)$ knots, $\left\{\mathbf{y}'\right\}$, is generated with the knots being evenly spread throughout the parameter space.
+Using the maximum and minimum values of $\mathbf{y}_t$ in each dimension, a new $d$-dimensional grid of $M' = \text{max}\left(\left\{M_{j}\right\}_{j = 1}^{N}\right)$ knots, $\left\{\mathbf{y'}_m\right\}_{m=1}^{M'}$, is generated with the knots being evenly spread throughout the parameter space.
 These parameters serve as common knots that allow the spline coefficients to be compared across time points.
@@
 These newly generated knots are weighted equally, so an unweighted smoothing spline model is used.
-Following \cite{greenSilverman1994}, with
+Following \cite{greenSilverman1994}, we let
 
 \begin{itemize}
-  \item $\mathbf{E}' = \left(E'_{qr}\right)_{1 \leq q, r, \leq m'}$ defined by $E'_{qr} = \eta_{4-d}\left(\mathbf{y}'_{q} - \mathbf{y}'_{r}\right)$.
-  \item $\mathbf{R}' = \left(R'_{qp}\right)_{1 \leq q \leq m', \ 1 \leq p \leq d + 1}$ is defined by $R'_{qp} = \phi_p(\mathbf{y}'_{q})$
+  \item $\mathbf{E}' = \left(E'_{qr}\right)_{1 \leq q, r, \leq M'}$ defined by $E'_{qr} = \eta_{d}\left(\|\mathbf{y}'_{q} - \mathbf{y}'_{r}\|\right)$.
+  \item $\mathbf{R}' = \left(R'_{mp}\right)_{1 \leq m \leq M', \ 1 \leq p \leq d + 1}$ is defined by $R'_{mp} = \phi_p(\mathbf{y'}_{m})$.
 \end{itemize}
-the coefficients that minimize $\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right)^{T}\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right) + \gamma\mathbf{s}_t^{T}\mathbf{E}'\mathbf{s}_t$ can be found by solving
+
+Then the coefficients that minimize $\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right)^{T}\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right) + \gamma\mathbf{s}_t^{T}\mathbf{E}'\mathbf{s}_t$ can be found by solving
 \begin{equation}
@@
   \end{array}
-  \right). \label{eq:7}
+  \right). \label{eq:5}
 \end{equation}
@@
 This function takes the form
-\begin{equation}
-  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i\|t - t_i\|^{3} + \sum_{p=1}^{2}\tau_p \phi(t)_p. \label{eq:7}
-\end{equation}
+
+\[%
+  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i\|t - t_i\|^{3} + \sum_{p=1}^{2}\tau_p \phi(t)_p
+.\]%
 We define matrix $\mathbf{A}$ as $\mathbf{A}_{qr} = \|t_q - t_r\|^{3}$, matrix $\mathbf{T}$ as $\mathbf{T}_{pr} = \phi_p(t_r)$, $\mathbf{\delta} = \left(\delta_1, \dots, \delta_T\right)^{T}$, and $\mathbf{\tau} = \left(\tau_1, \tau_2\right)^{T}$.
@@
   \end{array}
-  \right). \label{eq:8}
+  \right). \label{eq:6}
 \end{equation}
@@
 As with the tuning process for $\lambda$, the optimal value of $\gamma$ can be identified using cross-validation, with either leave-one-out or $k$-folds cross-validation being appropriate depending on the number of time points available.
-With a proper estimate of the optimum $\gamma$, we may denote the coefficients estimated by $g(t)$ as $g(t) = \Omega(t) = \left(\mathbf{s}_{\gamma}(t)^{T}, \mathbf{\alpha}_{\gamma}(t)^{T}\right)^{T}$.
+If using leave-one-out cross-validation, the mean squared error can be calculated as
+\begin{equation}
+  MSE(\gamma) = \frac{1}{T}\sum_{t=1}^{T}\frac{1}{N_t}\sum_{j=1}^{N_t}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\gamma}^{(t)}(\mathbf{y}_{jm}, t_j)\|^2. \label{eq:7}
+\end{equation}
+We then let $\hat{\gamma}$ be the value of $\gamma$ that minimizes $MSE(\gamma)$.
+
+With a proper estimate of the $\hat{\gamma}$, we may denote the coefficients estimated by $g(t)$ as $\Omega(t) = \left(\mathbf{s}_{\hat{\gamma}}(t)^{T}, \mathbf{\alpha}_{\hat{\gamma}}(t)^{T}\right)^{T}$.
+We use these coefficients in equation \ref{eq:2} in place of the constant coefficients.
 Therefore, the varying-coefficient models have the following outputs:
 \begin{equation}
-  f(\mathbf{y}, t) = \sum_{k=1}^{m'}\mathbf{s}_{\gamma}(t)_l \eta_{4-d}\left(\|\mathbf{y} - \mathbf{y'}_k\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\gamma}(t)_p \phi_p(\mathbf{y}). \label{eq:9}
+  f(\mathbf{y}, t) = \sum_{m=1}^{M'}\mathbf{s}_{\hat{\gamma}, l}(t) \eta_{d}\left(\|\mathbf{y} - \mathbf{y'}_m\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\hat{\gamma}, p, l}(t) \phi_p(\mathbf{y}), \quad l = 1, \dots, D.
+\label{eq:8}
 \end{equation}
@@
 \caption{Varying-Coefficient Smoothing Spline Model}\label{alg:varying_spline}
-  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{ijk}\right\}_{i=1, j = 1, k = 1}^{N, n_i, m_{ij}}$, $d$-dimensional input values $\left\{\mathbf{y}_{ijk}\right\}_{i=1, j=1, k=1}^{N, n_i, m_{ij}}$, weights $\left\{w_{ijk}\right\}_{i=1, j=1, k=1}^{N, n_i, m_{ij}}$, observation times $\left\{t_{ij}\right\}_{i=1, j=1}^{N, n_i}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$.}
+  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j = 1, m = 1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, weights $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and number of cross-validation folds $K$.}
   \KwResult{Analytic formula of $\hat{f}(\mathbf{y}, t)$ and optimal tuning parameter $\hat{\gamma}$.}
@@
   \For{$l = 1, 2, \dots, L$} {
-    Solve equation \ref{eq:5} to estimate $f_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}_t$, $\mathbf{y}_t$, and $w_t$\;
+    Solve equation \ref{eq:3} to estimate $f_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}_t$, $\mathbf{y}_t$, and $w_t$\;
     \For{$u = 1, 2, \dots, U_t$} {
-      Solve equation \ref{eq:5} to estimate $f^{(-u)}_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}^{(-u)}_t$, $\mathbf{y}^{(-u)}_t$, $w^{(-u)}_t$\;
-      Estimate $MSE(\lambda_{l})_u$
+      Solve equation \ref{eq:3} to estimate $f^{(-u)}_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}^{(-u)}_t$, $\mathbf{y}^{(-u)}_t$, $w^{(-u)}_t$\;
+      Estimate $MSE(\lambda_{l})_u$ using equation \ref{eq:4}\;
     }
+    Set $MSE(\lambda_l) = \frac{1}{U}\sum_{u=1}^{U}MSE(\lambda_l)_u$\;
   }
+  $\hat{\lambda}_t = \arg \min_{\lambda} MSE(\lambda)$\;
+  $f_t(\mathbf{y}) = f_{\hat{\lambda}_t}(\mathbf{y})$\;
 }
+
+  Let $M' = \max(M_{j})$, $\left\{\mathbf{y}'_{m}\right\}_{m=1}^{M'}$ be a grid spanning the range of values of $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
+  \For{$t = 1, 2, \dots, T$} {
+    Compute $\left\{\mathbf{x'}_{m, t}\right\}_{m=1}^{M'} = \left\{f_t(\mathbf{y'}_m)\right\}_{m=1}^{M'}$\;
+    Solve equation \ref{eq:5} to compute $\mathbf{s'}_t$ and $\mathbf{\alpha'}_t$\;
+    Set $\mathbf{\omega}_t = \left[\mathbf{s'}_t^{T}, \mathbf{\alpha'}_t^{T}\right]$\;
+  }
+  \For{$p = 1, 2, \dots, P$} {
+    Compute $g_{\gamma_p}(t)$ by solving equation \ref{eq:6}\;
+    \For{$t = 1, 2, \dots, T$} {
+      Compute $g_{\gamma_p}^{(t)}$ by solving \ref{eq:6}\;
+    }
+    Estimate $MSE(\gamma_p)$ using \ref{eq:7}\;
+  }
+  $\hat{\gamma} = \arg \min_{\gamma} MSE(\gamma)$\;
+  $\hat{f}(\mathbf{y}, t) = f_{\hat{\gamma}}(\mathbf{y}, t)$, where the form of $\hat{f}(\mathbf{y}, t)$ is given in \ref{eq:8}.
 \end{algorithm}

### 2023-08-15 paper/main.tex
add backfitting algorithm

@@
 
-\[%
-  \mathbf{x}_{jm} = \psi(\mathbf{y}_{jm}) = f_p(\mathbf{y}_{jm}, t_{j}) + f_{g(j)}(\mathbf{y}_{jm}, t_{j}) + f_{i(j)}(\mathbf{y}_{jm}, t_{j}) + f_j(\mathbf{y}_{jm}) + \epsilon_{jm}
-.\]%
+\begin{equation}
+  \mathbf{x}_{jm} = \psi(\mathbf{y}_{jm}) = f_p(\mathbf{y}_{jm}, t_{j}) + f_{g(j)}(\mathbf{y}_{jm}, t_{j}) + f_{i(j)}(\mathbf{y}_{jm}, t_{j}) + f_j(\mathbf{y}_{jm}) + \epsilon_{jm}. \label{eq:2}
+\end{equation}
 
@@
 \begin{equation}
-  f_l(\mathbf{y}) = \sum_{j=1}^{N}\sum_{m=1}^{M_j}s_{jm, l} \times \eta_{d}\left(\|\mathbf{y} - \mathbf{y}_{jm}\|\right) + \sum_{p = 1}^{d + 1}\alpha_{p, l} \times \phi_{p}(\mathbf{y}), \quad l = 1, 2, \dots, D, \label{eq:2}
+  f_l(\mathbf{y}) = \sum_{j=1}^{N}\sum_{m=1}^{M_j}s_{jm, l} \times \eta_{d}\left(\|\mathbf{y} - \mathbf{y}_{jm}\|\right) + \sum_{p = 1}^{d + 1}\alpha_{p, l} \times \phi_{p}(\mathbf{y}), \quad l = 1, 2, \dots, D, \label{eq:3}
 \end{equation}
@@
   \end{array}
-  \right). \label{eq:3}
+  \right). \label{eq:4}
 \end{equation}
@@
 \begin{equation}
-  MSE(\lambda) = \frac{1}{K}\sum_{k=1}^{K}\frac{1}{N_u}\sum_{j=1}^{N_u}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\lambda}^{(k)}(\mathbf{y}_{jm})\|^2. \label{eq:4}
+  MSE(\lambda) = \frac{1}{K}\sum_{k=1}^{K}\frac{1}{N_u}\sum_{j=1}^{N_u}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\lambda}^{(k)}(\mathbf{y}_{jm})\|^2. \label{eq:5}
 \end{equation}
@@
 The coefficients and tuning parameter that minimize $\mathcal{K}_{\lambda}(f)$ suffice for estimating $f_{j}(\mathbf{y})$, in which the coefficients are not time-varying.
-To reach estimates for the varying-coefficient models, equation \ref{eq:3} is solved using the data available at each time point within the relevant study level.
+To reach estimates for the varying-coefficient models, equation \ref{eq:4} is solved using the data available at each time point within the relevant study level.
 To estimate $f_p(\mathbf{y}, t)$, we use $\left\{\mathbf{x}_t\right\}_{t=1}^T$ and $\left\{\mathbf{y}_t\right\}_{t=1}^T$, yielding fitted function $\left\{\hat{f}_t(\mathbf{y})\right\}_{t=1}^T$.
@@
   \end{array}
-  \right). \label{eq:5}
+  \right). \label{eq:6}
 \end{equation}
@@
   \end{array}
-  \right). \label{eq:6}
+  \right). \label{eq:7}
 \end{equation}
@@
 \begin{equation}
-  MSE(\gamma) = \frac{1}{T}\sum_{t=1}^{T}\frac{1}{N_t}\sum_{j=1}^{N_t}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\gamma}^{(t)}(\mathbf{y}_{jm}, t_j)\|^2. \label{eq:7}
+  MSE(\gamma) = \frac{1}{T}\sum_{t=1}^{T}\frac{1}{N_t}\sum_{j=1}^{N_t}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\gamma}^{(t)}(\mathbf{y}_{jm}, t_j)\|^2. \label{eq:8}
 \end{equation}
@@
 
-With a proper estimate of the $\hat{\gamma}$, we may denote the coefficients estimated by $g(t)$ as $\Omega(t) = \left(\mathbf{s}_{\hat{\gamma}}(t)^{T}, \mathbf{\alpha}_{\hat{\gamma}}(t)^{T}\right)^{T}$.
-We use these coefficients in equation \ref{eq:2} in place of the constant coefficients.
+With a proper estimate of the $\hat{\gamma}$, we may denote the coefficients of $g(t)$ as $\Omega = \left(\delta^{T}, \tau^{T}\right)$, and those estimated by $g(t)$ as $\omega(t) = \left(\mathbf{s}_{\hat{\gamma}}(t)^{T}, \mathbf{\alpha}_{\hat{\gamma}}(t)^{T}\right)^{T}$.
+We use these coefficients in equation \ref{eq:3} in place of the constant coefficients.
 Therefore, the varying-coefficient models have the following outputs:
@@
   f(\mathbf{y}, t) = \sum_{m=1}^{M'}\mathbf{s}_{\hat{\gamma}, l}(t) \eta_{d}\left(\|\mathbf{y} - \mathbf{y'}_m\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\hat{\gamma}, p, l}(t) \phi_p(\mathbf{y}), \quad l = 1, \dots, D.
-\label{eq:8}
+\label{eq:9}
 \end{equation}
@@
   \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j = 1, m = 1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, weights $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and number of cross-validation folds $K$.}
-  \KwResult{Analytic formula of $\hat{f}(\mathbf{y}, t)$ and optimal tuning parameter $\hat{\gamma}$.}
+  \KwResult{Analytic formula of $\hat{f}(\mathbf{y}, t)$, coefficients $\Omega = \left(\mathbf{\delta}^{T}, \mathbf{\tau}^{T}\right)$, and optimal tuning parameter $\hat{\gamma}$.}
 
@@
   \For{$l = 1, 2, \dots, L$} {
-    Solve equation \ref{eq:3} to estimate $f_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}_t$, $\mathbf{y}_t$, and $w_t$\;
+    Solve equation \ref{eq:4} to estimate $f_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}_t$, $\mathbf{y}_t$, and $w_t$\;
     \For{$u = 1, 2, \dots, U_t$} {
-      Solve equation \ref{eq:3} to estimate $f^{(-u)}_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}^{(-u)}_t$, $\mathbf{y}^{(-u)}_t$, $w^{(-u)}_t$\;
-      Estimate $MSE(\lambda_{l})_u$ using equation \ref{eq:4}\;
+      Solve equation \ref{eq:4} to estimate $f^{(-u)}_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}^{(-u)}_t$, $\mathbf{y}^{(-u)}_t$, $w^{(-u)}_t$\;
+      Estimate $MSE(\lambda_{l})_u$ using equation \ref{eq:5}\;
     }
@@
     Compute $\left\{\mathbf{x'}_{m, t}\right\}_{m=1}^{M'} = \left\{f_t(\mathbf{y'}_m)\right\}_{m=1}^{M'}$\;
-    Solve equation \ref{eq:5} to compute $\mathbf{s'}_t$ and $\mathbf{\alpha'}_t$\;
+    Solve equation \ref{eq:6} to compute $\mathbf{s'}_t$ and $\mathbf{\alpha'}_t$\;
     Set $\mathbf{\omega}_t = \left[\mathbf{s'}_t^{T}, \mathbf{\alpha'}_t^{T}\right]$\;
@@
   \For{$p = 1, 2, \dots, P$} {
-    Compute $g_{\gamma_p}(t)$ by solving equation \ref{eq:6}\;
+    Compute $g_{\gamma_p}(t)$ by solving equation \ref{eq:7}\;
     \For{$t = 1, 2, \dots, T$} {
-      Compute $g_{\gamma_p}^{(t)}$ by solving \ref{eq:6}\;
+      Compute $g_{\gamma_p}^{(t)}$ by solving \ref{eq:7}\;
     }
-    Estimate $MSE(\gamma_p)$ using \ref{eq:7}\;
+    Estimate $MSE(\gamma_p)$ using \ref{eq:8}\;
   }
   $\hat{\gamma} = \arg \min_{\gamma} MSE(\gamma)$\;
-  $\hat{f}(\mathbf{y}, t) = f_{\hat{\gamma}}(\mathbf{y}, t)$, where the form of $\hat{f}(\mathbf{y}, t)$ is given in \ref{eq:8}.
+  $\hat{f}(\mathbf{y}, t) = f_{\hat{\gamma}}(\mathbf{y}, t)$, where the form of $\hat{f}(\mathbf{y}, t)$ is given in \ref{eq:9}.
+\end{algorithm}
+
+While the procedure outlined in Algorithm \ref{alg:varying_spline} yields an appropriate estimator when working with unnested data, this approach must be adapted to work within the additive model structure of model \ref{eq:2}.
+As discussed in \cite{hastieGeneralizedAdditiveModels1990}, the additive model $\psi(\mathbf{y})$ can be estimated using a backfitting procedure, which uses the model's additive nature to iteratively fit each distinct smoothing spline function until convergence.
+Specifically, we use the fact that, under the model,
+
+\[%
+  f_p(\mathbf{y}_{jm}, t_{j}) = E\left[\mathbf{x}_{jm} - f_{g(j)}(\mathbf{y}_{jm}, t_{j}) - f_{i(j)}(\mathbf{y}_{jm}, t_{j}) - f_{j}(\mathbf{y}_{jm})\right]
+,\]%
+with a similar result holding for each of the smooth functions making up the additive model. Thus, the backfitting algorithm to fit the proposed model is given below.
+
+\begin{algorithm}
+\caption{Backfitting Algorithm for Additive Spline Model}\label{alg:backfitting}
+  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$, and $\epsilon$.}
+  \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$.}
+
+  Estimate $f_{p}^{(0)}(\mathbf{y}, t)$ and $\Omega_{p}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_{j}\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+  \For{$g = 1, 2, \dots, G$} {
+    Estimate $f_{g}^{(0)}(\mathbf{y}, t)$ and $\Omega_{g}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{t_j\right\}_{j=1}^{N_g}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $K$\;
+  }
+  \For{$i = 1, 2, \dots, I$} {
+    Estimate $f_{i}^{(0)}(\mathbf{y}, t)$ and $\Omega_{i}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{t_j\right\}_{j=1}^{N_i}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+  }
+  \For{$j = 1, 2, \dots, N$} {
+    Estimate $f_{j}^{(0)}(\mathbf{y})$ and $\omega_{j}^{(0)}$ by solving \ref{eq:4} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{m=1}^{M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{m=1}^{M_j}$, $\left\{w_{jm}\right\}_{m = 1}^{M_j}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $K$\;
+  }
+  $n \gets 0$\;
+  \While{$\max\left(\|\Omega_p^{(n)} - \Omega_{p}^{(n-1)}\|, \left\{\|\Omega_{g}^{(n)} - \Omega_{g}^{(n-1)}\|\right\}_{g=1}^{G}, \left\{\|\Omega_{i}^{(n)} - \Omega_{i}^{(n-1)}\|\right\}_{i=1}^{I}, \left\{\|\omega_{j}^{(n)} - \omega_{j}^{(n-1)}\|\right\}_{j=1}^{N}\right) \geq \epsilon$} {
+    Estimate $f_{p}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{p}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{g(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_j^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_{j}\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+    \For{$g = 1, 2, \dots, G$} {
+      Estimate $f_{g}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{g}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_{j}^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{t_j\right\}_{j=1}^{N_g}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $K$\;
+    }
+    \For{$i = 1, 2, \dots, I$} {
+      Estimate $f_{i}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{i}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_j^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{t_j\right\}_{j=1}^{N_i}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+    }
+    \For{$j = 1, 2, \dots, N$} {
+      Estimate $f_{j}^{(n+1)}(\mathbf{y})$ and $\omega_{j}^{(n+1)}$ by solving \ref{eq:4} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j)\right\}_{m=1}^{M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{m=1}^{M_j}$, $\left\{w_{jm}\right\}_{m = 1}^{M_j}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $K$\;
+    }
+    $n \gets n + 1$
+  }
 \end{algorithm}
@@
 For the time being, this extension is left to future work.
-
-As discussed in \cite{hastieGeneralizedAdditiveModels1990}, the additive model $\psi(\mathbf{y})$ can be estimated using a backfitting procedure, which uses the model's additive nature to iteratively fit each distinct smoothing spline function until convergence.
-Specifically, we use the fact that, under the model,
-
-\[%
-f_p(\mathbf{y}_{ijk}, t_{ij}) = E\left[\mathbf{x}_{ijk} - f_{g(i)}(\mathbf{y}_{ijk}, t_{ij}) - f_{i}(\mathbf{y}_{ijk}, t_{ij}) - f_{ij}(\mathbf{y}_{ijk})\right]
-,\]%
-with a similar result holding for each of the smooth functions making up the additive model. Thus, the backfitting algorithm to fit the proposed model is given below.
-
-\zielinski{The following algorithm is currently used as a placeholder while I work out the specifics.}
-
-\zielinski{Try using a ratio of the norm of the difference in model coefficients and the norm of the old model to measure convergence.
-This will set the change in the model coefficients (as measured by the norm of the difference vector/matrix) relative to the magnitude of the model coefficients.}
-
-\begin{algorithm}
-\caption{Backfitting Algorithm for Additive Spline Model}\label{alg:backfitting}
-  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{ijk}\right\}_{i=1, j=1, k=1}^{N, n_i, m_{ij}}$, $d$-dimensional input values $\left\{\mathbf{y}_{ijk}\right\}_{i=1, j=1, k=1}^{N, n_i, m_{ij}}$, and observation times $\left\{t_{ij}\right\}_{i=1, j=1}^{N, n_i}$.}
-  \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{N}$, and $\left\{\hat{f}_{ij}\right\}_{i=1, j=1}^{N, n_i}$, and optimal tuning parameters $\hat{\lambda}_p$, $\hat{\gamma}_p$, $\left\{\hat{\lambda}_g\right\}_{g=1}^{G}$, $\left\{\hat{\gamma}\right\}_{g=1}^{G}$, $\left\{\hat{\lambda}_{i}\right\}_{i=1}^{N}$, $\left\{\hat{\gamma}_{i}\right\}_{i=1}^{N}$, $\left\{\hat{\lambda}_{ij}\right\}_{i=1, j=1}^{N, n_i}$.}
-
-\For{$t = 1, 2, \dots, T$} {
-  Apply HDMDE algorithm with input $\left(\left\{Y_{i, t}\right\}_{i = 1}^{I_t}, N_0, \epsilon, \alpha\right)$ and obtain $N_t$, $\left\{\mu_{j, t}\right\}_{j = 1}^{N_t}$, and $\left\{\theta_{j, t}\right\}_{j = 1}^{N_t}$\;
-}
-
-  Apply Isomap to parameterize $\left\{\mu_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$ by the $d$-dimensional parameters $\left\{r_{j, t}\right\}_{j = 1, t = 1}^{N_t, T}$.
-Formally set $\pi_{f_t(0)}(\mu_{j, t}) \gets r_{j, t}$ for $j = 1, 2, \dots, N_t$, $t = 1, 2, \dots, T$\;
-
-\For{t = 1, 2, \dots, T} {
-  Apply modified PME algorithm with $\pi_{f_t(0)}(\mu_{j, N_t, t}) \gets \left\{r_{j, t}\right\}_{j = 1}^{N_t}$ and obtain $f_t$, $\lambda_t^*$, and $\tau_t$\;
-  $\left\{r_{j, t}\right\} \gets \pi_{f_t}(\mu_{j, t})$ for $j = 1, \dots, N_t$ and $X_{j, t} \gets f_t(r_{j, t})$ for $j = 1, \dots, N_t$\;
-}
-
-  Let $N = \max(N_t)$, $\left\{r_i^*\right\}_{i=1}^{N}$ be a grid spanning the range of estimated values of $\left\{r_{i, j}\right\}_{i=1, t=1}^{N_t, T}$\;
-
-\For{t = 1, 2, \dots, T} {
-  Set $X_{i, t} = f_t(r_{i}^*)$ for $i = 1, \dots, N$\;
-  Compute $f_t^*(r)$ by solving \eqref{eq:13}\;
-  Set $w_t = \frac{\frac{1}{\tau_t}}{\sum_{i=1}^{T}\frac{1}{\tau_i}}$\;
-}
-
-  Define $\boldsymbol{\omega}$ by setting $\boldsymbol{\omega}_t = \left[\boldsymbol{s}_t, \boldsymbol{\alpha}_t\right]$\;
-  \For{l = 1, 2, \dots, L} {
-    Compute $g_{\gamma_l}(t)$ by solving \eqref{eq:16}\;
-    \For{t = 1, 2, \dots, T} {
-      Compute $g_{\gamma_l}^{(t)}$ by solving \eqref{eq:16}\;
-    }
-    Estimate $MSD(\gamma_l)$ using \eqref{eq:18}\;
-  }
-  $\gamma^* = \arg\min_{\gamma}MSD(\gamma)$\;
-  $f^*(t, \boldsymbol{r}) = f_{\gamma^*}(t, \boldsymbol{r})$, where the form of $f^*(t, \boldsymbol{r})$ is given in \eqref{eq:17}
-\end{algorithm}
 

### 2023-08-15 paper/main.tex
update backfitting algorithm formatting
update backfitting algorithm, hierarchical pme tuning discussion

@@
 \begin{equation}
-  \mathbf{x}_{jm} = \psi(\mathbf{y}_{jm}) = f_p(\mathbf{y}_{jm}, t_{j}) + f_{g(j)}(\mathbf{y}_{jm}, t_{j}) + f_{i(j)}(\mathbf{y}_{jm}, t_{j}) + f_j(\mathbf{y}_{jm}) + \epsilon_{jm}. \label{eq:2}
+  \mathbf{x}_{jm} = \psi_j(\mathbf{y}_{jm}, t_j) = f_p(\mathbf{y}_{jm}, t_{j}) + f_{g(j)}(\mathbf{y}_{jm}, t_{j}) + f_{i(j)}(\mathbf{y}_{jm}, t_{j}) + f_j(\mathbf{y}_{jm}) + \epsilon_{jm}.
+\label{eq:2}
 \end{equation}
@@
 While the procedure outlined in Algorithm \ref{alg:varying_spline} yields an appropriate estimator when working with unnested data, this approach must be adapted to work within the additive model structure of model \ref{eq:2}.
-As discussed in \cite{hastieGeneralizedAdditiveModels1990}, the additive model $\psi(\mathbf{y})$ can be estimated using a backfitting procedure, which uses the model's additive nature to iteratively fit each distinct smoothing spline function until convergence.
+As discussed in \cite{hastieGeneralizedAdditiveModels1990}, the additive model $\psi(\mathbf{y}, t)$ can be estimated using a backfitting procedure, which uses the model's additive nature to iteratively fit each distinct smoothing spline function until convergence.
 Specifically, we use the fact that, under the model,
@@
 ,\]%
-with a similar result holding for each of the smooth functions making up the additive model. Thus, the backfitting algorithm to fit the proposed model is given below.
+with a similar result holding for each of the smooth functions making up the additive model. Thus, the backfitting algorithm to fit the proposed model is given in Algorithm \ref{alg:backfitting}.
 
 \begin{algorithm}
+\footnotesize
 \caption{Backfitting Algorithm for Additive Spline Model}\label{alg:backfitting}
@@
     }
-    $n \gets n + 1$
+    $n \gets n + 1$\;
   }
+  $\hat{f}_p \gets f_{p}^{(n)}$, $\left\{\hat{f}_g\right\}_{g = 1}^{G} \gets \left\{f_{g}^{(n)}\right\}_{g = 1}^{G}$, $\left\{\hat{f}_i\right\}_{i=1}^{I} \gets \left\{f_{i}^{(n)}\right\}_{i=1}^{I}$, and $\left\{\hat{f}_j\right\}_{j=1}^{N} \gets \left\{f_j^{(n)}\right\}_{j=1}^{N}$.
 \end{algorithm}
@@
 
-A clear concern with the use of this procedure in this case is the number of functions to be fit, which can quickly increase the computational burden of fitting the models.
+A clear concern with the use of this procedure in this case is the number of functions to be estimated, which can quickly increase the computational burden of fitting the models.
 However, the nesting of the different levels of the data helps to limit the number of function combinations that must be fit, helping to keep computational requirements manageable.
@@
 
-With the structure of the additive spline model established, we now incorporate it into the larger manifold learning procedure.
+With the structure of the additive spline model established, we now incorporate it into the larger manifold learning process.
 This procedure is based on the Principal Manifold Estimation (PME) algorithm introduced in \cite{mengPrincipalManifoldEstimation2021}, which iterates between estimating a function to embed observations in $D$-dimensional space and reparameterizing those observations in the $d$-dimensional space to estimate the underlying manifold.
+The PME algorithm is largely split into three distinct steps: data reduction, fitting, and tuning.
+The split between these three steps enforces a process by which candidate functions are fit for each in a vector of smoothing values in the fitting stage, then the tuning step is used to select the smoothing value that results in the lowest error, as measured by mean squared distance,
+\[%
+  \mathcal{D}(\hat{f}) = \frac{1}{N}\sum_{i=1}^{N}\|x_i - \hat{f}_\lambda\left(\pi_{\hat{f}_\lambda}(x_i)\right)\|_{\mathbb{R}^{D}}^2
+,\]%
+where $\hat{f}_\lambda$ represents the estimated embedding map given smoothing parameter $\lambda$, $\pi_{\hat{f}_\lambda}(x)$ denotes the projection index that obtains a $d$-dimensional parameterization of $D$-dimensional input $x$, and $N$ represents the total number of observations in the dataset.
+
+While this approach to tuning the model parameters is beneficial because it allows a complete comparison of all parameter values under consideration, the additive model described in equation \ref{eq:2} requires many more parameters, meaning that taking the same tuning approach would result in the need to find an optimal point in a very high-dimensional space.
+This would greatly increase the computational complexity of the algorithm, likely to the point of infeasibility in an practical situation.
+Instead, the proposed algorithm avoids this problem by taking a greedy approach to tuning the model, where the smoothing parameters for each model are chosen individually, rather than simultaneously.
+The proposed algorithm is given as Algorithm \ref{alg:hpme}.
+
+\begin{algorithm}
+\caption{Backfitting Algorithm for Additive Spline Model}\label{alg:hpme}
+  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$, and $\epsilon$.}
+  \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$.}
+
+\end{algorithm}
 

### 2023-08-15 paper/main.tex
add preliminary hierarchical pme discussion

@@
 Instead, the proposed algorithm avoids this problem by taking a greedy approach to tuning the model, where the smoothing parameters for each model are chosen individually, rather than simultaneously.
-The proposed algorithm is given as Algorithm \ref{alg:hpme}.
+
+Thus, the hierarchical PME algorithm proposed here is composed of initialization and data reduction and fitting steps.
+The first step uses the PME algorithm to estimate an embedding map from one image.
+The HDMDE algorithm, detailed in \cite{mengPrincipalManifoldEstimation2021}, is used to reduce the data from each image into a set of component centers and associated weights.
+Then, the embedding map estimated using the PME algorithm can be used to provide initial parameterizations of the component centers for every image.
+These initial parameterizations are then used in the fitting step.
+
+Like the PME algorithm, the fitting step uses an iterative approach to alternate between using the parameterizations to estimate an embedding function with the form of the additive spline model in equation \ref{eq:2}, then using this newly generated embedding function to reestimate the most appropriate parameterizations for the reduced set of components identified in the data reduction process.
+Once the error of the embedding function as measured by the sum of squared distances measured at the component centers declines by less than a given threshold $\epsilon$, the iterative procedure stops and the embedding function and associated parameterizations are finalized.
+The details of this process are given in Algorithm \ref{alg:hpme}.
 
 \begin{algorithm}
-\caption{Backfitting Algorithm for Additive Spline Model}\label{alg:hpme}
-  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$, and $\epsilon$.}
-  \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$.}
+\caption{Hierarchical Principal Manifold Estimation Algorithm}\label{alg:hpme}
+  \KwData{Data points $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$ in $\mathbb{R}^{D}$; a positive integer $N_0 < \sum_{j=1}^{N}M_j$; $\alpha, \epsilon, \epsilon^{*} \in (0, 1)$; candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}, \left\{\gamma_p\right\}_{p=1}^{P}$; $itr \geq 1$, which is the maximum number of iterations allowed.}
+  \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$, .}
 

### 2023-08-16 paper/main.tex
add hierarchical pme algorithm

@@
 
-The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{ijk})$, which can be modeled as
-
+The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{jm})$, which can be modeled as 
 \begin{equation}
-  \mathbf{x}_{jm} = \psi_j(\mathbf{y}_{jm}, t_j) = f_p(\mathbf{y}_{jm}, t_{j}) + f_{g(j)}(\mathbf{y}_{jm}, t_{j}) + f_{i(j)}(\mathbf{y}_{jm}, t_{j}) + f_j(\mathbf{y}_{jm}) + \epsilon_{jm}.
-\label{eq:2}
+  \mathbf{x}_{jm} = \psi(\mathbf{y}_{jm}, t_j) = f_p(\mathbf{y}_{jm}, t_{j}) + f_{g(j)}(\mathbf{y}_{jm}, t_{j}) + f_{i(j)}(\mathbf{y}_{jm}, t_{j}) + f_j(\mathbf{y}_{jm}) + \epsilon_{jm}. \label{eq:2}
 \end{equation}
+The embedding map $\psi(\mathbf{y})$ is associated with a projection index $\kappa_{\psi}(\mathbf{x}): \mathbb{R}^{D} \to \mathbb{R}^{d}$, which, informally, outputs the $d$-dimensional parameterization $\mathbf{y}$ that minimizes $\|\mathbf{x} - \psi(\mathbf{y})\|_{\mathbb{R}^{D}}$ and follows the generalized definition of projection indices given in \cite{mengPrincipalManifoldEstimation2021}.
 
@@
 \caption{Backfitting Algorithm for Additive Spline Model}\label{alg:backfitting}
-  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$, and $\epsilon$.}
+  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, weights $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$ and $\left\{\gamma_p\right\}_{p=1}^{P}$, $\epsilon$ and $K$.}
   \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$.}
@@
 \caption{Hierarchical Principal Manifold Estimation Algorithm}\label{alg:hpme}
-  \KwData{Data points $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$ in $\mathbb{R}^{D}$; a positive integer $N_0 < \sum_{j=1}^{N}M_j$; $\alpha, \epsilon, \epsilon^{*} \in (0, 1)$; candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}, \left\{\gamma_p\right\}_{p=1}^{P}$; $itr \geq 1$, which is the maximum number of iterations allowed.}
+  \KwData{Data points $\left\{\mathbf{v}_{jm}\right\}_{j=1, m=1}^{N, M'_j}$ in $\mathbb{R}^{D}$; group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$; a positive integer $N_0 < \min\left(\left\{M'_j\right\}_{j=1}^{N}\right)$; $\alpha, \epsilon, \epsilon^{*} \in (0, 1)$; candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}, \left\{\gamma_p\right\}_{p=1}^{P}$; $itr \geq 1$, which is the maximum number of iterations allowed.}
   \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$, .}
-
+  \textbf{Initialization}: Apply PME with input $\left(\left\{\mathbf{v}_{1m}\right\}_{m=1}^{M'_1}, N_0, \alpha, \epsilon, \epsilon^{*}, \left\{\lambda_l\right\}_{l=1}^{L}, itr\right)$ and obtain $f^{(1)}$\;
+  $\psi^{(1)} \gets f^{(1)}$\;
+  \textbf{Data Reduction}: Apply HDMDE algorithm with inputs $\left(\left\{\left\{\mathbf{v}_{jm}\right\}_{m=1}^{M_j}\right\}_{j=1}^{N}, N_0, \epsilon, \alpha\right)$ and obtain $\left\{M_j\right\}_{j=1}^{N}$, $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, and $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
+  \textbf{Parameterization}: Use the projection index of $\psi^{(1)}$, $\kappa_{\psi^{(1)}}$, to parameterize $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$ by $d$-dimensional parameters $\left\{y_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
+  $\mathcal{E} \gets 2 \times \epsilon^{*}$, $n \gets 1$, and $\mathcal{D}(\psi^{(1)}) \gets \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - \psi^{(1)}\left(\kappa_{\psi^{(1)}}(\mathbf{x}_{jm})\right)\|_{\mathbb{R}^{D}}^2$\;
+  \While{$\mathcal{E} \geq \epsilon^{*}$ and $n < itr$} {
+    Compute $f_{p}^{(n + 1)}$, $\left\{f_{g}^{(n + 1)}\right\}_{g = 1}^{G}$, $\left\{f_{i}^{(n + 1)}\right\}_{i=1}^{I}$, and $\left\{f_{j}^{(n+1)}\right\}_{j=1}^{N}$ by using Algorithm \ref{alg:backfitting} with inputs $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\kappa_{\psi^{(n)}}(\mathbf{x}_{jm})\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_j\right\}_{j=1}^{N}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{g(j)\right\}_{j=1}^{N}$, $\left\{i(j)\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $\epsilon$, and $K$\;
+    $\left\{\psi^{(n+1)}(\mathbf{y}, t) \gets f_{p}^{(n + 1)}(\mathbf{y}, t) + f_{g(j)}^{(n+1)}(\mathbf{y}, t) + f_{i(j)}^{(n + 1)}(\mathbf{y}, t) + f_{j}^{(n+1)}(\mathbf{y})\right\}_{j=1}^{N}$\;
+    $\mathcal{D}(\psi^{n+1}) \gets \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - \psi^{n + 1}\left(\kappa_{\psi^{n+1}}(\mathbf{x}_{jm})\right)\|_{\mathbb{R}^{D}}^2$\;
+    $\mathcal{E} \gets |\mathcal{D}(\psi^{(n+1)}) - \mathcal{D}(\psi^{(n)})| / \mathcal{D}(\psi^{(n)})$\;
+    $n \gets n + 1$\;
+  }
+  $\hat{f}_p \gets f_{p}^{(n)}$, $\left\{\hat{f}_{g} \gets f_{g}^{(n)}\right\}_{g=1}^{G}$, $\left\{\hat{f}_{i} \gets f_{i}^{(n)}\right\}_{i=1}^{I}$, $\left\{\hat{f}_j \gets f_{j}^{(n)}\right\}_{j=1}^{N}$.
 \end{algorithm}

### 2023-10-02 paper/main.tex
Updates from Overleaf

@@
 Most commonly-used manifold learning algorithms, including Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), Locally Linear Embedding (\cite{roweisNonlinearDimensionalityReduction2000}), Laplacian Eigenmaps (\cite{belkinLaplacianEigenmapsDimensionality2003}), and Diffusion Maps (\cite{coifmanDiffusionMaps2006}) construct a graph of similarities between observations, which is then used to determine the structure of the manifold.
-\cite{mengPrincipalManifoldEstimation2021} presents an alternative principal manifold framework for manifold estimation which extends to higher dimensions the principal curve framework introduced in \cite{hastiePrincipalCurves1989}, in which principal curves are defined as one-dimensional curves that pass through the middle of a high-dimensional dataset.
+\cite{mengPrincipalManifoldEstimation2021} present an alternative principal manifold framework for manifold estimation which extends to higher dimensions the principal curve framework introduced by \cite{hastiePrincipalCurves1989}, in which principal curves are defined as one-dimensional curves that pass through the middle of a high-dimensional dataset.
 
@@
 treatment groups in a clinical trial).
-This goal is particularly relevant in the case of the complex neurodegenerative conditions mentioned above, where the high degree of heterogeneity in patient experiences have led to attempts to identify subgroups of patients based on a range of criteria such as symptom expression, onset, or rate of progression [Citations].
+This goal is particularly relevant in the case of the complex neurodegenerative conditions mentioned above, where the high degree of heterogeneity in patient experiences has led to attempts to identify subgroups of patients based on a range of criteria such as symptom expression, onset, or rate of progression [Citations].
 Fundamentally, if a low dimensional manifold is thought to effectively represent data related to the patient experience of a neurodegenerative condition, and the experience of that condition differs by group, then the low-dimensional representation of that experience should also differ by group.
 
-A method to accurately and efficiently estimate the manifolds underlying data at the group level must take advantage of the grouped structure of observations found in the dataset to which it is to be applied.
-To develop an estimate of the manifold underlying the hippocampi of participants in the ADNI study, it is necessary to recognize the basic nesting structure of the data, in which brain images are nested within participants, which may belong to a latent disease subgroup for which patients tend to have more related symptom experiences or underlying physiological developments, which are in turn nested within study groups based on Alzheimer's disease status, which are finally nested within the greater study population.
-The goal of this article is to develop a manifold learning approach that appropriately accounts for this hierarchical structure, estimates the underlying manifold at each level of the hierarchy, and allows for statistical testing of differences between those low-dimensional manifolds.
+A method to accurately and efficiently estimate the manifolds underlying data at the group level must take advantage of the grouped structure of observations.
+To develop an estimate of the manifold underlying the hippocampi of participants in the ADNI study, it is necessary to recognize the basic nesting structure of the data, in which brain images are nested within participants possibly belonging to a latent disease subgroup where patients tend to have more related symptom experiences or underlying physiological developments, which are in turn nested within study groups based on Alzheimer's disease status, that are finally nested within the greater study population.
+In this paper, we develop a manifold learning approach that appropriately accounts for this hierarchical structure when modeling the underlying manifolds.
+We further propose an approach for estimation of the underlying manifold at each level of the hierarchy allowing for statistical testing of differences between those low-dimensional manifolds.
 
-There have been several approaches taken to address hierarchical structure in manifold learning problems.
+There have been attempts to address the hierarchical structure in manifold learning problems.
 \cite{bhatiaHierarchicalManifoldLearning2012} introduced a modification of Laplacian eigenmaps to estimate manifolds from recursively subdivided regions of medical images while enforcing similar alignment of the manifolds estimated from related regions.
-\cite{freicheCharacterizingMyocardialIschemia2022} builds on this approach to incorporate multiple imaging modalities and automate the tuning of the algorithm's hyperparameters.
-
-\cite{gaoHierarchicalManifoldLearning2019} proposed a two-step diffusion maps method for hierarchical data with variation over several domains, where diffusion maps is applied first to reduce the images for each individual to $d_1$ dimensions, resulting in individual-specific manifold estimates, then diffusion maps is applied again on the concatenated parameterizations to reduce over the individuals to $d_2$ dimensions, thus estimating a group-level manifold.
+\cite{freicheCharacterizingMyocardialIschemia2022} built on this approach to incorporate multiple imaging modalities and automate the tuning of the algorithm's hyperparameters.
+\cite{gaoHierarchicalManifoldLearning2019} proposed a method based on two-step diffusion maps for hierarchical data with variation over several domains, where diffusion maps are applied first to reduce the dimension of the images for each individual, resulting in individual-specific manifold estimates, then diffusion maps are applied again on the concatenated parameterizations to reduce over the individuals, thus estimating a group-level manifold.
 This approach allows the first step of dimension reduction to capture the most important qualities of each image separately before reducing across subjects, but the two step procedure ensures that the manifold estimated at the individual level has no relation to the manifold estimated at the group level, posing problems for the interpretability of the embedding mappings and comparisons of manifolds between levels.
@@
 
-This article builds on the principal manifold estimation (PME) algorithm introduced in \cite{mengPrincipalManifoldEstimation2021}, which uses an iterative procedure to specify an embedding mapping from the low-dimensional manifold to a high-dimensional space that takes the form of a smoothing spline.
+In this work, we build on the principal manifold estimation (PME) algorithm introduced by \cite{mengPrincipalManifoldEstimation2021}, which uses an iterative procedure to specify an embedding mapping from the low-dimensional manifold to a high-dimensional space that takes the form of a smoothing spline.
 Drawing from literature in hierarchical and additive models, we replace the smoothing spline in the PME algorithm with an additive smoothing spline model (\cite{gelmanDataAnalysisUsing2007}, \cite{hastieGeneralizedAdditiveModels1990}, \cite{gelmanBayesianDataAnalysis2014}).
 While the specific model structure may change to meet the needs of the application, \cite{brumbackSmoothingSplineModels1998} and \cite{schulamFrameworkIndividualizingPredictions2015a} provide examples of the general model framework used.
-This ultimately yields a nonlinear dimension reduction algorithm that takes advantage of a statistically principled model to address nesting structures present in the input data.
+This ultimately yields a nonlinear dimension reduction algorithm that takes advantage of a statistically principled model to address nesting structures of the input data.
 As it is based on the PME framework, the algorithm outputs an explicit representation of the embedding function, making it possible to both generate new data in high-dimensional space using low-dimensional parameters and map new input data to the estimated manifold.
@@
 
-The following sections are organized as follows: Section 2 specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper.
+The following sections are organized as follows.
+Section \ref{s:methods} specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper.
 Section 3 applies this method to several simulated datasets using a variety of underlying manifold structures and tests performance on the MNIST dataset of handwritten digits.
@@
 
-\section{Methods}
+\section{Methods}\label{s:methods}
 
@@
 We begin this section with a brief review of additive models.
-Consider a regression scenario where output variable $\mathbf{y} = \left(y_1, \dots, y_n\right)^{T}$ are thought to be influenced by covariate vectors $\mathbf{x}_i = \left(x_{i1}, \dots, x_{ip}\right)$.
+Consider a regression scenario where the dependent variable $\mathbf{y} = \left(y_1, \dots, y_n\right)^{T}$ is thought to be associated with independent covariate vectors $\mathbf{x}_i = \left(x_{i1}, \dots, x_{ip}\right)$.
 In this situation, the standard multiple linear regression model assumes that the dependence of $y_i$ on $\mathbf{x}_i$ takes the linear form
@@
 \end{equation}
-where again $\epsilon_i$ has mean 0 and variance $\sigma^2$, but the $f_j$s represent arbitrary functions which are commonly smooth.
+where again $\epsilon_i$ has mean 0 and variance $\sigma^2$, however for each $j$, $f_j$ represents an arbitrary function commonly assumed to be smooth.
 While the functions in model \ref{eq:1} only take as parameters the value of one covariate, with each covariate being represented in the model once, this does not need to be the case.
@@
 
-With this background in place, we turn to the motivating situation of analyzing the subcortical surfaces of ADNI participants using MR images, with a focus on the hippocampus.
-ADNI is a longitudinal observational study, where several imaging modalities are used to record neuroimages from participants over time.
+With this background in place, we turn to the motivating example of analyzing the subcortical surfaces of ADNI participants using MR images, with a focus on the hippocampus, although the framework can be applied to modeling other structures in the brain.
+ADNI is a longitudinal observational study, where several imaging modalities are used to record brain images from participants over time.
 Participants can generally be considered to belong to three groups based on cognitive state: elderly healthy controls, those displaying symptoms of mild cognitive impairment, and those diagnosed with AD.
 In practice, participants may transition between these categorizations as their symptoms progress.
-However, because patients may be diagnosed with AD at different stages, accurately modeling these transitions is challenging and is beyond the scope of this article.
+However, because patients may be diagnosed with AD at different stages, accurately modeling these transitions is challenging and is beyond the scope of this manuscript.
 For simplicity, we use the groupings recorded at the time each participant was recruited to the study.
@@
 
+\section{Aim 2: Principal Manifold Estimation for Hierarchical Data} \label{sec:hpme}
+
+\subsection{Background}
+
+Neurodegenerative diseases such as Alzheimer's disease (AD) and Parkinson's disease are highly complex, chronic conditions that exhibit substantial heterogeneity between patients.
+To enable researchers to identify potential biomarkers of progression for these conditions, large multi-site longitudinal observational studies have been conducted that collect information from a wide range of participants over long durations across several modalities, including neuroimaging (\cite{muellerWaysEarlyDiagnosis2005}, \cite{marekParkinsonProgressionMarker2011}).
+Neuroimaging data in particular have been shown to be sensitive to developments in the early stages of disease progression, often before diagnosis occurs (\cite{jackHypotheticalModelDynamic2010}).
+
+While observations from neuroimaging data are typically high-dimensional, it is commonly assumed that the data lie along a low-dimensional manifold.
+Under this assumption, manifold learning algorithms can be used to recover this underlying manifold and parameterize this high-dimensional input data in the learned low-dimensional space.
+Most commonly-used manifold learning algorithms, including Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), Locally Linear Embedding (\cite{roweisNonlinearDimensionalityReduction2000}), Laplacian Eigenmaps (\cite{belkinLaplacianEigenmapsDimensionality2003}), and Diffusion Maps (\cite{coifmanDiffusionMaps2006}) construct a graph of similarities between observations, which is then used to determine the structure of the manifold.
+\cite{mengPrincipalManifoldEstimation2021} present an alternative principal manifold framework for manifold estimation which extends to higher dimensions the principal curve framework introduced by \cite{hastiePrincipalCurves1989}, in which principal curves are defined as one-dimensional curves that pass through the middle of a high-dimensional dataset.
+
+In neuroimaging settings, manifold learning algorithms are most often used to estimate the underlying manifold at the population level, with vectorized image intensities being used as the input data.
+\cite{yueParameterizationWhiteMatter2016} used manifold learning to estimate a low-dimensional parameterization of the corpus callosum at the individual level.
+However, in many cases, the outcome of interest is framed as a comparison between different groups of individuals (e.g.
+treatment groups in a clinical trial).
+This goal is particularly relevant in the case of the complex neurodegenerative conditions mentioned above, where the high degree of heterogeneity in patient experiences has led to attempts to identify subgroups of patients based on a range of criteria such as symptom expression, onset, or rate of progression (\cite{ferreiraBiologicalSubtypesAlzheimer2020}, \cite{venutoPredictingAmbulatoryCapacity}).
+Fundamentally, if a low dimensional manifold is thought to effectively represent data related to the patient experience of a neurodegenerative condition, and the experience of that condition differs by group, then the low-dimensional representation of that experience should also differ by group.
+
+A method to accurately and efficiently estimate the manifolds underlying data at the group level must take advantage of the grouped structure of observations.
+To develop an estimate of the manifold underlying the hippocampi of participants in the ADNI study, it is necessary to recognize the basic nesting structure of the data, in which brain images are nested within participants possibly belonging to a latent disease subgroup where patients tend to have more related symptom experiences or underlying physiological developments, which are in turn nested within study groups based on Alzheimer's disease status, that are finally nested within the greater study population.
+In this paper, we develop a manifold learning approach that appropriately accounts for this hierarchical structure when modeling the underlying manifolds.
+We further propose an approach for estimation of the underlying manifold at each level of the hierarchy allowing for statistical testing of differences between those low-dimensional manifolds.
+
+There have been attempts to address the hierarchical structure in manifold learning problems.
+\cite{bhatiaHierarchicalManifoldLearning2012} introduced a modification of Laplacian eigenmaps to estimate manifolds from recursively subdivided regions of medical images while enforcing similar alignment of the manifolds estimated from related regions.
+\cite{freicheCharacterizingMyocardialIschemia2022} built on this approach to incorporate multiple imaging modalities and automate the tuning of the algorithm's hyperparameters.
+\cite{gaoHierarchicalManifoldLearning2019} proposed a method based on two-step diffusion maps for hierarchical data with variation over several domains, where diffusion maps are applied first to reduce the dimension of the images for each individual, resulting in individual-specific manifold estimates, then diffusion maps are applied again on the concatenated parameterizations to reduce over the individuals, thus estimating a group-level manifold.
+This approach allows the first step of dimension reduction to capture the most important qualities of each image separately before reducing across subjects, but the two step procedure ensures that the manifold estimated at the individual level has no relation to the manifold estimated at the group level, posing problems for the interpretability of the embedding mappings and comparisons of manifolds between levels.
+
+\cite{guerreroGroupconstrainedManifoldLearning2017} noted that the grouping behavior of observations recorded longitudinally from the same individual adversely impacted the creation of the neighborhood graph between data points when estimating a manifold for the entire population.
+The authors proposed a constraint specifying that, a given observation's collection of nearest neighbors may include at most one point from each individual or other specified grouping.
+This encourages a more even representation of the entire space of observations in the neighborhood graph, and may be applied to any manifold learning algorithm that features a neighborhood graph to specify proximities between observations.
+However, it simply mitigates the effects of the nesting structure on a population-level manifold estimate rather than enabling a detailed comparison of manifold estimates for each group.
+
+In this work, we build on the principal manifold estimation (PME) algorithm introduced by \cite{mengPrincipalManifoldEstimation2021}, which uses an iterative procedure to specify an embedding mapping from the low-dimensional manifold to a high-dimensional space that takes the form of a smoothing spline.
+Drawing from literature in hierarchical and additive models, we replace the smoothing spline in the PME algorithm with an additive smoothing spline model (\cite{gelmanDataAnalysisUsing2007}, \cite{hastieGeneralizedAdditiveModels1990}, \cite{gelmanBayesianDataAnalysis2014}).
+While the specific model structure may change to meet the needs of the application, \cite{brumbackSmoothingSplineModels1998} and \cite{schulamFrameworkIndividualizingPredictions2015a} provide examples of the general model framework used.
+This ultimately yields a nonlinear dimension reduction algorithm that takes advantage of a statistically principled model to address nesting structures of the input data.
+As it is based on the PME framework, the algorithm outputs an explicit representation of the embedding function, making it possible to both generate new data in high-dimensional space using low-dimensional parameters and map new input data to the estimated manifold.
+Smoothing spline analysis of variance allows the statistical comparison of the estimated manifolds between groups or individuals (\cite{wangMixedEffectsSmoothing1998}).
+
+The following sections are organized as follows.
+Section \ref{s:methods} specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper.
+Section 3 applies this method to several simulated datasets using a variety of underlying manifold structures and tests performance on the MNIST dataset of handwritten digits.
+Section 4 demonstrates how the model specified in Section 2 may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset, while Section 5 provides a concluding discussion of the proposed algorithm and opportunities for further study.
+
+\subsection{Methods}\label{s:methods}
+
+\subsubsection{Additive Smoothing Spline Model}
+
+We begin this section with a brief review of additive models.
+Consider a regression scenario where the dependent variable $\mathbf{y} = \left(y_1, \dots, y_n\right)^{T}$ is thought to be associated with independent covariate vectors $\mathbf{x}_i = \left(x_{i1}, \dots, x_{ip}\right)$.
+In this situation, the standard multiple linear regression model assumes that the dependence of $y_i$ on $\mathbf{x}_i$ takes the linear form
+\[%
+  y_i = \alpha + x_{i1}\beta_1 + \dots + x_{ip}\beta_p + \epsilon_i
+,\]%
+where $\epsilon$ has mean 0 and variance $\sigma^2$.
+Additive models are a direct extension of this model, where the additive structure of the model is retained, but the assumption of linearity is discarded.
+A simple additive model takes the form
+\begin{equation}
+  y_i = \alpha + \sum_{j=1}^{p}f_j(X_{ij}) + \epsilon_i, \label{eq:1}
+\end{equation}
+where again $\epsilon_i$ has mean 0 and variance $\sigma^2$, however for each $j$, $f_j$ represents an arbitrary function commonly assumed to be smooth.
+While the functions in model \ref{eq:1} only take as parameters the value of one covariate, with each covariate being represented in the model once, this does not need to be the case.
+The structure of additive models seeks to balance benefits of interpretability found in additive models with the flexibility of nonlinear, and frequently, nonparametric functional forms.
+Further information can be found in \cite{hastieGeneralizedAdditiveModels1990}.
+
+With this background in place, we turn to the motivating example of analyzing the subcortical surfaces of ADNI participants using MR images, with a focus on the hippocampus, although the framework can be applied to modeling other structures in the brain.
+ADNI is a longitudinal observational study, where several imaging modalities are used to record brain images from participants over time.
+Participants can generally be considered to belong to three groups based on cognitive state: elderly healthy controls, those displaying symptoms of mild cognitive impairment, and those diagnosed with AD.
+In practice, participants may transition between these categorizations as their symptoms progress.
+However, because patients may be diagnosed with AD at different stages, accurately modeling these transitions is challenging and is beyond the scope of this manuscript.
+For simplicity, we use the groupings recorded at the time each participant was recruited to the study.
+
+Individual participants are typically scheduled for imaging at six month intervals, though variability exists in terms of the recorded intervals.
+Also of note, the numbers of observations contributed by each participant varies, so groups may be unbalanced.
+We denote the number of groups under consideration using $G$, the number of individuals included in the study as $I$, the number of images available as $N$, and the number of observations available for the $j$th image, with $j = 1, 2, \dots, N$, as $M_j$.
+The individual to which the $j$th image belongs is denoted as $i(j)$, the group of the $i(j)$th individual is represented by $g(j)$, and we use $t_j$ to denote the time at which the $j$th image was recorded, with all times being relative to the $i(j)$th individual's baseline visit.
+Finally, we use $N_t$, $N_i$, and $N_g$ to denote the number of images available for time point $t$, individual $i$, and group $g$, respectively.
+
+Within a segmented structural MRI, the voxels estimated to mark the surface of the hippocampus can be assumed to lie along a two-dimensional manifold that has been wrapped into the three-dimensional image space.
+More generally, we let $d$ denote the intrinsic dimension of the low-dimensional manifold, and $D$ represent the dimension of the high-dimensional input space.
+For the $j$th image with $j = 1, \dots, N$, we denote the $m$th voxel in the $D$-dimensional space as $\mathbf{x}_{jm} \in \mathbb{R}^{D}$, its associated $d$-dimensional parameterization as $\mathbf{y}_{jm} \in \mathbb{R}^{d}$, and its weight as $w_{jm}$.
+We use $\mathbf{x}_g$, $\mathbf{x}_i$, $\mathbf{x}_{j}$, and $\mathbf{x}_t$ to denote the input voxels from all images in the $g$th group, $i$th individual, $j$th image, and $t$th time point, respectively, and use similar notation for all other relevant terms.
+
+The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{jm})$, which can be modeled as 
+\begin{equation}
+  \mathbf{x}_{jm} = \psi(\mathbf{y}_{jm}, t_j) = f_p(\mathbf{y}_{jm}, t_{j}) + f_{g(j)}(\mathbf{y}_{jm}, t_{j}) + f_{i(j)}(\mathbf{y}_{jm}, t_{j}) + f_j(\mathbf{y}_{jm}) + \epsilon_{jm}. \label{eq:2}
+\end{equation}
+The embedding map $\psi(\mathbf{y})$ is associated with a projection index $\kappa_{\psi}(\mathbf{x}): \mathbb{R}^{D} \to \mathbb{R}^{d}$, which, informally, outputs the $d$-dimensional parameterization $\mathbf{y}$ that minimizes $\|\mathbf{x} - \psi(\mathbf{y})\|_{\mathbb{R}^{D}}$ and follows the generalized definition of projection indices given in \cite{mengPrincipalManifoldEstimation2021}.
+
+Each function in this additive model represents a different level of nesting in the data, with $f_p(\mathbf{y}, t)$ estimating the embedding at the population level, and $f_g(\mathbf{y}, t)$, $f_i(\mathbf{y}, t)$, and $f_{j}(\mathbf{y})$ describing deviations from the population estimates specific to the group, individual, and image in question, respectively.
+As estimating longitudinal changes in the underlying manifold may be of interest, $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$ and $f_i(\mathbf{y}, t)$ are each denoted as varying-coefficient models, where the coefficients of each function are dependent on time (\cite{hastieVaryingCoefficientModels1993}).
+Thus, the model is capable of estimating separate longitudinal changes for population, group, and individual estimates.
+In cases where longitudinal changes are not of interest, each varying-coefficient model may be constrained to be held constant with respect to time.
+As the $j$th image is recorded at a single time point, $f_{j}(\mathbf{y}, t)$ may not be dependent on time.
+We specify $f_{j}(\mathbf{y})$ as a smoothing spline, while the functions $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$, and $f_i(\mathbf{y}, t)$ are each varying-coefficient smoothing spline models, estimated by using a smoothing spline to smooth over smoothing spline coefficients calculated at each time point available in the data.
+
+The smoothing spline functions $f(\mathbf{y})$ minimize 
+\[%
+  \mathcal{K}_{\lambda}(f) = \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - f(\mathbf{y}_{jm})\|^2 + \lambda\|\nabla^{\otimes 2}f\|^2
+,\]%
+where $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is given. Thus, $f(\mathbf{y})$ takes the form
+\begin{equation}
+  f_l(\mathbf{y}) = \sum_{j=1}^{N}\sum_{m=1}^{M_j}s_{jm, l} \times \eta_{d}\left(\|\mathbf{y} - \mathbf{y}_{jm}\|\right) + \sum_{p = 1}^{d + 1}\alpha_{p, l} \times \phi_{p}(\mathbf{y}), \quad l = 1, 2, \dots, D, \label{eq:3}
+\end{equation}
+with $\phi_p(\mathbf{y})$ representing the basis of the linear space of polynomials on $\mathbb{R}^{d}$ with degree $\leq 1$ and $\eta_{d}(y) = y^{4-d}\log y$ when $\log y \neq 0$ and $d$ is even, $\eta_{d}(y) = 0$ when $\log y = 0$ and $d$ is even, and $\eta_{d}(y) = y^{4-d}$ when $d$ is odd.
+
+Without accounting for the additive structure of the model, and after concatenating all relevant values into $M = \sum_{j=1}^{N}M_{j}$-dimensional vectors of input values $\mathbf{y}$, weights $\mathbf{w}$, and outputs $\mathbf{x}$, we can estimate $f$ by obtaining the values of $s_l \in \mathbb{R}^{M}$ and $\alpha_l \in \mathbb{R}^{d + 1}$ for $l = 1, \dots, D$ that minimize $\|\mathbf{W}^{\frac{1}{2}}\left(\mathbf{x}_{l} - \mathbf{E}s_l - \mathbf{R}\alpha_l\right)\|^2_{\mathbb{R}^{M}} + \lambda\|\mathbf{E}^{\frac{1}{2}}s_l\|^2_{\mathbb{R}^{M}}$ under the constraint that $\mathbf{R}^{T}s_l = 0$, where
+\begin{itemize}
+  \item $\mathbf{E} = (E_{qr})_{1 \leq q, r, \leq M}$ is defined by $E_{qr} = \eta_{d}\left(\|\mathbf{y}_{q} - \mathbf{y}_{r}\|\right))$.
+  \item $\mathbf{R} = \left(R_{mp}\right)_{1 \leq m \leq M, \ 1 \leq p \leq d + 1}$ is defined by $R_{mp} = \phi_p(\mathbf{y}_{m})$.
+  \item $\mathbf{W} = diag(w_{1}, \dots, w_{M})$.
+\end{itemize}
+
+With $\upsilon_l$ representing Lagrange multipliers, the solutions for $s_l$ and $\alpha_l$ can be found by solving
+
+\begin{equation}
+   \left(\begin{array}{ccc}
+    2\mathbf{EWE} + 2\lambda\mathbf{E} & 2\mathbf{EWR} & \mathbf{R} \\
+    2\mathbf{R}^{T}\mathbf{WE} & 2\mathbf{R}^{T}\mathbf{WR} & \mathbf{0} \\
+    \mathbf{R}^{T} & \mathbf{0} & \mathbf{0}
+  \end{array}\right)\left(
+  \begin{array}{c}
+    s_l \\
+    \alpha_l \\
+    \upsilon_l
+  \end{array}
+  \right) = \left(
+  \begin{array}{c}
+    2\mathbf{EW}\mathbf{x}_l \\
+    2\mathbf{R}^{T}\mathbf{W}\mathbf{x}_l \\
+    \mathbf{0}
+  \end{array}
+  \right). \label{eq:4}
+\end{equation}
+The optimal value of $\lambda$ can be found using cross-validation, where with images being randomly split into $K$ folds, functions may be evaluated by the mean squared error, written as
+\begin{equation}
+  MSE(\lambda) = \frac{1}{K}\sum_{k=1}^{K}\frac{1}{N_u}\sum_{j=1}^{N_u}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\lambda}^{(k)}(\mathbf{y}_{jm})\|^2. \label{eq:5}
+\end{equation}
+Here, $f_{\lambda}^{(k)}(\mathbf{y})$ denotes the estimated spline function given tuning parameter $\lambda$ and with the data for the $k$th fold excluded.
+
+The coefficients and tuning parameter that minimize $\mathcal{K}_{\lambda}(f)$ suffice for estimating $f_{j}(\mathbf{y})$, in which the coefficients are not time-varying.
+To reach estimates for the varying-coefficient models, equation \ref{eq:4} is solved using the data available at each time point within the relevant study level.
+To estimate $f_p(\mathbf{y}, t)$, we use $\left\{\mathbf{x}_t\right\}_{t=1}^T$ and $\left\{\mathbf{y}_t\right\}_{t=1}^T$, yielding fitted function $\left\{\hat{f}_t(\mathbf{y})\right\}_{t=1}^T$.
+Using the maximum and minimum values of $\mathbf{y}_t$ in each dimension, a new $d$-dimensional grid of $M' = \text{max}\left(\left\{M_{j}\right\}_{j = 1}^{N}\right)$ knots, $\left\{\mathbf{y'}_m\right\}_{m=1}^{M'}$, is generated with the knots being evenly spread throughout the parameter space.
+These parameters serve as common knots that allow the spline coefficients to be compared across time points.
+They are then used to compute a new set of corresponding outcome values $\left\{\mathbf{x}'_t\right\}_{t=1}^{T}$.
+The spline functions are reestimated as $\left\{\hat{f}'_t\right\}_{t=1}^{T}$ using $\left\{\mathbf{x}'_t\right\}_{t=1}^{T}$ and $\mathbf{y}'$.
+These newly generated knots are weighted equally, so an unweighted smoothing spline model is used.
+Following \cite{greenSilverman1994}, we let
+
+\begin{itemize}
+  \item $\mathbf{E}' = \left(E'_{qr}\right)_{1 \leq q, r, \leq M'}$ defined by $E'_{qr} = \eta_{d}\left(\|\mathbf{y}'_{q} - \mathbf{y}'_{r}\|\right)$.
+  \item $\mathbf{R}' = \left(R'_{mp}\right)_{1 \leq m \leq M', \ 1 \leq p \leq d + 1}$ is defined by $R'_{mp} = \phi_p(\mathbf{y'}_{m})$.
+\end{itemize}
+
+Then the coefficients that minimize $\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right)^{T}\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right) + \gamma\mathbf{s}_t^{T}\mathbf{E}'\mathbf{s}_t$ can be found by solving
+\begin{equation}
+  \left(
+  \begin{array}{cc}
+    \mathbf{E}' + \lambda_{t}\mathbf{I} & \mathbf{R}'^{T} \\
+    \mathbf{R}' & \mathbf{0}
+  \end{array}
+  \right)\left(
+  \begin{array}{c}
+    \mathbf{s}'_t \\
+    \mathbf{\alpha}'_t
+  \end{array}
+  \right) = \left(
+  \begin{array}{c}
+    \mathbf{x}'_t \\
+    \mathbf{0}
+  \end{array}
+  \right). \label{eq:6}
+\end{equation}
+
+The coefficients found at each time point $t$, $\mathbf{s'}_{t}$ and $\mathbf{\alpha'}_{t}$ are concatenated in matrix $\mathbf{\omega}_{t} = \left(\mathbf{s'}_{t}^T, \mathbf{\alpha'}_{t}^T\right)$.
+This coefficient matrix is then used as the output value in a cubic smoothing spline function using the time points as input values.
+This function takes the form
+
+\[%
+  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i\|t - t_i\|^{3} + \sum_{p=1}^{2}\tau_p \phi(t)_p
+.\]%
+We define matrix $\mathbf{A}$ as $\mathbf{A}_{qr} = \|t_q - t_r\|^{3}$, matrix $\mathbf{T}$ as $\mathbf{T}_{pr} = \phi_p(t_r)$, $\mathbf{\delta} = \left(\delta_1, \dots, \delta_T\right)^{T}$, and $\mathbf{\tau} = \left(\tau_1, \tau_2\right)^{T}$.
+Then, the coefficients of $g_{\gamma}(t)$, given a value of $\gamma$, can be found by solving
+\begin{equation}
+  \left(
+  \begin{array}{cc}
+    \mathbf{A} + \gamma\mathbf{I} & \mathbf{T}^{T} \\
+    \mathbf{T} & \mathbf{0}
+  \end{array}
+  \right)\left(
+  \begin{array}{c}
+    \mathbf{\delta} \\
+    \mathbf{\tau}
+  \end{array}
+  \right) = \left(
+  \begin{array}{c}
+    \mathbf{\omega} \\
+    \mathbf{0}
+  \end{array}
+  \right). \label{eq:7}
+\end{equation}
+
+As with the tuning process for $\lambda$, the optimal value of $\gamma$ can be identified using cross-validation, with either leave-one-out or $k$-folds cross-validation being appropriate depending on the number of time points available.
+If using leave-one-out cross-validation, the mean squared error can be calculated as
+\begin{equation}
+  MSE(\gamma) = \frac{1}{T}\sum_{t=1}^{T}\frac{1}{N_t}\sum_{j=1}^{N_t}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\gamma}^{(t)}(\mathbf{y}_{jm}, t_j)\|^2. \label{eq:8}
+\end{equation}
+We then let $\hat{\gamma}$ be the value of $\gamma$ that minimizes $MSE(\gamma)$.
+
+With a proper estimate of the $\hat{\gamma}$, we may denote the coefficients of $g(t)$ as $\Omega = \left(\delta^{T}, \tau^{T}\right)$, and those estimated by $g(t)$ as $\omega(t) = \left(\mathbf{s}_{\hat{\gamma}}(t)^{T}, \mathbf{\alpha}_{\hat{\gamma}}(t)^{T}\right)^{T}$.
+We use these coefficients in equation \ref{eq:3} in place of the constant coefficients.
+Therefore, the varying-coefficient models have the following outputs:
+\begin{equation}
+  f(\mathbf{y}, t) = \sum_{m=1}^{M'}\mathbf{s}_{\hat{\gamma}, l}(t) \eta_{d}\left(\|\mathbf{y} - \mathbf{y'}_m\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\hat{\gamma}, p, l}(t) \phi_p(\mathbf{y}), \quad l = 1, \dots, D.
+\label{eq:9}
+\end{equation}
+The fitting process for the varying-coefficient smoothing spline model is given formally as Algorithm \ref{alg:varying_spline}.
+
+\SetKwComment{Comment}{/* }{ */}
+
+\RestyleAlgo{ruled}
+\LinesNumbered
+
+\begin{algorithm}
+\caption{Varying-Coefficient Smoothing Spline Model}\label{alg:varying_spline}
+  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j = 1, m = 1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, weights $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and number of cross-validation folds $K$.}
+  \KwResult{Analytic formula of $\hat{f}(\mathbf{y}, t)$, coefficients $\Omega = \left(\mathbf{\delta}^{T}, \mathbf{\tau}^{T}\right)$, and optimal tuning parameter $\hat{\gamma}$.}
+
+\For{$t = 1, 2, \dots, T$} {
+  Set $U_t$ equal to the number of images included in the data at time $t$\;
+  \For{$l = 1, 2, \dots, L$} {
+    Solve equation \ref{eq:4} to estimate $f_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}_t$, $\mathbf{y}_t$, and $w_t$\;
+    \For{$u = 1, 2, \dots, U_t$} {
+      Solve equation \ref{eq:4} to estimate $f^{(-u)}_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}^{(-u)}_t$, $\mathbf{y}^{(-u)}_t$, $w^{(-u)}_t$\;
+      Estimate $MSE(\lambda_{l})_u$ using equation \ref{eq:5}\;
+    }
+    Set $MSE(\lambda_l) = \frac{1}{U}\sum_{u=1}^{U}MSE(\lambda_l)_u$\;
+  }
+  $\hat{\lambda}_t = \arg \min_{\lambda} MSE(\lambda)$\;
+  $f_t(\mathbf{y}) = f_{\hat{\lambda}_t}(\mathbf{y})$\;
+}
+
+  Let $M' = \max(M_{j})$, $\left\{\mathbf{y}'_{m}\right\}_{m=1}^{M'}$ be a grid spanning the range of values of $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
+  \For{$t = 1, 2, \dots, T$} {
+    Compute $\left\{\mathbf{x'}_{m, t}\right\}_{m=1}^{M'} = \left\{f_t(\mathbf{y'}_m)\right\}_{m=1}^{M'}$\;
+    Solve equation \ref{eq:6} to compute $\mathbf{s'}_t$ and $\mathbf{\alpha'}_t$\;
+    Set $\mathbf{\omega}_t = \left[\mathbf{s'}_t^{T}, \mathbf{\alpha'}_t^{T}\right]$\;
+  }
+  \For{$p = 1, 2, \dots, P$} {
+    Compute $g_{\gamma_p}(t)$ by solving equation \ref{eq:7}\;
+    \For{$t = 1, 2, \dots, T$} {
+      Compute $g_{\gamma_p}^{(t)}$ by solving \ref{eq:7}\;
+    }
+    Estimate $MSE(\gamma_p)$ using \ref{eq:8}\;
+  }
+  $\hat{\gamma} = \arg \min_{\gamma} MSE(\gamma)$\;
+  $\hat{f}(\mathbf{y}, t) = f_{\hat{\gamma}}(\mathbf{y}, t)$, where the form of $\hat{f}(\mathbf{y}, t)$ is given in \ref{eq:9}.
+\end{algorithm}
+
+While the procedure outlined in Algorithm \ref{alg:varying_spline} yields an appropriate estimator when working with unnested data, this approach must be adapted to work within the additive model structure of model \ref{eq:2}.
+As discussed in \cite{hastieGeneralizedAdditiveModels1990}, the additive model $\psi(\mathbf{y}, t)$ can be estimated using a backfitting procedure, which uses the model's additive nature to iteratively fit each distinct smoothing spline function until convergence.
+Specifically, we use the fact that, under the model,
+
+\[%
+  f_p(\mathbf{y}_{jm}, t_{j}) = E\left[\mathbf{x}_{jm} - f_{g(j)}(\mathbf{y}_{jm}, t_{j}) - f_{i(j)}(\mathbf{y}_{jm}, t_{j}) - f_{j}(\mathbf{y}_{jm})\right]
+,\]%
+with a similar result holding for each of the smooth functions making up the additive model. Thus, the backfitting algorithm to fit the proposed model is given in Algorithm \ref{alg:backfitting}.
+
+\begin{algorithm}
+\footnotesize
+\caption{Backfitting Algorithm for Additive Spline Model}\label{alg:backfitting}
+  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, weights $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$ and $\left\{\gamma_p\right\}_{p=1}^{P}$, $\epsilon$ and $K$.}
+  \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$.}
+
+  Estimate $f_{p}^{(0)}(\mathbf{y}, t)$ and $\Omega_{p}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_{j}\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+  \For{$g = 1, 2, \dots, G$} {
+    Estimate $f_{g}^{(0)}(\mathbf{y}, t)$ and $\Omega_{g}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{t_j\right\}_{j=1}^{N_g}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $K$\;
+  }
+  \For{$i = 1, 2, \dots, I$} {
+    Estimate $f_{i}^{(0)}(\mathbf{y}, t)$ and $\Omega_{i}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{t_j\right\}_{j=1}^{N_i}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+  }
+  \For{$j = 1, 2, \dots, N$} {
+    Estimate $f_{j}^{(0)}(\mathbf{y})$ and $\omega_{j}^{(0)}$ by solving \ref{eq:4} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{m=1}^{M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{m=1}^{M_j}$, $\left\{w_{jm}\right\}_{m = 1}^{M_j}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $K$\;
+  }
+  $n \gets 0$\;
+  \While{$\max\left(\|\Omega_p^{(n)} - \Omega_{p}^{(n-1)}\|, \left\{\|\Omega_{g}^{(n)} - \Omega_{g}^{(n-1)}\|\right\}_{g=1}^{G}, \left\{\|\Omega_{i}^{(n)} - \Omega_{i}^{(n-1)}\|\right\}_{i=1}^{I}, \left\{\|\omega_{j}^{(n)} - \omega_{j}^{(n-1)}\|\right\}_{j=1}^{N}\right) \geq \epsilon$} {
+    Estimate $f_{p}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{p}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{g(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_j^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_{j}\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+    \For{$g = 1, 2, \dots, G$} {
+      Estimate $f_{g}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{g}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_{j}^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{t_j\right\}_{j=1}^{N_g}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $K$\;
+    }
+    \For{$i = 1, 2, \dots, I$} {
+      Estimate $f_{i}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{i}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_j^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{t_j\right\}_{j=1}^{N_i}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+    }
+    \For{$j = 1, 2, \dots, N$} {
+      Estimate $f_{j}^{(n+1)}(\mathbf{y})$ and $\omega_{j}^{(n+1)}$ by solving \ref{eq:4} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j)\right\}_{m=1}^{M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{m=1}^{M_j}$, $\left\{w_{jm}\right\}_{m = 1}^{M_j}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $K$\;
+    }
+    $n \gets n + 1$\;
+  }
+  $\hat{f}_p \gets f_{p}^{(n)}$, $\left\{\hat{f}_g\right\}_{g = 1}^{G} \gets \left\{f_{g}^{(n)}\right\}_{g = 1}^{G}$, $\left\{\hat{f}_i\right\}_{i=1}^{I} \gets \left\{f_{i}^{(n)}\right\}_{i=1}^{I}$, and $\left\{\hat{f}_j\right\}_{j=1}^{N} \gets \left\{f_j^{(n)}\right\}_{j=1}^{N}$.
+\end{algorithm}
+
+As mentioned previously, this proposed additive spline model structure has several benefits.
+First, it maintains interpretability at each level of the model.
+It is possible to visually inspect the embedding maps of the population and each group and individual to better understand how they interact.
+While the model described here is intended to be applicable to the specific situation involving the modeling of data from the ADNI study, this general modeling approach can be conveniently adapted to a variety of scenarios.
+Additionally, basing our handling of the data's hierarchical structure on a statistical model allows us to assess statistical properties of the estimated manifold, where less statistically principled approaches make this option difficult.
+Of particular interest, \cite{wangMixedEffectsSmoothing1998} and \cite{brumbackSmoothingSplineModels1998} explore connections between smoothing spline models and mixed-effects models.
+The smoothing spline analysis of variance proposed by \cite{wangMixedEffectsSmoothing1998} may be applicable to the model structure proposed here, which would enable a detailed decomposition of the variation in manifold structure by level.
+For the time being, this extension is left to future work.
+
+A clear concern with the use of this procedure in this case is the number of functions to be estimated, which can quickly increase the computational burden of fitting the models.
+However, the nesting of the different levels of the data helps to limit the number of function combinations that must be fit, helping to keep computational requirements manageable.
+\cite{brumbackSmoothingSplineModels1998} used restricted maximum likelihood estimation to reach an estimate of their proposed additive spline model, which may potentially provide computational improvements.
+However, in the case that the model should be modified to meet the needs of a new dataset, the backfitting algorithm will be more easily adaptable.
+
+\subsubsection{Hierarchical Principal Manifold Estimation}
+
+With the structure of the additive spline model established, we now incorporate it into the larger manifold learning process.
+This procedure is based on the Principal Manifold Estimation (PME) algorithm introduced in \cite{mengPrincipalManifoldEstimation2021}, which iterates between estimating a function to embed observations in $D$-dimensional space and reparameterizing those observations in the $d$-dimensional space to estimate the underlying manifold.
+The PME algorithm is largely split into three distinct steps: data reduction, fitting, and tuning.
+The split between these three steps enforces a process by which candidate functions are fit for each in a vector of smoothing values in the fitting stage, then the tuning step is used to select the smoothing value that results in the lowest error, as measured by mean squared distance,
+\[%
+  \mathcal{D}(\hat{f}) = \frac{1}{N}\sum_{i=1}^{N}\|x_i - \hat{f}_\lambda\left(\pi_{\hat{f}_\lambda}(x_i)\right)\|_{\mathbb{R}^{D}}^2
+,\]%
+where $\hat{f}_\lambda$ represents the estimated embedding map given smoothing parameter $\lambda$, $\pi_{\hat{f}_\lambda}(x)$ denotes the projection index that obtains a $d$-dimensional parameterization of $D$-dimensional input $x$, and $N$ represents the total number of observations in the dataset.
+
+While this approach to tuning the model parameters is beneficial because it allows a complete comparison of all parameter values under consideration, the additive model described in equation \ref{eq:2} requires many more parameters, meaning that taking the same tuning approach would result in the need to find an optimal point in a very high-dimensional space.
+This would greatly increase the computational complexity of the algorithm, likely to the point of infeasibility in an practical situation.
+Instead, the proposed algorithm avoids this problem by taking a greedy approach to tuning the model, where the smoothing parameters for each model are chosen individually, rather than simultaneously.
+
+Thus, the hierarchical PME algorithm proposed here is composed of initialization and data reduction and fitting steps.
+The first step uses the PME algorithm to estimate an embedding map from one image.
+The HDMDE algorithm, detailed in \cite{mengPrincipalManifoldEstimation2021}, is used to reduce the data from each image into a set of component centers and associated weights.
+Then, the embedding map estimated using the PME algorithm can be used to provide initial parameterizations of the component centers for every image.
+These initial parameterizations are then used in the fitting step.
+
+Like the PME algorithm, the fitting step uses an iterative approach to alternate between using the parameterizations to estimate an embedding function with the form of the additive spline model in equation \ref{eq:2}, then using this newly generated embedding function to reestimate the most appropriate parameterizations for the reduced set of components identified in the data reduction process.
+Once the error of the embedding function as measured by the sum of squared distances measured at the component centers declines by less than a given threshold $\epsilon$, the iterative procedure stops and the embedding function and associated parameterizations are finalized.
+The details of this process are given in Algorithm \ref{alg:hpme}.
+
+\begin{algorithm}
+\caption{Hierarchical Principal Manifold Estimation Algorithm}\label{alg:hpme}
+  \KwData{Data points $\left\{\mathbf{v}_{jm}\right\}_{j=1, m=1}^{N, M'_j}$ in $\mathbb{R}^{D}$; group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$; a positive integer $N_0 < \min\left(\left\{M'_j\right\}_{j=1}^{N}\right)$; $\alpha, \epsilon, \epsilon^{*} \in (0, 1)$; candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}, \left\{\gamma_p\right\}_{p=1}^{P}$; $itr \geq 1$, which is the maximum number of iterations allowed.}
+  \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$, .}
+  \textbf{Initialization}: Apply PME with input $\left(\left\{\mathbf{v}_{1m}\right\}_{m=1}^{M'_1}, N_0, \alpha, \epsilon, \epsilon^{*}, \left\{\lambda_l\right\}_{l=1}^{L}, itr\right)$ and obtain $f^{(1)}$\;
+  $\psi^{(1)} \gets f^{(1)}$\;
+  \textbf{Data Reduction}: Apply HDMDE algorithm with inputs $\left(\left\{\left\{\mathbf{v}_{jm}\right\}_{m=1}^{M_j}\right\}_{j=1}^{N}, N_0, \epsilon, \alpha\right)$ and obtain $\left\{M_j\right\}_{j=1}^{N}$, $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, and $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
+  \textbf{Parameterization}: Use the projection index of $\psi^{(1)}$, $\kappa_{\psi^{(1)}}$, to parameterize $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$ by $d$-dimensional parameters $\left\{y_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
+  $\mathcal{E} \gets 2 \times \epsilon^{*}$, $n \gets 1$, and $\mathcal{D}(\psi^{(1)}) \gets \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - \psi^{(1)}\left(\kappa_{\psi^{(1)}}(\mathbf{x}_{jm})\right)\|_{\mathbb{R}^{D}}^2$\;
+  \While{$\mathcal{E} \geq \epsilon^{*}$ and $n < itr$} {
+    Compute $f_{p}^{(n + 1)}$, $\left\{f_{g}^{(n + 1)}\right\}_{g = 1}^{G}$, $\left\{f_{i}^{(n + 1)}\right\}_{i=1}^{I}$, and $\left\{f_{j}^{(n+1)}\right\}_{j=1}^{N}$ by using Algorithm \ref{alg:backfitting} with inputs $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\kappa_{\psi^{(n)}}(\mathbf{x}_{jm})\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_j\right\}_{j=1}^{N}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{g(j)\right\}_{j=1}^{N}$, $\left\{i(j)\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $\epsilon$, and $K$\;
+    $\left\{\psi^{(n+1)}(\mathbf{y}, t) \gets f_{p}^{(n + 1)}(\mathbf{y}, t) + f_{g(j)}^{(n+1)}(\mathbf{y}, t) + f_{i(j)}^{(n + 1)}(\mathbf{y}, t) + f_{j}^{(n+1)}(\mathbf{y})\right\}_{j=1}^{N}$\;
+    $\mathcal{D}(\psi^{n+1}) \gets \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - \psi^{n + 1}\left(\kappa_{\psi^{n+1}}(\mathbf{x}_{jm})\right)\|_{\mathbb{R}^{D}}^2$\;
+    $\mathcal{E} \gets |\mathcal{D}(\psi^{(n+1)}) - \mathcal{D}(\psi^{(n)})| / \mathcal{D}(\psi^{(n)})$\;
+    $n \gets n + 1$\;
+  }
+  $\hat{f}_p \gets f_{p}^{(n)}$, $\left\{\hat{f}_{g} \gets f_{g}^{(n)}\right\}_{g=1}^{G}$, $\left\{\hat{f}_{i} \gets f_{i}^{(n)}\right\}_{i=1}^{I}$, $\left\{\hat{f}_j \gets f_{j}^{(n)}\right\}_{j=1}^{N}$.
+\end{algorithm}
+
+\subsection{Proposed Work}
+Further work will be dedicated to testing the proposed approach in several settings.
+First, the algorithm will be tested with a simplified additive model on the MNIST dataset of handwritten images.
+This will provide an opportunity to test the method on a relatively small imaging dataset, serving as a proof of concept and indicating where problems may arise when the algorithm is used with more complex data structures.
+The method will next be applied to a series of simulated datasets that were generated to reflect the hierarchical structure present in the ADNI dataset.
+In both of these cases, the estimates of the hierarchical approach will be compared to alternative methods, with the fit of each method to the data being compared.
+Finally, the hierarchical approach will be applied to structural MRI data from the ADNI dataset.
+Specifically, the algorithm can be used to estimate the manifolds underlying the hippocampus of study participants.
+Of interest in this case will be a comparison of the varying-coefficient models by group, which may provide a visual confirmation that the algorithm's estimates are detecting the most relevant information included in the data.
+
 \newpage

### 2023-10-02 paper/main.tex
Add changes made to oral qualifier proposal

@@
 
-\section*{Abstract}
+\section{Abstract} \label{s:abstract}
 
@@
 Neurodegenerative diseases such as Alzheimer's disease (AD) and Parkinson's disease are highly complex, chronic conditions that exhibit substantial heterogeneity between patients.
-To enable researchers to identify potential biomarkers of progression for these conditions, large multi-site longitudinal observational studies have been conducted that collect information from a wide range of participants over long durations across several modalities, including neuroimaging [Cite ADNI and PPMI studies].
-Neuroimaging data in particular have been shown to be sensitive to developments in the early stages of disease progression, often before diagnosis occurs [Find citation from reading].
+To enable researchers to identify potential biomarkers of progression for these conditions, large multi-site longitudinal observational studies have been conducted that collect information from a wide range of participants over long durations across several modalities, including neuroimaging (\cite{muellerWaysEarlyDiagnosis2005}, \cite{marekParkinsonProgressionMarker2011}).
+Neuroimaging data in particular have been shown to be sensitive to developments in the early stages of disease progression, often before diagnosis occurs (\cite{jackHypotheticalModelDynamic2010}).
 
-While observations from neuroimaging data are typically high-dimensional, it is commonly assumed that the data lie along a low-dimensional manifold [Relevant citations demonstrating this].
+While observations from neuroimaging data are typically high-dimensional, it is commonly assumed that the data lie along a low-dimensional manifold.
 Under this assumption, manifold learning algorithms can be used to recover this underlying manifold and parameterize this high-dimensional input data in the learned low-dimensional space.
@@
 
-In neuroimaging settings, manifold learning algorithms are most often used to estimate the underlying manifold at the population level, with vectorized image intensities being used as the input data [Citations].
+In neuroimaging settings, manifold learning algorithms are most often used to estimate the underlying manifold at the population level, with vectorized image intensities being used as the input data.
 \cite{yueParameterizationWhiteMatter2016} used manifold learning to estimate a low-dimensional parameterization of the corpus callosum at the individual level.
@@
 treatment groups in a clinical trial).
-This goal is particularly relevant in the case of the complex neurodegenerative conditions mentioned above, where the high degree of heterogeneity in patient experiences has led to attempts to identify subgroups of patients based on a range of criteria such as symptom expression, onset, or rate of progression [Citations].
+This goal is particularly relevant in the case of the complex neurodegenerative conditions mentioned above, where the high degree of heterogeneity in patient experiences has led to attempts to identify subgroups of patients based on a range of criteria such as symptom expression, onset, or rate of progression (\cite{ferreiraBiologicalSubtypesAlzheimer2020}, \cite{venutoPredictingAmbulatoryCapacity}).
 Fundamentally, if a low dimensional manifold is thought to effectively represent data related to the patient experience of a neurodegenerative condition, and the experience of that condition differs by group, then the low-dimensional representation of that experience should also differ by group.
@@
 
-\section{Application to Simulated and MNIST Data}
-
-\subsection{Application to Simulated Data}
-
-\subsection{Application to MNIST Data}
-
-\section{Application to ADNI Data}
-
-\section{Discussion}
-
-\section{Aim 2: Principal Manifold Estimation for Hierarchical Data} \label{sec:hpme}
-
-\subsection{Background}
-
-Neurodegenerative diseases such as Alzheimer's disease (AD) and Parkinson's disease are highly complex, chronic conditions that exhibit substantial heterogeneity between patients.
-To enable researchers to identify potential biomarkers of progression for these conditions, large multi-site longitudinal observational studies have been conducted that collect information from a wide range of participants over long durations across several modalities, including neuroimaging (\cite{muellerWaysEarlyDiagnosis2005}, \cite{marekParkinsonProgressionMarker2011}).
-Neuroimaging data in particular have been shown to be sensitive to developments in the early stages of disease progression, often before diagnosis occurs (\cite{jackHypotheticalModelDynamic2010}).
-
-While observations from neuroimaging data are typically high-dimensional, it is commonly assumed that the data lie along a low-dimensional manifold.
-Under this assumption, manifold learning algorithms can be used to recover this underlying manifold and parameterize this high-dimensional input data in the learned low-dimensional space.
-Most commonly-used manifold learning algorithms, including Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), Locally Linear Embedding (\cite{roweisNonlinearDimensionalityReduction2000}), Laplacian Eigenmaps (\cite{belkinLaplacianEigenmapsDimensionality2003}), and Diffusion Maps (\cite{coifmanDiffusionMaps2006}) construct a graph of similarities between observations, which is then used to determine the structure of the manifold.
-\cite{mengPrincipalManifoldEstimation2021} present an alternative principal manifold framework for manifold estimation which extends to higher dimensions the principal curve framework introduced by \cite{hastiePrincipalCurves1989}, in which principal curves are defined as one-dimensional curves that pass through the middle of a high-dimensional dataset.
-
-In neuroimaging settings, manifold learning algorithms are most often used to estimate the underlying manifold at the population level, with vectorized image intensities being used as the input data.
-\cite{yueParameterizationWhiteMatter2016} used manifold learning to estimate a low-dimensional parameterization of the corpus callosum at the individual level.
-However, in many cases, the outcome of interest is framed as a comparison between different groups of individuals (e.g.
-treatment groups in a clinical trial).
-This goal is particularly relevant in the case of the complex neurodegenerative conditions mentioned above, where the high degree of heterogeneity in patient experiences has led to attempts to identify subgroups of patients based on a range of criteria such as symptom expression, onset, or rate of progression (\cite{ferreiraBiologicalSubtypesAlzheimer2020}, \cite{venutoPredictingAmbulatoryCapacity}).
-Fundamentally, if a low dimensional manifold is thought to effectively represent data related to the patient experience of a neurodegenerative condition, and the experience of that condition differs by group, then the low-dimensional representation of that experience should also differ by group.
-
-A method to accurately and efficiently estimate the manifolds underlying data at the group level must take advantage of the grouped structure of observations.
-To develop an estimate of the manifold underlying the hippocampi of participants in the ADNI study, it is necessary to recognize the basic nesting structure of the data, in which brain images are nested within participants possibly belonging to a latent disease subgroup where patients tend to have more related symptom experiences or underlying physiological developments, which are in turn nested within study groups based on Alzheimer's disease status, that are finally nested within the greater study population.
-In this paper, we develop a manifold learning approach that appropriately accounts for this hierarchical structure when modeling the underlying manifolds.
-We further propose an approach for estimation of the underlying manifold at each level of the hierarchy allowing for statistical testing of differences between those low-dimensional manifolds.
-
-There have been attempts to address the hierarchical structure in manifold learning problems.
-\cite{bhatiaHierarchicalManifoldLearning2012} introduced a modification of Laplacian eigenmaps to estimate manifolds from recursively subdivided regions of medical images while enforcing similar alignment of the manifolds estimated from related regions.
-\cite{freicheCharacterizingMyocardialIschemia2022} built on this approach to incorporate multiple imaging modalities and automate the tuning of the algorithm's hyperparameters.
-\cite{gaoHierarchicalManifoldLearning2019} proposed a method based on two-step diffusion maps for hierarchical data with variation over several domains, where diffusion maps are applied first to reduce the dimension of the images for each individual, resulting in individual-specific manifold estimates, then diffusion maps are applied again on the concatenated parameterizations to reduce over the individuals, thus estimating a group-level manifold.
-This approach allows the first step of dimension reduction to capture the most important qualities of each image separately before reducing across subjects, but the two step procedure ensures that the manifold estimated at the individual level has no relation to the manifold estimated at the group level, posing problems for the interpretability of the embedding mappings and comparisons of manifolds between levels.
-
-\cite{guerreroGroupconstrainedManifoldLearning2017} noted that the grouping behavior of observations recorded longitudinally from the same individual adversely impacted the creation of the neighborhood graph between data points when estimating a manifold for the entire population.
-The authors proposed a constraint specifying that, a given observation's collection of nearest neighbors may include at most one point from each individual or other specified grouping.
-This encourages a more even representation of the entire space of observations in the neighborhood graph, and may be applied to any manifold learning algorithm that features a neighborhood graph to specify proximities between observations.
-However, it simply mitigates the effects of the nesting structure on a population-level manifold estimate rather than enabling a detailed comparison of manifold estimates for each group.
-
-In this work, we build on the principal manifold estimation (PME) algorithm introduced by \cite{mengPrincipalManifoldEstimation2021}, which uses an iterative procedure to specify an embedding mapping from the low-dimensional manifold to a high-dimensional space that takes the form of a smoothing spline.
-Drawing from literature in hierarchical and additive models, we replace the smoothing spline in the PME algorithm with an additive smoothing spline model (\cite{gelmanDataAnalysisUsing2007}, \cite{hastieGeneralizedAdditiveModels1990}, \cite{gelmanBayesianDataAnalysis2014}).
-While the specific model structure may change to meet the needs of the application, \cite{brumbackSmoothingSplineModels1998} and \cite{schulamFrameworkIndividualizingPredictions2015a} provide examples of the general model framework used.
-This ultimately yields a nonlinear dimension reduction algorithm that takes advantage of a statistically principled model to address nesting structures of the input data.
-As it is based on the PME framework, the algorithm outputs an explicit representation of the embedding function, making it possible to both generate new data in high-dimensional space using low-dimensional parameters and map new input data to the estimated manifold.
-Smoothing spline analysis of variance allows the statistical comparison of the estimated manifolds between groups or individuals (\cite{wangMixedEffectsSmoothing1998}).
-
-The following sections are organized as follows.
-Section \ref{s:methods} specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper.
-Section 3 applies this method to several simulated datasets using a variety of underlying manifold structures and tests performance on the MNIST dataset of handwritten digits.
-Section 4 demonstrates how the model specified in Section 2 may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset, while Section 5 provides a concluding discussion of the proposed algorithm and opportunities for further study.
-
-\subsection{Methods}\label{s:methods}
-
-\subsubsection{Additive Smoothing Spline Model}
-
-We begin this section with a brief review of additive models.
-Consider a regression scenario where the dependent variable $\mathbf{y} = \left(y_1, \dots, y_n\right)^{T}$ is thought to be associated with independent covariate vectors $\mathbf{x}_i = \left(x_{i1}, \dots, x_{ip}\right)$.
-In this situation, the standard multiple linear regression model assumes that the dependence of $y_i$ on $\mathbf{x}_i$ takes the linear form
-\[%
-  y_i = \alpha + x_{i1}\beta_1 + \dots + x_{ip}\beta_p + \epsilon_i
-,\]%
-where $\epsilon$ has mean 0 and variance $\sigma^2$.
-Additive models are a direct extension of this model, where the additive structure of the model is retained, but the assumption of linearity is discarded.
-A simple additive model takes the form
-\begin{equation}
-  y_i = \alpha + \sum_{j=1}^{p}f_j(X_{ij}) + \epsilon_i, \label{eq:1}
-\end{equation}
-where again $\epsilon_i$ has mean 0 and variance $\sigma^2$, however for each $j$, $f_j$ represents an arbitrary function commonly assumed to be smooth.
-While the functions in model \ref{eq:1} only take as parameters the value of one covariate, with each covariate being represented in the model once, this does not need to be the case.
-The structure of additive models seeks to balance benefits of interpretability found in additive models with the flexibility of nonlinear, and frequently, nonparametric functional forms.
-Further information can be found in \cite{hastieGeneralizedAdditiveModels1990}.
-
-With this background in place, we turn to the motivating example of analyzing the subcortical surfaces of ADNI participants using MR images, with a focus on the hippocampus, although the framework can be applied to modeling other structures in the brain.
-ADNI is a longitudinal observational study, where several imaging modalities are used to record brain images from participants over time.
-Participants can generally be considered to belong to three groups based on cognitive state: elderly healthy controls, those displaying symptoms of mild cognitive impairment, and those diagnosed with AD.
-In practice, participants may transition between these categorizations as their symptoms progress.
-However, because patients may be diagnosed with AD at different stages, accurately modeling these transitions is challenging and is beyond the scope of this manuscript.
-For simplicity, we use the groupings recorded at the time each participant was recruited to the study.
-
-Individual participants are typically scheduled for imaging at six month intervals, though variability exists in terms of the recorded intervals.
-Also of note, the numbers of observations contributed by each participant varies, so groups may be unbalanced.
-We denote the number of groups under consideration using $G$, the number of individuals included in the study as $I$, the number of images available as $N$, and the number of observations available for the $j$th image, with $j = 1, 2, \dots, N$, as $M_j$.
-The individual to which the $j$th image belongs is denoted as $i(j)$, the group of the $i(j)$th individual is represented by $g(j)$, and we use $t_j$ to denote the time at which the $j$th image was recorded, with all times being relative to the $i(j)$th individual's baseline visit.
-Finally, we use $N_t$, $N_i$, and $N_g$ to denote the number of images available for time point $t$, individual $i$, and group $g$, respectively.
-
-Within a segmented structural MRI, the voxels estimated to mark the surface of the hippocampus can be assumed to lie along a two-dimensional manifold that has been wrapped into the three-dimensional image space.
-More generally, we let $d$ denote the intrinsic dimension of the low-dimensional manifold, and $D$ represent the dimension of the high-dimensional input space.
-For the $j$th image with $j = 1, \dots, N$, we denote the $m$th voxel in the $D$-dimensional space as $\mathbf{x}_{jm} \in \mathbb{R}^{D}$, its associated $d$-dimensional parameterization as $\mathbf{y}_{jm} \in \mathbb{R}^{d}$, and its weight as $w_{jm}$.
-We use $\mathbf{x}_g$, $\mathbf{x}_i$, $\mathbf{x}_{j}$, and $\mathbf{x}_t$ to denote the input voxels from all images in the $g$th group, $i$th individual, $j$th image, and $t$th time point, respectively, and use similar notation for all other relevant terms.
-
-The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{jm})$, which can be modeled as 
-\begin{equation}
-  \mathbf{x}_{jm} = \psi(\mathbf{y}_{jm}, t_j) = f_p(\mathbf{y}_{jm}, t_{j}) + f_{g(j)}(\mathbf{y}_{jm}, t_{j}) + f_{i(j)}(\mathbf{y}_{jm}, t_{j}) + f_j(\mathbf{y}_{jm}) + \epsilon_{jm}. \label{eq:2}
-\end{equation}
-The embedding map $\psi(\mathbf{y})$ is associated with a projection index $\kappa_{\psi}(\mathbf{x}): \mathbb{R}^{D} \to \mathbb{R}^{d}$, which, informally, outputs the $d$-dimensional parameterization $\mathbf{y}$ that minimizes $\|\mathbf{x} - \psi(\mathbf{y})\|_{\mathbb{R}^{D}}$ and follows the generalized definition of projection indices given in \cite{mengPrincipalManifoldEstimation2021}.
-
-Each function in this additive model represents a different level of nesting in the data, with $f_p(\mathbf{y}, t)$ estimating the embedding at the population level, and $f_g(\mathbf{y}, t)$, $f_i(\mathbf{y}, t)$, and $f_{j}(\mathbf{y})$ describing deviations from the population estimates specific to the group, individual, and image in question, respectively.
-As estimating longitudinal changes in the underlying manifold may be of interest, $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$ and $f_i(\mathbf{y}, t)$ are each denoted as varying-coefficient models, where the coefficients of each function are dependent on time (\cite{hastieVaryingCoefficientModels1993}).
-Thus, the model is capable of estimating separate longitudinal changes for population, group, and individual estimates.
-In cases where longitudinal changes are not of interest, each varying-coefficient model may be constrained to be held constant with respect to time.
-As the $j$th image is recorded at a single time point, $f_{j}(\mathbf{y}, t)$ may not be dependent on time.
-We specify $f_{j}(\mathbf{y})$ as a smoothing spline, while the functions $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$, and $f_i(\mathbf{y}, t)$ are each varying-coefficient smoothing spline models, estimated by using a smoothing spline to smooth over smoothing spline coefficients calculated at each time point available in the data.
-
-The smoothing spline functions $f(\mathbf{y})$ minimize 
-\[%
-  \mathcal{K}_{\lambda}(f) = \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - f(\mathbf{y}_{jm})\|^2 + \lambda\|\nabla^{\otimes 2}f\|^2
-,\]%
-where $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is given. Thus, $f(\mathbf{y})$ takes the form
-\begin{equation}
-  f_l(\mathbf{y}) = \sum_{j=1}^{N}\sum_{m=1}^{M_j}s_{jm, l} \times \eta_{d}\left(\|\mathbf{y} - \mathbf{y}_{jm}\|\right) + \sum_{p = 1}^{d + 1}\alpha_{p, l} \times \phi_{p}(\mathbf{y}), \quad l = 1, 2, \dots, D, \label{eq:3}
-\end{equation}
-with $\phi_p(\mathbf{y})$ representing the basis of the linear space of polynomials on $\mathbb{R}^{d}$ with degree $\leq 1$ and $\eta_{d}(y) = y^{4-d}\log y$ when $\log y \neq 0$ and $d$ is even, $\eta_{d}(y) = 0$ when $\log y = 0$ and $d$ is even, and $\eta_{d}(y) = y^{4-d}$ when $d$ is odd.
-
-Without accounting for the additive structure of the model, and after concatenating all relevant values into $M = \sum_{j=1}^{N}M_{j}$-dimensional vectors of input values $\mathbf{y}$, weights $\mathbf{w}$, and outputs $\mathbf{x}$, we can estimate $f$ by obtaining the values of $s_l \in \mathbb{R}^{M}$ and $\alpha_l \in \mathbb{R}^{d + 1}$ for $l = 1, \dots, D$ that minimize $\|\mathbf{W}^{\frac{1}{2}}\left(\mathbf{x}_{l} - \mathbf{E}s_l - \mathbf{R}\alpha_l\right)\|^2_{\mathbb{R}^{M}} + \lambda\|\mathbf{E}^{\frac{1}{2}}s_l\|^2_{\mathbb{R}^{M}}$ under the constraint that $\mathbf{R}^{T}s_l = 0$, where
-\begin{itemize}
-  \item $\mathbf{E} = (E_{qr})_{1 \leq q, r, \leq M}$ is defined by $E_{qr} = \eta_{d}\left(\|\mathbf{y}_{q} - \mathbf{y}_{r}\|\right))$.
-  \item $\mathbf{R} = \left(R_{mp}\right)_{1 \leq m \leq M, \ 1 \leq p \leq d + 1}$ is defined by $R_{mp} = \phi_p(\mathbf{y}_{m})$.
-  \item $\mathbf{W} = diag(w_{1}, \dots, w_{M})$.
-\end{itemize}
-
-With $\upsilon_l$ representing Lagrange multipliers, the solutions for $s_l$ and $\alpha_l$ can be found by solving
-
-\begin{equation}
-   \left(\begin{array}{ccc}
-    2\mathbf{EWE} + 2\lambda\mathbf{E} & 2\mathbf{EWR} & \mathbf{R} \\
-    2\mathbf{R}^{T}\mathbf{WE} & 2\mathbf{R}^{T}\mathbf{WR} & \mathbf{0} \\
-    \mathbf{R}^{T} & \mathbf{0} & \mathbf{0}
-  \end{array}\right)\left(
-  \begin{array}{c}
-    s_l \\
-    \alpha_l \\
-    \upsilon_l
-  \end{array}
-  \right) = \left(
-  \begin{array}{c}
-    2\mathbf{EW}\mathbf{x}_l \\
-    2\mathbf{R}^{T}\mathbf{W}\mathbf{x}_l \\
-    \mathbf{0}
-  \end{array}
-  \right). \label{eq:4}
-\end{equation}
-The optimal value of $\lambda$ can be found using cross-validation, where with images being randomly split into $K$ folds, functions may be evaluated by the mean squared error, written as
-\begin{equation}
-  MSE(\lambda) = \frac{1}{K}\sum_{k=1}^{K}\frac{1}{N_u}\sum_{j=1}^{N_u}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\lambda}^{(k)}(\mathbf{y}_{jm})\|^2. \label{eq:5}
-\end{equation}
-Here, $f_{\lambda}^{(k)}(\mathbf{y})$ denotes the estimated spline function given tuning parameter $\lambda$ and with the data for the $k$th fold excluded.
-
-The coefficients and tuning parameter that minimize $\mathcal{K}_{\lambda}(f)$ suffice for estimating $f_{j}(\mathbf{y})$, in which the coefficients are not time-varying.
-To reach estimates for the varying-coefficient models, equation \ref{eq:4} is solved using the data available at each time point within the relevant study level.
-To estimate $f_p(\mathbf{y}, t)$, we use $\left\{\mathbf{x}_t\right\}_{t=1}^T$ and $\left\{\mathbf{y}_t\right\}_{t=1}^T$, yielding fitted function $\left\{\hat{f}_t(\mathbf{y})\right\}_{t=1}^T$.
-Using the maximum and minimum values of $\mathbf{y}_t$ in each dimension, a new $d$-dimensional grid of $M' = \text{max}\left(\left\{M_{j}\right\}_{j = 1}^{N}\right)$ knots, $\left\{\mathbf{y'}_m\right\}_{m=1}^{M'}$, is generated with the knots being evenly spread throughout the parameter space.
-These parameters serve as common knots that allow the spline coefficients to be compared across time points.
-They are then used to compute a new set of corresponding outcome values $\left\{\mathbf{x}'_t\right\}_{t=1}^{T}$.
-The spline functions are reestimated as $\left\{\hat{f}'_t\right\}_{t=1}^{T}$ using $\left\{\mathbf{x}'_t\right\}_{t=1}^{T}$ and $\mathbf{y}'$.
-These newly generated knots are weighted equally, so an unweighted smoothing spline model is used.
-Following \cite{greenSilverman1994}, we let
-
-\begin{itemize}
-  \item $\mathbf{E}' = \left(E'_{qr}\right)_{1 \leq q, r, \leq M'}$ defined by $E'_{qr} = \eta_{d}\left(\|\mathbf{y}'_{q} - \mathbf{y}'_{r}\|\right)$.
-  \item $\mathbf{R}' = \left(R'_{mp}\right)_{1 \leq m \leq M', \ 1 \leq p \leq d + 1}$ is defined by $R'_{mp} = \phi_p(\mathbf{y'}_{m})$.
-\end{itemize}
-
-Then the coefficients that minimize $\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right)^{T}\left(\mathbf{x}'_t - \mathbf{E}'\mathbf{s}_t - \mathbf{R}'^{T}\mathbf{\alpha}_t\right) + \gamma\mathbf{s}_t^{T}\mathbf{E}'\mathbf{s}_t$ can be found by solving
-\begin{equation}
-  \left(
-  \begin{array}{cc}
-    \mathbf{E}' + \lambda_{t}\mathbf{I} & \mathbf{R}'^{T} \\
-    \mathbf{R}' & \mathbf{0}
-  \end{array}
-  \right)\left(
-  \begin{array}{c}
-    \mathbf{s}'_t \\
-    \mathbf{\alpha}'_t
-  \end{array}
-  \right) = \left(
-  \begin{array}{c}
-    \mathbf{x}'_t \\
-    \mathbf{0}
-  \end{array}
-  \right). \label{eq:6}
-\end{equation}
-
-The coefficients found at each time point $t$, $\mathbf{s'}_{t}$ and $\mathbf{\alpha'}_{t}$ are concatenated in matrix $\mathbf{\omega}_{t} = \left(\mathbf{s'}_{t}^T, \mathbf{\alpha'}_{t}^T\right)$.
-This coefficient matrix is then used as the output value in a cubic smoothing spline function using the time points as input values.
-This function takes the form
-
-\[%
-  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i\|t - t_i\|^{3} + \sum_{p=1}^{2}\tau_p \phi(t)_p
-.\]%
-We define matrix $\mathbf{A}$ as $\mathbf{A}_{qr} = \|t_q - t_r\|^{3}$, matrix $\mathbf{T}$ as $\mathbf{T}_{pr} = \phi_p(t_r)$, $\mathbf{\delta} = \left(\delta_1, \dots, \delta_T\right)^{T}$, and $\mathbf{\tau} = \left(\tau_1, \tau_2\right)^{T}$.
-Then, the coefficients of $g_{\gamma}(t)$, given a value of $\gamma$, can be found by solving
-\begin{equation}
-  \left(
-  \begin{array}{cc}
-    \mathbf{A} + \gamma\mathbf{I} & \mathbf{T}^{T} \\
-    \mathbf{T} & \mathbf{0}
-  \end{array}
-  \right)\left(
-  \begin{array}{c}
-    \mathbf{\delta} \\
-    \mathbf{\tau}
-  \end{array}
-  \right) = \left(
-  \begin{array}{c}
-    \mathbf{\omega} \\
-    \mathbf{0}
-  \end{array}
-  \right). \label{eq:7}
-\end{equation}
-
-As with the tuning process for $\lambda$, the optimal value of $\gamma$ can be identified using cross-validation, with either leave-one-out or $k$-folds cross-validation being appropriate depending on the number of time points available.
-If using leave-one-out cross-validation, the mean squared error can be calculated as
-\begin{equation}
-  MSE(\gamma) = \frac{1}{T}\sum_{t=1}^{T}\frac{1}{N_t}\sum_{j=1}^{N_t}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\gamma}^{(t)}(\mathbf{y}_{jm}, t_j)\|^2. \label{eq:8}
-\end{equation}
-We then let $\hat{\gamma}$ be the value of $\gamma$ that minimizes $MSE(\gamma)$.
-
-With a proper estimate of the $\hat{\gamma}$, we may denote the coefficients of $g(t)$ as $\Omega = \left(\delta^{T}, \tau^{T}\right)$, and those estimated by $g(t)$ as $\omega(t) = \left(\mathbf{s}_{\hat{\gamma}}(t)^{T}, \mathbf{\alpha}_{\hat{\gamma}}(t)^{T}\right)^{T}$.
-We use these coefficients in equation \ref{eq:3} in place of the constant coefficients.
-Therefore, the varying-coefficient models have the following outputs:
-\begin{equation}
-  f(\mathbf{y}, t) = \sum_{m=1}^{M'}\mathbf{s}_{\hat{\gamma}, l}(t) \eta_{d}\left(\|\mathbf{y} - \mathbf{y'}_m\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\hat{\gamma}, p, l}(t) \phi_p(\mathbf{y}), \quad l = 1, \dots, D.
-\label{eq:9}
-\end{equation}
-The fitting process for the varying-coefficient smoothing spline model is given formally as Algorithm \ref{alg:varying_spline}.
-
-\SetKwComment{Comment}{/* }{ */}
-
-\RestyleAlgo{ruled}
-\LinesNumbered
-
-\begin{algorithm}
-\caption{Varying-Coefficient Smoothing Spline Model}\label{alg:varying_spline}
-  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j = 1, m = 1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, weights $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and number of cross-validation folds $K$.}
-  \KwResult{Analytic formula of $\hat{f}(\mathbf{y}, t)$, coefficients $\Omega = \left(\mathbf{\delta}^{T}, \mathbf{\tau}^{T}\right)$, and optimal tuning parameter $\hat{\gamma}$.}
-
-\For{$t = 1, 2, \dots, T$} {
-  Set $U_t$ equal to the number of images included in the data at time $t$\;
-  \For{$l = 1, 2, \dots, L$} {
-    Solve equation \ref{eq:4} to estimate $f_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}_t$, $\mathbf{y}_t$, and $w_t$\;
-    \For{$u = 1, 2, \dots, U_t$} {
-      Solve equation \ref{eq:4} to estimate $f^{(-u)}_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}^{(-u)}_t$, $\mathbf{y}^{(-u)}_t$, $w^{(-u)}_t$\;
-      Estimate $MSE(\lambda_{l})_u$ using equation \ref{eq:5}\;
-    }
-    Set $MSE(\lambda_l) = \frac{1}{U}\sum_{u=1}^{U}MSE(\lambda_l)_u$\;
-  }
-  $\hat{\lambda}_t = \arg \min_{\lambda} MSE(\lambda)$\;
-  $f_t(\mathbf{y}) = f_{\hat{\lambda}_t}(\mathbf{y})$\;
-}
-
-  Let $M' = \max(M_{j})$, $\left\{\mathbf{y}'_{m}\right\}_{m=1}^{M'}$ be a grid spanning the range of values of $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
-  \For{$t = 1, 2, \dots, T$} {
-    Compute $\left\{\mathbf{x'}_{m, t}\right\}_{m=1}^{M'} = \left\{f_t(\mathbf{y'}_m)\right\}_{m=1}^{M'}$\;
-    Solve equation \ref{eq:6} to compute $\mathbf{s'}_t$ and $\mathbf{\alpha'}_t$\;
-    Set $\mathbf{\omega}_t = \left[\mathbf{s'}_t^{T}, \mathbf{\alpha'}_t^{T}\right]$\;
-  }
-  \For{$p = 1, 2, \dots, P$} {
-    Compute $g_{\gamma_p}(t)$ by solving equation \ref{eq:7}\;
-    \For{$t = 1, 2, \dots, T$} {
-      Compute $g_{\gamma_p}^{(t)}$ by solving \ref{eq:7}\;
-    }
-    Estimate $MSE(\gamma_p)$ using \ref{eq:8}\;
-  }
-  $\hat{\gamma} = \arg \min_{\gamma} MSE(\gamma)$\;
-  $\hat{f}(\mathbf{y}, t) = f_{\hat{\gamma}}(\mathbf{y}, t)$, where the form of $\hat{f}(\mathbf{y}, t)$ is given in \ref{eq:9}.
-\end{algorithm}
-
-While the procedure outlined in Algorithm \ref{alg:varying_spline} yields an appropriate estimator when working with unnested data, this approach must be adapted to work within the additive model structure of model \ref{eq:2}.
-As discussed in \cite{hastieGeneralizedAdditiveModels1990}, the additive model $\psi(\mathbf{y}, t)$ can be estimated using a backfitting procedure, which uses the model's additive nature to iteratively fit each distinct smoothing spline function until convergence.
-Specifically, we use the fact that, under the model,
-
-\[%
-  f_p(\mathbf{y}_{jm}, t_{j}) = E\left[\mathbf{x}_{jm} - f_{g(j)}(\mathbf{y}_{jm}, t_{j}) - f_{i(j)}(\mathbf{y}_{jm}, t_{j}) - f_{j}(\mathbf{y}_{jm})\right]
-,\]%
-with a similar result holding for each of the smooth functions making up the additive model. Thus, the backfitting algorithm to fit the proposed model is given in Algorithm \ref{alg:backfitting}.
-
-\begin{algorithm}
-\footnotesize
-\caption{Backfitting Algorithm for Additive Spline Model}\label{alg:backfitting}
-  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, weights $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$ and $\left\{\gamma_p\right\}_{p=1}^{P}$, $\epsilon$ and $K$.}
-  \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$.}
-
-  Estimate $f_{p}^{(0)}(\mathbf{y}, t)$ and $\Omega_{p}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_{j}\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
-  \For{$g = 1, 2, \dots, G$} {
-    Estimate $f_{g}^{(0)}(\mathbf{y}, t)$ and $\Omega_{g}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{t_j\right\}_{j=1}^{N_g}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $K$\;
-  }
-  \For{$i = 1, 2, \dots, I$} {
-    Estimate $f_{i}^{(0)}(\mathbf{y}, t)$ and $\Omega_{i}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{t_j\right\}_{j=1}^{N_i}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
-  }
-  \For{$j = 1, 2, \dots, N$} {
-    Estimate $f_{j}^{(0)}(\mathbf{y})$ and $\omega_{j}^{(0)}$ by solving \ref{eq:4} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{m=1}^{M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{m=1}^{M_j}$, $\left\{w_{jm}\right\}_{m = 1}^{M_j}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $K$\;
-  }
-  $n \gets 0$\;
-  \While{$\max\left(\|\Omega_p^{(n)} - \Omega_{p}^{(n-1)}\|, \left\{\|\Omega_{g}^{(n)} - \Omega_{g}^{(n-1)}\|\right\}_{g=1}^{G}, \left\{\|\Omega_{i}^{(n)} - \Omega_{i}^{(n-1)}\|\right\}_{i=1}^{I}, \left\{\|\omega_{j}^{(n)} - \omega_{j}^{(n-1)}\|\right\}_{j=1}^{N}\right) \geq \epsilon$} {
-    Estimate $f_{p}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{p}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{g(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_j^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_{j}\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
-    \For{$g = 1, 2, \dots, G$} {
-      Estimate $f_{g}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{g}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_{j}^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{t_j\right\}_{j=1}^{N_g}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $K$\;
-    }
-    \For{$i = 1, 2, \dots, I$} {
-      Estimate $f_{i}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{i}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_j^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{t_j\right\}_{j=1}^{N_i}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
-    }
-    \For{$j = 1, 2, \dots, N$} {
-      Estimate $f_{j}^{(n+1)}(\mathbf{y})$ and $\omega_{j}^{(n+1)}$ by solving \ref{eq:4} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j)\right\}_{m=1}^{M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{m=1}^{M_j}$, $\left\{w_{jm}\right\}_{m = 1}^{M_j}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $K$\;
-    }
-    $n \gets n + 1$\;
-  }
-  $\hat{f}_p \gets f_{p}^{(n)}$, $\left\{\hat{f}_g\right\}_{g = 1}^{G} \gets \left\{f_{g}^{(n)}\right\}_{g = 1}^{G}$, $\left\{\hat{f}_i\right\}_{i=1}^{I} \gets \left\{f_{i}^{(n)}\right\}_{i=1}^{I}$, and $\left\{\hat{f}_j\right\}_{j=1}^{N} \gets \left\{f_j^{(n)}\right\}_{j=1}^{N}$.
-\end{algorithm}
-
-As mentioned previously, this proposed additive spline model structure has several benefits.
-First, it maintains interpretability at each level of the model.
-It is possible to visually inspect the embedding maps of the population and each group and individual to better understand how they interact.
-While the model described here is intended to be applicable to the specific situation involving the modeling of data from the ADNI study, this general modeling approach can be conveniently adapted to a variety of scenarios.
-Additionally, basing our handling of the data's hierarchical structure on a statistical model allows us to assess statistical properties of the estimated manifold, where less statistically principled approaches make this option difficult.
-Of particular interest, \cite{wangMixedEffectsSmoothing1998} and \cite{brumbackSmoothingSplineModels1998} explore connections between smoothing spline models and mixed-effects models.
-The smoothing spline analysis of variance proposed by \cite{wangMixedEffectsSmoothing1998} may be applicable to the model structure proposed here, which would enable a detailed decomposition of the variation in manifold structure by level.
-For the time being, this extension is left to future work.
-
-A clear concern with the use of this procedure in this case is the number of functions to be estimated, which can quickly increase the computational burden of fitting the models.
-However, the nesting of the different levels of the data helps to limit the number of function combinations that must be fit, helping to keep computational requirements manageable.
-\cite{brumbackSmoothingSplineModels1998} used restricted maximum likelihood estimation to reach an estimate of their proposed additive spline model, which may potentially provide computational improvements.
-However, in the case that the model should be modified to meet the needs of a new dataset, the backfitting algorithm will be more easily adaptable.
-
-\subsubsection{Hierarchical Principal Manifold Estimation}
-
-With the structure of the additive spline model established, we now incorporate it into the larger manifold learning process.
-This procedure is based on the Principal Manifold Estimation (PME) algorithm introduced in \cite{mengPrincipalManifoldEstimation2021}, which iterates between estimating a function to embed observations in $D$-dimensional space and reparameterizing those observations in the $d$-dimensional space to estimate the underlying manifold.
-The PME algorithm is largely split into three distinct steps: data reduction, fitting, and tuning.
-The split between these three steps enforces a process by which candidate functions are fit for each in a vector of smoothing values in the fitting stage, then the tuning step is used to select the smoothing value that results in the lowest error, as measured by mean squared distance,
-\[%
-  \mathcal{D}(\hat{f}) = \frac{1}{N}\sum_{i=1}^{N}\|x_i - \hat{f}_\lambda\left(\pi_{\hat{f}_\lambda}(x_i)\right)\|_{\mathbb{R}^{D}}^2
-,\]%
-where $\hat{f}_\lambda$ represents the estimated embedding map given smoothing parameter $\lambda$, $\pi_{\hat{f}_\lambda}(x)$ denotes the projection index that obtains a $d$-dimensional parameterization of $D$-dimensional input $x$, and $N$ represents the total number of observations in the dataset.
-
-While this approach to tuning the model parameters is beneficial because it allows a complete comparison of all parameter values under consideration, the additive model described in equation \ref{eq:2} requires many more parameters, meaning that taking the same tuning approach would result in the need to find an optimal point in a very high-dimensional space.
-This would greatly increase the computational complexity of the algorithm, likely to the point of infeasibility in an practical situation.
-Instead, the proposed algorithm avoids this problem by taking a greedy approach to tuning the model, where the smoothing parameters for each model are chosen individually, rather than simultaneously.
-
-Thus, the hierarchical PME algorithm proposed here is composed of initialization and data reduction and fitting steps.
-The first step uses the PME algorithm to estimate an embedding map from one image.
-The HDMDE algorithm, detailed in \cite{mengPrincipalManifoldEstimation2021}, is used to reduce the data from each image into a set of component centers and associated weights.
-Then, the embedding map estimated using the PME algorithm can be used to provide initial parameterizations of the component centers for every image.
-These initial parameterizations are then used in the fitting step.
-
-Like the PME algorithm, the fitting step uses an iterative approach to alternate between using the parameterizations to estimate an embedding function with the form of the additive spline model in equation \ref{eq:2}, then using this newly generated embedding function to reestimate the most appropriate parameterizations for the reduced set of components identified in the data reduction process.
-Once the error of the embedding function as measured by the sum of squared distances measured at the component centers declines by less than a given threshold $\epsilon$, the iterative procedure stops and the embedding function and associated parameterizations are finalized.
-The details of this process are given in Algorithm \ref{alg:hpme}.
-
-\begin{algorithm}
-\caption{Hierarchical Principal Manifold Estimation Algorithm}\label{alg:hpme}
-  \KwData{Data points $\left\{\mathbf{v}_{jm}\right\}_{j=1, m=1}^{N, M'_j}$ in $\mathbb{R}^{D}$; group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$; a positive integer $N_0 < \min\left(\left\{M'_j\right\}_{j=1}^{N}\right)$; $\alpha, \epsilon, \epsilon^{*} \in (0, 1)$; candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}, \left\{\gamma_p\right\}_{p=1}^{P}$; $itr \geq 1$, which is the maximum number of iterations allowed.}
-  \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$, .}
-  \textbf{Initialization}: Apply PME with input $\left(\left\{\mathbf{v}_{1m}\right\}_{m=1}^{M'_1}, N_0, \alpha, \epsilon, \epsilon^{*}, \left\{\lambda_l\right\}_{l=1}^{L}, itr\right)$ and obtain $f^{(1)}$\;
-  $\psi^{(1)} \gets f^{(1)}$\;
-  \textbf{Data Reduction}: Apply HDMDE algorithm with inputs $\left(\left\{\left\{\mathbf{v}_{jm}\right\}_{m=1}^{M_j}\right\}_{j=1}^{N}, N_0, \epsilon, \alpha\right)$ and obtain $\left\{M_j\right\}_{j=1}^{N}$, $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, and $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
-  \textbf{Parameterization}: Use the projection index of $\psi^{(1)}$, $\kappa_{\psi^{(1)}}$, to parameterize $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$ by $d$-dimensional parameters $\left\{y_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
-  $\mathcal{E} \gets 2 \times \epsilon^{*}$, $n \gets 1$, and $\mathcal{D}(\psi^{(1)}) \gets \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - \psi^{(1)}\left(\kappa_{\psi^{(1)}}(\mathbf{x}_{jm})\right)\|_{\mathbb{R}^{D}}^2$\;
-  \While{$\mathcal{E} \geq \epsilon^{*}$ and $n < itr$} {
-    Compute $f_{p}^{(n + 1)}$, $\left\{f_{g}^{(n + 1)}\right\}_{g = 1}^{G}$, $\left\{f_{i}^{(n + 1)}\right\}_{i=1}^{I}$, and $\left\{f_{j}^{(n+1)}\right\}_{j=1}^{N}$ by using Algorithm \ref{alg:backfitting} with inputs $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\kappa_{\psi^{(n)}}(\mathbf{x}_{jm})\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_j\right\}_{j=1}^{N}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{g(j)\right\}_{j=1}^{N}$, $\left\{i(j)\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $\epsilon$, and $K$\;
-    $\left\{\psi^{(n+1)}(\mathbf{y}, t) \gets f_{p}^{(n + 1)}(\mathbf{y}, t) + f_{g(j)}^{(n+1)}(\mathbf{y}, t) + f_{i(j)}^{(n + 1)}(\mathbf{y}, t) + f_{j}^{(n+1)}(\mathbf{y})\right\}_{j=1}^{N}$\;
-    $\mathcal{D}(\psi^{n+1}) \gets \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - \psi^{n + 1}\left(\kappa_{\psi^{n+1}}(\mathbf{x}_{jm})\right)\|_{\mathbb{R}^{D}}^2$\;
-    $\mathcal{E} \gets |\mathcal{D}(\psi^{(n+1)}) - \mathcal{D}(\psi^{(n)})| / \mathcal{D}(\psi^{(n)})$\;
-    $n \gets n + 1$\;
-  }
-  $\hat{f}_p \gets f_{p}^{(n)}$, $\left\{\hat{f}_{g} \gets f_{g}^{(n)}\right\}_{g=1}^{G}$, $\left\{\hat{f}_{i} \gets f_{i}^{(n)}\right\}_{i=1}^{I}$, $\left\{\hat{f}_j \gets f_{j}^{(n)}\right\}_{j=1}^{N}$.
-\end{algorithm}
-
-\subsection{Proposed Work}
+\section{Proposed Work}
 Further work will be dedicated to testing the proposed approach in several settings.

### 2023-10-02 paper/main.tex
introduction edits

@@
 
-\section{Abstract} \label{s:abstract}
+\section*{Abstract} \label{s:abstract}
 
-\section{Introduction}
+\section{Introduction} \label{s:hpme_intro}
 
@@
 While observations from neuroimaging data are typically high-dimensional, it is commonly assumed that the data lie along a low-dimensional manifold.
-Under this assumption, manifold learning algorithms can be used to recover this underlying manifold and parameterize this high-dimensional input data in the learned low-dimensional space.
+Under this assumption, manifold learning algorithms can be used to recover this underlying manifold and parameterize the high-dimensional input data in the learned low-dimensional space.
 Most commonly-used manifold learning algorithms, including Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), Locally Linear Embedding (\cite{roweisNonlinearDimensionalityReduction2000}), Laplacian Eigenmaps (\cite{belkinLaplacianEigenmapsDimensionality2003}), and Diffusion Maps (\cite{coifmanDiffusionMaps2006}) construct a graph of similarities between observations, which is then used to determine the structure of the manifold.
@@
 
-In neuroimaging settings, manifold learning algorithms are most often used to estimate the underlying manifold at the population level, with vectorized image intensities being used as the input data.
+In neuroimaging settings, manifold learning algorithms are most often applied at the population level, with vectorized image intensities being used as the input data.
 \cite{yueParameterizationWhiteMatter2016} used manifold learning to estimate a low-dimensional parameterization of the corpus callosum at the individual level.
@@
 A method to accurately and efficiently estimate the manifolds underlying data at the group level must take advantage of the grouped structure of observations.
-To develop an estimate of the manifold underlying the hippocampi of participants in the ADNI study, it is necessary to recognize the basic nesting structure of the data, in which brain images are nested within participants possibly belonging to a latent disease subgroup where patients tend to have more related symptom experiences or underlying physiological developments, which are in turn nested within study groups based on Alzheimer's disease status, that are finally nested within the greater study population.
-In this paper, we develop a manifold learning approach that appropriately accounts for this hierarchical structure when modeling the underlying manifolds.
-We further propose an approach for estimation of the underlying manifold at each level of the hierarchy allowing for statistical testing of differences between those low-dimensional manifolds.
+To develop an estimate of the manifold underlying the hippocampi of participants in the Alzheimer's Disease Neuroimaging Initiative (ADNI) study, it is necessary to recognize the basic nesting structure of the data, in which brain images are nested within participants belong to a latent disease subgroup where patients tend to have more related symptom experiences or underlying physiological developments.
+These subgroups are in turn nested within study groups based on Alzheimer's disease status, which are finally nested within the greater study population.
+In this paper, we develop a manifold learning approach that appropriately accounts for this hierarchical structure when estimating the manifolds underlying such data.
+We further propose an approach for estimation of the underlying manifold at each level of the hierarchy, allowing for statistical testing of differences between those low-dimensional manifolds.
 
-There have been attempts to address the hierarchical structure in manifold learning problems.
+There have been attempts to address hierarchical structure in manifold learning problems.
 \cite{bhatiaHierarchicalManifoldLearning2012} introduced a modification of Laplacian eigenmaps to estimate manifolds from recursively subdivided regions of medical images while enforcing similar alignment of the manifolds estimated from related regions.
 \cite{freicheCharacterizingMyocardialIschemia2022} built on this approach to incorporate multiple imaging modalities and automate the tuning of the algorithm's hyperparameters.
-\cite{gaoHierarchicalManifoldLearning2019} proposed a method based on two-step diffusion maps for hierarchical data with variation over several domains, where diffusion maps are applied first to reduce the dimension of the images for each individual, resulting in individual-specific manifold estimates, then diffusion maps are applied again on the concatenated parameterizations to reduce over the individuals, thus estimating a group-level manifold.
-This approach allows the first step of dimension reduction to capture the most important qualities of each image separately before reducing across subjects, but the two step procedure ensures that the manifold estimated at the individual level has no relation to the manifold estimated at the group level, posing problems for the interpretability of the embedding mappings and comparisons of manifolds between levels.
+\cite{gaoHierarchicalManifoldLearning2019} proposed a two-step diffusion maps method for hierarchical data with variation over several domains, where diffusion maps is applied first to reduce the dimension of the images for each individual, resulting in individual-specific manifold estimates, then diffusion maps is applied again on the concatenated parameterizations to reduce over the individuals, thus estimating a group-level manifold.
+This approach allows the first step of dimension reduction to capture the most important qualities of each image separately before reducing across subjects, but the two step procedure does not ensure that the manifold estimated at the individual level is structurally related to the manifold estimated at the group level, posing problems for the interpretability of the embedding mappings and comparisons of manifolds between levels.
 
-\cite{guerreroGroupconstrainedManifoldLearning2017} noted that the grouping behavior of observations recorded longitudinally from the same individual adversely impacted the creation of the neighborhood graph between data points when estimating a manifold for the entire population.
-The authors proposed a constraint specifying that, a given observation's collection of nearest neighbors may include at most one point from each individual or other specified grouping.
-This encourages a more even representation of the entire space of observations in the neighborhood graph, and may be applied to any manifold learning algorithm that features a neighborhood graph to specify proximities between observations.
+\cite{guerreroGroupconstrainedManifoldLearning2017} noted that, when using graph-based embedding methods to estimate a manifold from population data, the grouping behavior of observations recorded longitudinally from the same individual adversely impacted the creation of the neighborhood graph between data points.
+The authors proposed a constraint specifying that a given observation's collection of nearest neighbors may include at most one point from each individual or other given grouping.
+This encourages a more even representation of the entire space of observations in the neighborhood graph, and may be applied to any manifold learning algorithm that features a neighborhood graph to represent proximities between observations.
 However, it simply mitigates the effects of the nesting structure on a population-level manifold estimate rather than enabling a detailed comparison of manifold estimates for each group.
@@
 Drawing from literature in hierarchical and additive models, we replace the smoothing spline in the PME algorithm with an additive smoothing spline model (\cite{gelmanDataAnalysisUsing2007}, \cite{hastieGeneralizedAdditiveModels1990}, \cite{gelmanBayesianDataAnalysis2014}).
-While the specific model structure may change to meet the needs of the application, \cite{brumbackSmoothingSplineModels1998} and \cite{schulamFrameworkIndividualizingPredictions2015a} provide examples of the general model framework used.
-This ultimately yields a nonlinear dimension reduction algorithm that takes advantage of a statistically principled model to address nesting structures of the input data.
+While the specific model structure may change to meet the needs of the application, \cite{brumbackSmoothingSplineModels1998} and \cite{schulamFrameworkIndividualizingPredictions2015a} provide examples of additive spline models used for different applications.
+The use of additive spline models ultimately yields a nonlinear dimension reduction algorithm that takes advantage of a statistically principled model to address nesting structures of the input data.
 As it is based on the PME framework, the algorithm outputs an explicit representation of the embedding function, making it possible to both generate new data in high-dimensional space using low-dimensional parameters and map new input data to the estimated manifold.
-Smoothing spline analysis of variance allows the statistical comparison of the estimated manifolds between groups or individuals (\cite{wangMixedEffectsSmoothing1998}).
+Smoothing spline analysis of variance may allow the statistical comparison of the estimated manifolds between groups or individuals (\cite{wangMixedEffectsSmoothing1998}).
 
@@
 Section \ref{s:methods} specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper.
-Section 3 applies this method to several simulated datasets using a variety of underlying manifold structures and tests performance on the MNIST dataset of handwritten digits.
-Section 4 demonstrates how the model specified in Section 2 may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset, while Section 5 provides a concluding discussion of the proposed algorithm and opportunities for further study.
+Section \ref{s:hpme_mnist_simulation} applies this method to the MNIST dataset of handwritten digits, and compares the results to those produced using PME.
+The performance of the two approaches is then compared when they are applied to a series of simulated datasets of varying complexities.
+Section \ref{s:hpme_adni} discusses proposed work, which will demonstrate how the model specified in Section \ref{s:methods} may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset, while Section \ref{s:hpme_discussion} provides a concluding discussion of the proposed algorithm and opportunities for further study.
 
@@
 
-\section{Proposed Work}
+\section{Application to MNIST Data and Simulations}\label{s:hpme_mnist_simulation}
+
+\section{Proposed Work} \label{s:hpme_adni}
 Further work will be dedicated to testing the proposed approach in several settings.
@@
 
+\section{Discussion} \label{s:hpme_discussion}
+
 \newpage

### 2023-10-03 paper/main.tex
edits

@@
 We begin this section with a brief review of additive models.
-Consider a regression scenario where the dependent variable $\mathbf{y} = \left(y_1, \dots, y_n\right)^{T}$ is thought to be associated with independent covariate vectors $\mathbf{x}_i = \left(x_{i1}, \dots, x_{ip}\right)$.
+Consider a regression scenario where the dependent variable $\mathbf{y} = \left(y_1, \dots, y_n\right)^{T}$ is thought to be associated with independent covariate vectors $\mathbf{x}_i = \left(x_{i1}, \dots, x_{ip}\right)$, for $i = 1, \dots, n$, with $n$ representing the number of observations.
 In this situation, the standard multiple linear regression model assumes that the dependence of $y_i$ on $\mathbf{x}_i$ takes the linear form
@@
 where again $\epsilon_i$ has mean 0 and variance $\sigma^2$, however for each $j$, $f_j$ represents an arbitrary function commonly assumed to be smooth.
-While the functions in model \ref{eq:1} only take as parameters the value of one covariate, with each covariate being represented in the model once, this does not need to be the case.
-The structure of additive models seeks to balance benefits of interpretability found in additive models with the flexibility of nonlinear, and frequently, nonparametric functional forms.
+While the functions in model \ref{eq:1} only take as parameters the value of one covariate, with each covariate being represented in the model once, this does not need to be the case, as functions specifying interactions between covariates may also be included.
+The structure of additive models seeks to balance benefits of interpretability found in linear regression models with the flexibility of nonlinear, and frequently, nonparametric functional forms.
 Further information can be found in \cite{hastieGeneralizedAdditiveModels1990}.
@@
 
-Individual participants are typically scheduled for imaging at six month intervals, though variability exists in terms of the recorded intervals.
+Individual participants are typically scheduled for imaging at six month intervals, though variability exists in the recorded intervals.
 Also of note, the numbers of observations contributed by each participant varies, so groups may be unbalanced.
-We denote the number of groups under consideration using $G$, the number of individuals included in the study as $I$, the number of images available as $N$, and the number of observations available for the $j$th image, with $j = 1, 2, \dots, N$, as $M_j$.
+We denote the number of groups under consideration using $G$, the number of individuals included in the study as $I$, the number of images available as $N$, and the number of observations available for the $j$th image as $Q_j$, with $j = 1, 2, \dots, N$.
 The individual to which the $j$th image belongs is denoted as $i(j)$, the group of the $i(j)$th individual is represented by $g(j)$, and we use $t_j$ to denote the time at which the $j$th image was recorded, with all times being relative to the $i(j)$th individual's baseline visit.
+This time benchmark may be changed to age, time from diagnosis, time from symptom onset, or any other measure of time, depending on the setting.
 Finally, we use $N_t$, $N_i$, and $N_g$ to denote the number of images available for time point $t$, individual $i$, and group $g$, respectively.
@@
 More generally, we let $d$ denote the intrinsic dimension of the low-dimensional manifold, and $D$ represent the dimension of the high-dimensional input space.
-For the $j$th image with $j = 1, \dots, N$, we denote the $m$th voxel in the $D$-dimensional space as $\mathbf{x}_{jm} \in \mathbb{R}^{D}$, its associated $d$-dimensional parameterization as $\mathbf{y}_{jm} \in \mathbb{R}^{d}$, and its weight as $w_{jm}$.
+For the $j$th image with $j = 1, \dots, N$, we denote the $q$th voxel in the $D$-dimensional space as $\mathbf{x}_{jq} \in \mathbb{R}^{D}$, its associated $d$-dimensional parameterization as $\mathbf{y}_{jq} \in \mathbb{R}^{d}$, and its weight as $w_{jq}$.
 We use $\mathbf{x}_g$, $\mathbf{x}_i$, $\mathbf{x}_{j}$, and $\mathbf{x}_t$ to denote the input voxels from all images in the $g$th group, $i$th individual, $j$th image, and $t$th time point, respectively, and use similar notation for all other relevant terms.
 
-The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{jm})$, which can be modeled as 
+The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{jq})$, which can be modeled as 
 \begin{equation}
-  \mathbf{x}_{jm} = \psi(\mathbf{y}_{jm}, t_j) = f_p(\mathbf{y}_{jm}, t_{j}) + f_{g(j)}(\mathbf{y}_{jm}, t_{j}) + f_{i(j)}(\mathbf{y}_{jm}, t_{j}) + f_j(\mathbf{y}_{jm}) + \epsilon_{jm}. \label{eq:2}
+  \mathbf{x}_{jq} = \psi(\mathbf{y}_{jq}, t_j) = f_p(\mathbf{y}_{jq}, t_{j}) + f_{g(j)}(\mathbf{y}_{jq}, t_{j}) + f_{i(j)}(\mathbf{y}_{jq}, t_{j}) + f_j(\mathbf{y}_{jq}) + \epsilon_{jq}. \label{eq:2}
 \end{equation}
@@
 \[%
-  \mathcal{K}_{\lambda}(f) = \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - f(\mathbf{y}_{jm})\|^2 + \lambda\|\nabla^{\otimes 2}f\|^2
+  \mathcal{K}_{\lambda}(f) = \sum_{j=1}^{N}\sum_{q=1}^{Q_j}w_{jq}\|\mathbf{x}_{jq} - f(\mathbf{y}_{jq})\|^2_{\mathbb{R}^{d}} + \lambda\|\nabla^{\otimes 2}f\|^2_{L^2(\mathbb{R}^{d})}
 ,\]%
-where $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is given. Thus, $f(\mathbf{y})$ takes the form
+where $\nabla^{\otimes 2}f$ denotes the Hessian matrix of $f$, and $\lambda$ is a given parameter used to penalize curvature in $f$. Thus, $f(\mathbf{y})$ takes the form
 \begin{equation}
-  f_l(\mathbf{y}) = \sum_{j=1}^{N}\sum_{m=1}^{M_j}s_{jm, l} \times \eta_{d}\left(\|\mathbf{y} - \mathbf{y}_{jm}\|\right) + \sum_{p = 1}^{d + 1}\alpha_{p, l} \times \phi_{p}(\mathbf{y}), \quad l = 1, 2, \dots, D, \label{eq:3}
+  f_l(\mathbf{y}) = \sum_{j=1}^{N}\sum_{q=1}^{Q_j}s_{jq, l} \times \eta_{d}\left(\|\mathbf{y} - \mathbf{y}_{jq}\|\right) + \sum_{p = 1}^{d + 1}\alpha_{p, l} \times \phi_{p}(\mathbf{y}), \quad l = 1, 2, \dots, D, \label{eq:3}
 \end{equation}
@@
 
-Without accounting for the additive structure of the model, and after concatenating all relevant values into $M = \sum_{j=1}^{N}M_{j}$-dimensional vectors of input values $\mathbf{y}$, weights $\mathbf{w}$, and outputs $\mathbf{x}$, we can estimate $f$ by obtaining the values of $s_l \in \mathbb{R}^{M}$ and $\alpha_l \in \mathbb{R}^{d + 1}$ for $l = 1, \dots, D$ that minimize $\|\mathbf{W}^{\frac{1}{2}}\left(\mathbf{x}_{l} - \mathbf{E}s_l - \mathbf{R}\alpha_l\right)\|^2_{\mathbb{R}^{M}} + \lambda\|\mathbf{E}^{\frac{1}{2}}s_l\|^2_{\mathbb{R}^{M}}$ under the constraint that $\mathbf{R}^{T}s_l = 0$, where
+Without accounting for the additive structure of the model, and after concatenating all relevant values into $Q = \sum_{j=1}^{N}Q_{j}$-dimensional vectors of input values $\mathbf{y}$, weights $\mathbf{w}$, and outputs $\mathbf{x}$, we can estimate $f$ by obtaining the values of $s_l \in \mathbb{R}^{Q}$ and $\alpha_l \in \mathbb{R}^{d + 1}$ for $l = 1, \dots, D$ that minimize $\|\mathbf{W}^{\frac{1}{2}}\left(\mathbf{x}_{l} - \mathbf{E}s_l - \mathbf{R}\alpha_l\right)\|^2_{\mathbb{R}^{Q}} + \lambda\|\mathbf{E}^{\frac{1}{2}}s_l\|^2_{\mathbb{R}^{Q}}$ under the constraint that $\mathbf{R}^{T}s_l = 0$, where
 \begin{itemize}
-  \item $\mathbf{E} = (E_{qr})_{1 \leq q, r, \leq M}$ is defined by $E_{qr} = \eta_{d}\left(\|\mathbf{y}_{q} - \mathbf{y}_{r}\|\right))$.
-  \item $\mathbf{R} = \left(R_{mp}\right)_{1 \leq m \leq M, \ 1 \leq p \leq d + 1}$ is defined by $R_{mp} = \phi_p(\mathbf{y}_{m})$.
-  \item $\mathbf{W} = diag(w_{1}, \dots, w_{M})$.
+  \item $\mathbf{E} = (E_{qr})_{1 \leq q, r, \leq Q}$ is defined by $E_{qr} = \eta_{d}\left(\|\mathbf{y}_{q} - \mathbf{y}_{r}\|\right))$.
+  \item $\mathbf{R} = \left(R_{mp}\right)_{1 \leq m \leq Q, \ 1 \leq p \leq d + 1}$ is defined by $R_{mp} = \phi_p(\mathbf{y}_{m})$.
+  \item $\mathbf{W} = diag(w_{1}, \dots, w_{Q})$.
 \end{itemize}
@@
 \begin{equation}
-  MSE(\lambda) = \frac{1}{K}\sum_{k=1}^{K}\frac{1}{N_u}\sum_{j=1}^{N_u}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\lambda}^{(k)}(\mathbf{y}_{jm})\|^2. \label{eq:5}
+  MSD(\lambda) = \frac{1}{K}\sum_{k=1}^{K}\frac{1}{N_u}\sum_{j=1}^{N_u}\frac{1}{Q_j}\sum_{q=1}^{Q_j}\|\mathbf{x}_{jq} - f_{\lambda}^{(k)}(\mathbf{y}_{jq})\|^2. \label{eq:5}
 \end{equation}
@@
 To estimate $f_p(\mathbf{y}, t)$, we use $\left\{\mathbf{x}_t\right\}_{t=1}^T$ and $\left\{\mathbf{y}_t\right\}_{t=1}^T$, yielding fitted function $\left\{\hat{f}_t(\mathbf{y})\right\}_{t=1}^T$.
-Using the maximum and minimum values of $\mathbf{y}_t$ in each dimension, a new $d$-dimensional grid of $M' = \text{max}\left(\left\{M_{j}\right\}_{j = 1}^{N}\right)$ knots, $\left\{\mathbf{y'}_m\right\}_{m=1}^{M'}$, is generated with the knots being evenly spread throughout the parameter space.
+Using the maximum and minimum values of $\mathbf{y}_t$ in each dimension, a new $d$-dimensional grid of $Q' = \text{max}\left(\left\{Q_{j}\right\}_{j = 1}^{N}\right)$ knots, $\left\{\mathbf{y'}_q\right\}_{q=1}^{Q'}$, is generated with the knots being evenly spread throughout the parameter space.
 These parameters serve as common knots that allow the spline coefficients to be compared across time points.
@@
 \begin{itemize}
-  \item $\mathbf{E}' = \left(E'_{qr}\right)_{1 \leq q, r, \leq M'}$ defined by $E'_{qr} = \eta_{d}\left(\|\mathbf{y}'_{q} - \mathbf{y}'_{r}\|\right)$.
-  \item $\mathbf{R}' = \left(R'_{mp}\right)_{1 \leq m \leq M', \ 1 \leq p \leq d + 1}$ is defined by $R'_{mp} = \phi_p(\mathbf{y'}_{m})$.
+  \item $\mathbf{E}' = \left(E'_{qr}\right)_{1 \leq q, r, \leq Q'}$ defined by $E'_{qr} = \eta_{d}\left(\|\mathbf{y}'_{q} - \mathbf{y}'_{r}\|\right)$.
+  \item $\mathbf{R}' = \left(R'_{mp}\right)_{1 \leq m \leq Q', \ 1 \leq p \leq d + 1}$ is defined by $R'_{mp} = \phi_p(\mathbf{y'}_{m})$.
 \end{itemize}
@@
 \begin{equation}
-  MSE(\gamma) = \frac{1}{T}\sum_{t=1}^{T}\frac{1}{N_t}\sum_{j=1}^{N_t}\frac{1}{M_j}\sum_{m=1}^{M_j}\|\mathbf{x}_{jm} - f_{\gamma}^{(t)}(\mathbf{y}_{jm}, t_j)\|^2. \label{eq:8}
+  MSD(\gamma) = \frac{1}{T}\sum_{t=1}^{T}\frac{1}{N_t}\sum_{j=1}^{N_t}\frac{1}{Q_j}\sum_{q=1}^{Q_j}\|\mathbf{x}_{jq} - f_{\gamma}^{(t)}(\mathbf{y}_{jq}, t_j)\|^2. \label{eq:8}
 \end{equation}
-We then let $\hat{\gamma}$ be the value of $\gamma$ that minimizes $MSE(\gamma)$.
+We then let $\hat{\gamma}$ be the value of $\gamma$ that minimizes $MSD(\gamma)$.
 
@@
 \begin{equation}
-  f(\mathbf{y}, t) = \sum_{m=1}^{M'}\mathbf{s}_{\hat{\gamma}, l}(t) \eta_{d}\left(\|\mathbf{y} - \mathbf{y'}_m\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\hat{\gamma}, p, l}(t) \phi_p(\mathbf{y}), \quad l = 1, \dots, D.
+  f(\mathbf{y}, t) = \sum_{q=1}^{Q'}\mathbf{s}_{\hat{\gamma}, l}(t) \eta_{d}\left(\|\mathbf{y} - \mathbf{y'}_m\|\right) + \sum_{p=1}^{d + 1}\mathbf{\alpha}_{\hat{\gamma}, p, l}(t) \phi_p(\mathbf{y}), \quad l = 1, \dots, D.
 \label{eq:9}
@@
 \caption{Varying-Coefficient Smoothing Spline Model}\label{alg:varying_spline}
-  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j = 1, m = 1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, weights $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and number of cross-validation folds $K$.}
+  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jq}\right\}_{j = 1, q = 1}^{N, Q_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jq}\right\}_{j=1, q=1}^{N, Q_{j}}$, weights $\left\{w_{jq}\right\}_{j=1, q=1}^{N, Q_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and number of cross-validation folds $K$.}
   \KwResult{Analytic formula of $\hat{f}(\mathbf{y}, t)$, coefficients $\Omega = \left(\mathbf{\delta}^{T}, \mathbf{\tau}^{T}\right)$, and optimal tuning parameter $\hat{\gamma}$.}
@@
       Solve equation \ref{eq:4} to estimate $f^{(-u)}_{\lambda_l, t}(\mathbf{y})$ with inputs $\mathbf{x}^{(-u)}_t$, $\mathbf{y}^{(-u)}_t$, $w^{(-u)}_t$\;
-      Estimate $MSE(\lambda_{l})_u$ using equation \ref{eq:5}\;
+      Estimate $MSD(\lambda_{l})_u$ using equation \ref{eq:5}\;
     }
-    Set $MSE(\lambda_l) = \frac{1}{U}\sum_{u=1}^{U}MSE(\lambda_l)_u$\;
+    Set $MSD(\lambda_l) = \frac{1}{U}\sum_{u=1}^{U}MSD(\lambda_l)_u$\;
   }
-  $\hat{\lambda}_t = \arg \min_{\lambda} MSE(\lambda)$\;
+  $\hat{\lambda}_t = \arg \min_{\lambda} MSD(\lambda)$\;
   $f_t(\mathbf{y}) = f_{\hat{\lambda}_t}(\mathbf{y})$\;
@@
 
-  Let $M' = \max(M_{j})$, $\left\{\mathbf{y}'_{m}\right\}_{m=1}^{M'}$ be a grid spanning the range of values of $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
+  Let $Q' = \max(Q_{j})$, $\left\{\mathbf{y}'_{q}\right\}_{q=1}^{Q'}$ be a grid spanning the range of values of $\left\{\mathbf{y}_{jq}\right\}_{j=1, q=1}^{N, Q_j}$\;
   \For{$t = 1, 2, \dots, T$} {
-    Compute $\left\{\mathbf{x'}_{m, t}\right\}_{m=1}^{M'} = \left\{f_t(\mathbf{y'}_m)\right\}_{m=1}^{M'}$\;
+    Compute $\left\{\mathbf{x'}_{q, t}\right\}_{q=1}^{Q'} = \left\{f_t(\mathbf{y'}_q)\right\}_{q=1}^{Q'}$\;
     Solve equation \ref{eq:6} to compute $\mathbf{s'}_t$ and $\mathbf{\alpha'}_t$\;
@@
     }
-    Estimate $MSE(\gamma_p)$ using \ref{eq:8}\;
+    Estimate $MSD(\gamma_p)$ using \ref{eq:8}\;
   }
-  $\hat{\gamma} = \arg \min_{\gamma} MSE(\gamma)$\;
+  $\hat{\gamma} = \arg \min_{\gamma} MSD(\gamma)$\;
   $\hat{f}(\mathbf{y}, t) = f_{\hat{\gamma}}(\mathbf{y}, t)$, where the form of $\hat{f}(\mathbf{y}, t)$ is given in \ref{eq:9}.
@@
 \[%
-  f_p(\mathbf{y}_{jm}, t_{j}) = E\left[\mathbf{x}_{jm} - f_{g(j)}(\mathbf{y}_{jm}, t_{j}) - f_{i(j)}(\mathbf{y}_{jm}, t_{j}) - f_{j}(\mathbf{y}_{jm})\right]
+  f_p(\mathbf{y}_{jq}, t_{j}) = E\left[\mathbf{x}_{jq} - f_{g(j)}(\mathbf{y}_{jq}, t_{j}) - f_{i(j)}(\mathbf{y}_{jq}, t_{j}) - f_{j}(\mathbf{y}_{jq})\right]
 ,\]%
-with a similar result holding for each of the smooth functions making up the additive model. Thus, the backfitting algorithm to fit the proposed model is given in Algorithm \ref{alg:backfitting}.
+with a similar result holding for each of the smooth functions making up the additive model.
+The model in its entirety is considered to have converged when the maximum of the distances between the coefficients of functions estimated for the same element in consecutive steps is below a given threshold.
+Details of the backfitting algorithm to fit the proposed model is given in Algorithm \ref{alg:backfitting}.
 
@@
 \caption{Backfitting Algorithm for Additive Spline Model}\label{alg:backfitting}
-  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, weights $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$ and $\left\{\gamma_p\right\}_{p=1}^{P}$, $\epsilon$ and $K$.}
+  \KwData{$D$-dimensional output values $\left\{\mathbf{x}_{jq}\right\}_{j=1, q=1}^{N, Q_{j}}$, $d$-dimensional input values $\left\{\mathbf{y}_{jq}\right\}_{j=1, q=1}^{N, Q_{j}}$, observation times $\left\{t_{j}\right\}_{j=1}^{N}$, weights $\left\{w_{jq}\right\}_{j=1, q=1}^{N, Q_j}$, group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$, candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}$ and $\left\{\gamma_p\right\}_{p=1}^{P}$, $\epsilon$ and $K$.}
   \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$.}
 
-  Estimate $f_{p}^{(0)}(\mathbf{y}, t)$ and $\Omega_{p}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_{j}\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+  Estimate $f_{p}^{(0)}(\mathbf{y}, t)$ and $\Omega_{p}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jq}\right\}_{j=1, q=1}^{N, Q_j}$, $\left\{\mathbf{y}_{jq}\right\}_{j=1, q=1}^{N, Q_j}$, $\left\{w_{jq}\right\}_{j=1, q=1}^{N, Q_j}$, $\left\{t_{j}\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
   \For{$g = 1, 2, \dots, G$} {
-    Estimate $f_{g}^{(0)}(\mathbf{y}, t)$ and $\Omega_{g}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{t_j\right\}_{j=1}^{N_g}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $K$\;
+    Estimate $f_{g}^{(0)}(\mathbf{y}, t)$ and $\Omega_{g}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jq} - f_{p}^{(0)}(\mathbf{y}_{jq}, t_j)\right\}_{j=1, q=1}^{N_g, Q_j}$, $\left\{\mathbf{y}_{jq}\right\}_{j=1, q=1}^{N_g, Q_j}$, $\left\{w_{jq}\right\}_{j=1, q=1}^{N_g, Q_j}$, $\left\{t_j\right\}_{j=1}^{N_g}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $K$\;
   }
   \For{$i = 1, 2, \dots, I$} {
-    Estimate $f_{i}^{(0)}(\mathbf{y}, t)$ and $\Omega_{i}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{t_j\right\}_{j=1}^{N_i}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+    Estimate $f_{i}^{(0)}(\mathbf{y}, t)$ and $\Omega_{i}^{(0)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jq} - f_{p}^{(0)}(\mathbf{y}_{jq}, t_j) - f_{g(j)}^{(0)}(\mathbf{y}_{jq}, t_j)\right\}_{j=1, q=1}^{N_i, Q_j}$, $\left\{\mathbf{y}_{jq}\right\}_{j=1, q=1}^{N_i, Q_j}$, $\left\{w_{jq}\right\}_{j=1, q=1}^{N_i, Q_j}$, $\left\{t_j\right\}_{j=1}^{N_i}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
   }
   \For{$j = 1, 2, \dots, N$} {
-    Estimate $f_{j}^{(0)}(\mathbf{y})$ and $\omega_{j}^{(0)}$ by solving \ref{eq:4} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(0)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(0)}(\mathbf{y}_{jm}, t_j)\right\}_{m=1}^{M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{m=1}^{M_j}$, $\left\{w_{jm}\right\}_{m = 1}^{M_j}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $K$\;
+    Estimate $f_{j}^{(0)}(\mathbf{y})$ and $\omega_{j}^{(0)}$ by solving \ref{eq:4} with inputs $\left\{\mathbf{x}_{jq} - f_{p}^{(0)}(\mathbf{y}_{jq}, t_j) - f_{g(j)}^{(0)}(\mathbf{y}_{jq}, t_j) - f_{i(j)}^{(0)}(\mathbf{y}_{jq}, t_j)\right\}_{q=1}^{Q_j}$, $\left\{\mathbf{y}_{jq}\right\}_{q=1}^{Q_j}$, $\left\{w_{jq}\right\}_{q = 1}^{Q_j}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $K$\;
   }
@@
   \While{$\max\left(\|\Omega_p^{(n)} - \Omega_{p}^{(n-1)}\|, \left\{\|\Omega_{g}^{(n)} - \Omega_{g}^{(n-1)}\|\right\}_{g=1}^{G}, \left\{\|\Omega_{i}^{(n)} - \Omega_{i}^{(n-1)}\|\right\}_{i=1}^{I}, \left\{\|\omega_{j}^{(n)} - \omega_{j}^{(n-1)}\|\right\}_{j=1}^{N}\right) \geq \epsilon$} {
-    Estimate $f_{p}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{p}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{g(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_j^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_{j}\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+    Estimate $f_{p}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{p}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jq} - f_{g(j)}^{(n)}(\mathbf{y}_{jq}, t_j) - f_{i(j)}^{(n)}(\mathbf{y}_{jq}, t_j) - f_j^{(n)}(\mathbf{y}_{jq})\right\}_{j=1, q=1}^{N, Q_j}$, $\left\{\mathbf{y}_{jq}\right\}_{j=1, q=1}^{N, Q_j}$, $\left\{w_{jq}\right\}_{j=1, q=1}^{N, Q_j}$, $\left\{t_{j}\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
     \For{$g = 1, 2, \dots, G$} {
-      Estimate $f_{g}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{g}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n)}(\mathbf{y}_{jm}, t_j) - f_{j}^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_g, M_j}$, $\left\{t_j\right\}_{j=1}^{N_g}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $K$\;
+      Estimate $f_{g}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{g}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jq} - f_{p}^{(n+1)}(\mathbf{y}_{jq}, t_j) - f_{i(j)}^{(n)}(\mathbf{y}_{jq}, t_j) - f_{j}^{(n)}(\mathbf{y}_{jq})\right\}_{j=1, q=1}^{N_g, Q_j}$, $\left\{\mathbf{y}_{jq}\right\}_{j=1, q=1}^{N_g, Q_j}$, $\left\{w_{jq}\right\}_{j=1, q=1}^{N_g, Q_j}$, $\left\{t_j\right\}_{j=1}^{N_g}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $K$\;
     }
     \For{$i = 1, 2, \dots, I$} {
-      Estimate $f_{i}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{i}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_j^{(n)}(\mathbf{y}_{jm})\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N_i, M_j}$, $\left\{t_j\right\}_{j=1}^{N_i}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
+      Estimate $f_{i}^{(n+1)}(\mathbf{y}, t)$ and $\Omega_{i}^{(n+1)}$ using Algorithm \ref{alg:varying_spline} with inputs $\left\{\mathbf{x}_{jq} - f_{p}^{(n+1)}(\mathbf{y}_{jq}, t_j) - f_{g(j)}^{(n+1)}(\mathbf{y}_{jq}, t_j) - f_j^{(n)}(\mathbf{y}_{jq})\right\}_{j=1, q=1}^{N_i, Q_j}$, $\left\{\mathbf{y}_{jq}\right\}_{j=1, q=1}^{N_i, Q_j}$, $\left\{w_{jq}\right\}_{j=1, q=1}^{N_i, Q_j}$, $\left\{t_j\right\}_{j=1}^{N_i}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, and $K$\;
     }
     \For{$j = 1, 2, \dots, N$} {
-      Estimate $f_{j}^{(n+1)}(\mathbf{y})$ and $\omega_{j}^{(n+1)}$ by solving \ref{eq:4} with inputs $\left\{\mathbf{x}_{jm} - f_{p}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{g(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j) - f_{i(j)}^{(n+1)}(\mathbf{y}_{jm}, t_j)\right\}_{m=1}^{M_j}$, $\left\{\mathbf{y}_{jm}\right\}_{m=1}^{M_j}$, $\left\{w_{jm}\right\}_{m = 1}^{M_j}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $K$\;
+      Estimate $f_{j}^{(n+1)}(\mathbf{y})$ and $\omega_{j}^{(n+1)}$ by solving \ref{eq:4} with inputs $\left\{\mathbf{x}_{jq} - f_{p}^{(n+1)}(\mathbf{y}_{jq}, t_j) - f_{g(j)}^{(n+1)}(\mathbf{y}_{jq}, t_j) - f_{i(j)}^{(n+1)}(\mathbf{y}_{jq}, t_j)\right\}_{q=1}^{Q_j}$, $\left\{\mathbf{y}_{jq}\right\}_{q=1}^{Q_j}$, $\left\{w_{jq}\right\}_{m = 1}^{Q_j}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $K$\;
     }
@@
 
-As mentioned previously, this proposed additive spline model structure has several benefits.
+This proposed additive spline model structure has several benefits.
 First, it maintains interpretability at each level of the model.
@@
 The PME algorithm is largely split into three distinct steps: data reduction, fitting, and tuning.
-The split between these three steps enforces a process by which candidate functions are fit for each in a vector of smoothing values in the fitting stage, then the tuning step is used to select the smoothing value that results in the lowest error, as measured by mean squared distance,
+The split between these three steps enforces a process by which candidate functions are fit for each element in a vector of candidate smoothing values in the fitting stage, then the tuning step is used to select the smoothing value that results in the lowest error, as measured by mean squared distance,
 \[%
@@
 
-While this approach to tuning the model parameters is beneficial because it allows a complete comparison of all parameter values under consideration, the additive model described in equation \ref{eq:2} requires many more parameters, meaning that taking the same tuning approach would result in the need to find an optimal point in a very high-dimensional space.
+While this approach to tuning the model parameters is beneficial because it allows a complete comparison of all parameter values under consideration, the additive model described in equation \ref{eq:2} requires many more parameters, meaning that taking the same tuning approach would call for finding an optimal point in a very high-dimensional space.
 This would greatly increase the computational complexity of the algorithm, likely to the point of infeasibility in an practical situation.
-Instead, the proposed algorithm avoids this problem by taking a greedy approach to tuning the model, where the smoothing parameters for each model are chosen individually, rather than simultaneously.
+Instead, the proposed algorithm avoids this problem by taking a greedy approach to tuning the model, where the smoothing parameters for each model are chosen individually during the spline model's tuning step, rather than simultaneously.
 
-Thus, the hierarchical PME algorithm proposed here is composed of initialization and data reduction and fitting steps.
+Thus, the hierarchical PME algorithm proposed here is composed of initialization, data reduction, and fitting steps.
 The first step uses the PME algorithm to estimate an embedding map from one image.
-The HDMDE algorithm, detailed in \cite{mengPrincipalManifoldEstimation2021}, is used to reduce the data from each image into a set of component centers and associated weights.
-Then, the embedding map estimated using the PME algorithm can be used to provide initial parameterizations of the component centers for every image.
-These initial parameterizations are then used in the fitting step.
+The HDMDE algorithm, detailed in \cite{mengPrincipalManifoldEstimation2021}, is then used to reduce the data from each image into a set of component centers and associated weights.
+The embedding map estimated using the PME algorithm can be used to provide initial parameterizations of the component centers for every image.
+These parameterizations are then used as starting points in the fitting step.
 
@@
 \caption{Hierarchical Principal Manifold Estimation Algorithm}\label{alg:hpme}
-  \KwData{Data points $\left\{\mathbf{v}_{jm}\right\}_{j=1, m=1}^{N, M'_j}$ in $\mathbb{R}^{D}$; group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$; a positive integer $N_0 < \min\left(\left\{M'_j\right\}_{j=1}^{N}\right)$; $\alpha, \epsilon, \epsilon^{*} \in (0, 1)$; candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}, \left\{\gamma_p\right\}_{p=1}^{P}$; $itr \geq 1$, which is the maximum number of iterations allowed.}
+  \KwData{Data points $\left\{\mathbf{v}_{jq}\right\}_{j=1, q=1}^{N, Q'_j}$ in $\mathbb{R}^{D}$; group and individual indicator values $\left\{g(j)\right\}_{j=1}^{N}$ and $\left\{i(j)\right\}_{j=1}^{N}$; a positive integer $N_0 < \min\left(\left\{Q'_j\right\}_{j=1}^{N}\right)$; $\alpha, \epsilon, \epsilon^{*} \in (0, 1)$; candidate tuning parameters $\left\{\lambda_l\right\}_{l=1}^{L}, \left\{\gamma_p\right\}_{p=1}^{P}$; $itr \geq 1$, which is the maximum number of iterations allowed.}
   \KwResult{Analytic formulas of $\hat{f}_p$, $\left\{\hat{f}_g\right\}_{g=1}^{G}, \left\{\hat{f}_i\right\}_{i=1}^{I}$, and $\left\{\hat{f}_{j}\right\}_{j=1}^{N}$, .}
-  \textbf{Initialization}: Apply PME with input $\left(\left\{\mathbf{v}_{1m}\right\}_{m=1}^{M'_1}, N_0, \alpha, \epsilon, \epsilon^{*}, \left\{\lambda_l\right\}_{l=1}^{L}, itr\right)$ and obtain $f^{(1)}$\;
+  \textbf{Initialization}: Apply PME with input $\left(\left\{\mathbf{v}_{1m}\right\}_{q=1}^{Q'_1}, N_0, \alpha, \epsilon, \epsilon^{*}, \left\{\lambda_l\right\}_{l=1}^{L}, itr\right)$ and obtain $f^{(1)}$\;
   $\psi^{(1)} \gets f^{(1)}$\;
-  \textbf{Data Reduction}: Apply HDMDE algorithm with inputs $\left(\left\{\left\{\mathbf{v}_{jm}\right\}_{m=1}^{M_j}\right\}_{j=1}^{N}, N_0, \epsilon, \alpha\right)$ and obtain $\left\{M_j\right\}_{j=1}^{N}$, $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, and $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
-  \textbf{Parameterization}: Use the projection index of $\psi^{(1)}$, $\kappa_{\psi^{(1)}}$, to parameterize $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$ by $d$-dimensional parameters $\left\{y_{jm}\right\}_{j=1, m=1}^{N, M_j}$\;
-  $\mathcal{E} \gets 2 \times \epsilon^{*}$, $n \gets 1$, and $\mathcal{D}(\psi^{(1)}) \gets \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - \psi^{(1)}\left(\kappa_{\psi^{(1)}}(\mathbf{x}_{jm})\right)\|_{\mathbb{R}^{D}}^2$\;
+  \textbf{Data Reduction}: Apply HDMDE algorithm with inputs $\left(\left\{\left\{\mathbf{v}_{jq}\right\}_{q=1}^{Q_j}\right\}_{j=1}^{N}, N_0, \epsilon, \alpha\right)$ and obtain $\left\{Q_j\right\}_{j=1}^{N}$, $\left\{\mathbf{x}_{jq}\right\}_{j=1, q=1}^{N, Q_j}$, and $\left\{w_{jq}\right\}_{j=1, q=1}^{N, Q_j}$\;
+  \textbf{Parameterization}: Use the projection index of $\psi^{(1)}$, $\kappa_{\psi^{(1)}}$, to parameterize $\left\{\mathbf{x}_{jq}\right\}_{j=1, q=1}^{N, Q_j}$ by $d$-dimensional parameters $\left\{y_{jq}\right\}_{j=1, q=1}^{N, Q_j}$\;
+  $\mathcal{E} \gets 2 \times \epsilon^{*}$, $n \gets 1$, and $\mathcal{D}(\psi^{(1)}) \gets \sum_{j=1}^{N}\sum_{q=1}^{Q_j}w_{jq}\|\mathbf{x}_{jq} - \psi^{(1)}\left(\kappa_{\psi^{(1)}}(\mathbf{x}_{jq})\right)\|_{\mathbb{R}^{D}}^2$\;
   \While{$\mathcal{E} \geq \epsilon^{*}$ and $n < itr$} {
-    Compute $f_{p}^{(n + 1)}$, $\left\{f_{g}^{(n + 1)}\right\}_{g = 1}^{G}$, $\left\{f_{i}^{(n + 1)}\right\}_{i=1}^{I}$, and $\left\{f_{j}^{(n+1)}\right\}_{j=1}^{N}$ by using Algorithm \ref{alg:backfitting} with inputs $\left\{\mathbf{x}_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{\kappa_{\psi^{(n)}}(\mathbf{x}_{jm})\right\}_{j=1, m=1}^{N, M_j}$, $\left\{t_j\right\}_{j=1}^{N}$, $\left\{w_{jm}\right\}_{j=1, m=1}^{N, M_j}$, $\left\{g(j)\right\}_{j=1}^{N}$, $\left\{i(j)\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $\epsilon$, and $K$\;
+    Compute $f_{p}^{(n + 1)}$, $\left\{f_{g}^{(n + 1)}\right\}_{g = 1}^{G}$, $\left\{f_{i}^{(n + 1)}\right\}_{i=1}^{I}$, and $\left\{f_{j}^{(n+1)}\right\}_{j=1}^{N}$ by using Algorithm \ref{alg:backfitting} with inputs $\left\{\mathbf{x}_{jq}\right\}_{j=1, q=1}^{N, Q_j}$, $\left\{\kappa_{\psi^{(n)}}(\mathbf{x}_{jq})\right\}_{j=1, q=1}^{N, Q_j}$, $\left\{t_j\right\}_{j=1}^{N}$, $\left\{w_{jq}\right\}_{j=1, q=1}^{N, Q_j}$, $\left\{g(j)\right\}_{j=1}^{N}$, $\left\{i(j)\right\}_{j=1}^{N}$, $\left\{\lambda_l\right\}_{l=1}^{L}$, $\left\{\gamma_p\right\}_{p=1}^{P}$, $\epsilon$, and $K$\;
     $\left\{\psi^{(n+1)}(\mathbf{y}, t) \gets f_{p}^{(n + 1)}(\mathbf{y}, t) + f_{g(j)}^{(n+1)}(\mathbf{y}, t) + f_{i(j)}^{(n + 1)}(\mathbf{y}, t) + f_{j}^{(n+1)}(\mathbf{y})\right\}_{j=1}^{N}$\;
-    $\mathcal{D}(\psi^{n+1}) \gets \sum_{j=1}^{N}\sum_{m=1}^{M_j}w_{jm}\|\mathbf{x}_{jm} - \psi^{n + 1}\left(\kappa_{\psi^{n+1}}(\mathbf{x}_{jm})\right)\|_{\mathbb{R}^{D}}^2$\;
+    $\mathcal{D}(\psi^{n+1}) \gets \sum_{j=1}^{N}\sum_{q=1}^{Q_j}w_{jq}\|\mathbf{x}_{jq} - \psi^{n + 1}\left(\kappa_{\psi^{n+1}}(\mathbf{x}_{jq})\right)\|_{\mathbb{R}^{D}}^2$\;
     $\mathcal{E} \gets |\mathcal{D}(\psi^{(n+1)}) - \mathcal{D}(\psi^{(n)})| / \mathcal{D}(\psi^{(n)})$\;

### 2023-10-03 paper/main.tex
edits, added preliminary mnist results

@@
 The performance of the two approaches is then compared when they are applied to a series of simulated datasets of varying complexities.
-Section \ref{s:hpme_adni} discusses proposed work, which will demonstrate how the model specified in Section \ref{s:methods} may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset, while Section \ref{s:hpme_discussion} provides a concluding discussion of the proposed algorithm and opportunities for further study.
+Section \ref{s:hpme_adni} discusses proposed work, which will demonstrate how the model specified in Section \ref{s:methods} may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset.
+This will be followed by a concluding discussion of the proposed algorithm and opportunities for further study.
 
@@
 
+To demonstrate the potential benefits of the proposed hierarchical estimation approach when compared to the existing PME algorithm, we implemented the hierarchical PME algorithm using a simplified version of the additive spline model presented in Section \ref{s:methods} for use on images from the MNIST database.
+The MNIST database includes 70,000 $28 \times 28$ images of handwritten numerical digits written by different individuals, with 60,000 in a training set and an additional 10,000 in a test set.
+To test the hierarchical algorithm, we considered a reduced set of 100 images of the numeral ``5", where the individual pixels of each image with intensity greater than 0 can be considered to create a 2-dimensional data cloud, through which a 1-dimensional curve passes.
+Each representation of the numeral must follow the same general path to be recognizable as a ``5", but differences in individual handwriting styles result in variation between images.
+We have no meaningful way of grouping images by individual or subgroup in this case, and there are no temporal relationships between images, so we assume the embedding function used in Algorithm \ref{alg:hpme} takes the form \[%
+  \mathbf{x}_{jq} = \psi(\mathbf{y}_{jq}) = f_p(\mathbf{y}_{jq}) + f_j(\mathbf{y}_{jq}) + \epsilon_{jq}
+.\]%
+
+Plots depicting how the population-level manifold estimate differs from the manifolds estimated for each image are shown for four images in Figure \ref{fig:mnist_results}.
+Here, we see that the population estimate, represented in red, clearly captures the shape of a numeral ``5" and does not show any clear bias compared to the data points included in the plots.
+Meanwhile, the estimates for the individual images, shown in blue, are more responsive to the image data to which it is fit, indicating that the additive model fit successfully to the data, and providing a simple demonstration of the improvements made available by incorporating hierarchical information into the manifold learning procedure.
+
+\begin{figure}%
+  \centering
+  \subfloat[\centering MNIST Image 1]{{\includegraphics[width=9cm]{mnist_5plot1}}}%
+  \hfill
+  \subfloat[\centering MNIST Image 2]{{\includegraphics[width=9cm]{mnist_5plot2}}}
+  \vfill
+  \subfloat[\centering MNIST Image 3]{{\includegraphics[width=9cm]{mnist_5plot3}}}
+  \hfill
+  \subfloat[\centering MNIST Image 4]{{\includegraphics[width=9cm]{mnist_5plot4}}}
+  \caption{Comparison of PME, Population, and Image-Level Results for MNIST Data}
+  \label{fig:mnist_results}
+\end{figure}
+
 \section{Proposed Work} \label{s:hpme_adni}
 Further work will be dedicated to testing the proposed approach in several settings.
-First, the algorithm will be tested with a simplified additive model on the MNIST dataset of handwritten images.
-This will provide an opportunity to test the method on a relatively small imaging dataset, serving as a proof of concept and indicating where problems may arise when the algorithm is used with more complex data structures.
+First, more extensive performance comparisons between the PME algorithm and its hierarchical extension will be conducted when applied to the MNIST dataset.
 The method will next be applied to a series of simulated datasets that were generated to reflect the hierarchical structure present in the ADNI dataset.
-In both of these cases, the estimates of the hierarchical approach will be compared to alternative methods, with the fit of each method to the data being compared.
+Again, the estimates of the hierarchical approach will be compared to alternative methods, with the fit of each method to the data being compared.
 Finally, the hierarchical approach will be applied to structural MRI data from the ADNI dataset.
@@
 Of interest in this case will be a comparison of the varying-coefficient models by group, which may provide a visual confirmation that the algorithm's estimates are detecting the most relevant information included in the data.
-
-\section{Discussion} \label{s:hpme_discussion}
 

### 2025-03-18 manuscript/main.tex
rename paper directory

--- 2023-10-03 paper/main.tex
@@
 A method to accurately and efficiently estimate the manifolds underlying data at the group level must take advantage of the grouped structure of observations.
-To develop an estimate of the manifold underlying the hippocampi of participants in the Alzheimer's Disease Neuroimaging Initiative (ADNI) study, it is necessary to recognize the basic nesting structure of the data, in which brain images are nested within participants belong to a latent disease subgroup where patients tend to have more related symptom experiences or underlying physiological developments.
+To develop an estimate of the manifold underlying the hippocampi of participants in the Alzheimer's Disease Neuroimaging Initiative (ADNI) study, it is necessary to recognize the basic nesting structure of the data, in which brain images are possibly nested within participants belong to a latent disease subgroup where patients tend to have more related symptom experiences or underlying physiological developments.
 These subgroups are in turn nested within study groups based on Alzheimer's disease status, which are finally nested within the greater study population.

### 2025-03-18 code/adni_mri_preprocess.R
initial code commit

@@ -1,0 +1,529 @@
+library(oro.dicom)
+library(oro.nifti)
+library(neurobase)
+library(fslr)
+# library(extrantsr)
+library(stringr)
+library(foreach)
+library(tidyverse)
+library(lubridate)
+# library(ANTsRCore)
+
+sub_dirs <- list.dirs("data/adni", recursive = FALSE)
+
+nii_scans <- list()
+
+processed_scans <- list()
+
+hipp_mask <- readNIfTI(
+  paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm_Hipp_mask_dil8.nii.gz")
+)
+
+mni_nifti <- readNIfTI(
+  paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm.nii.gz")
+)
+
+ncores <- parallel::detectCores() - 1
+cl <- parallel::makeCluster(ncores, type = "FORK")
+doParallel::registerDoParallel(cl = cl)
+
+# foreach (dir_idx = 1:length(sub_dirs)) %dopar% {
+foreach(dir_idx = 1:8) %dopar% {
+  img_dirs <- list.files(sub_dirs[dir_idx], recursive = TRUE, full.names = TRUE) %>%
+    gsub(pattern = "([^/]+$)", replacement = "") %>%
+    unique()
+
+  nii_scans[[dir_idx]] <- list()
+
+  for (img_idx in 1:length(img_dirs)) {
+    proc_dir <- paste0("data/adni_processed_fsl", gsub(pattern = "data/adni", replacement = "", img_dirs[img_idx]))
+    if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
+      next
+    } else {
+      all_slices <- readDICOM(img_dirs[img_idx])
+      nii <- dicom2nifti(all_slices)
+
+      run_first_all(
+        nii,
+        oprefix = proc_dir,
+        verbose = FALSE,
+        opts = "-d"
+      )
+    }
+  }
+}
+
+parallel::stopCluster(cl = cl)
+
+test_dicom <- readDICOM(
+  list.files(sub_dirs[1], recursive = TRUE, full.names = TRUE)[1]
+)
+
+adni_info <- read_csv("data/adni_info.csv")
+adni_info <- adni_info  |> 
+  rename(
+    image_id = "Image Data ID",
+    subid = "Subject",
+    date = "Acq Date"
+  ) |>
+  mutate(
[…]

### 2025-03-18 code/functions/fit_mnist_additive_model.R

@@ -1,0 +1,209 @@
+fit_mnist_additive_model <- function(x, params, weights, lambda, k, ids, epsilon = 0.05, max_iter = 100) {
+  D <- ncol(x)
+  d <- ncol(params)
+
+  fold_vec_full <- rep(1:k, ceiling(length(unique(ids)) / k))
+  id_folds <- fold_vec_full[
+    sample(
+      1:length(unique(ids)),
+      length(unique(ids)),
+      replace = FALSE
+    )
+  ]
+
+  print("Initializing population model")
+  init_population_embedding <- fit_weighted_spline(
+    x,
+    params,
+    weights,
+    lambda,
+    id_folds[ids]
+  )
+
+  print("Initializing ID models")
+  init_id_embeddings <- list()
+  id_x <- list()
+  id_params <- list()
+  id_weights <- list()
+  id_coefs <- list()
+  mean_x <- list()
+  for (id_idx in 1:length(unique(ids))) {
+    id_x[[id_idx]] <- x[ids == unique(ids)[id_idx], ]
+    id_params[[id_idx]] <- params[ids == unique(ids)[id_idx], ] %>%
+      matrix(ncol = 1)
+    # id_weights[[id_idx]] <- weights[ids == unique(ids)[id_idx]]
+    id_weights[[id_idx]] <- rep(1 / nrow(id_x[[id_idx]]), nrow(id_x[[id_idx]]))
+    fold_vec <- rep(1:k, ceiling(nrow(id_x[[id_idx]]) / k))
+    id_folds <- fold_vec[
+      sample(
+        1:nrow(id_x[[id_idx]]),
+        size = nrow(id_x[[id_idx]]),
+        replace = FALSE
+      )
+    ]
+    id_preds <- map(
+      1:nrow(id_params[[id_idx]]),
+      ~ init_population_embedding$embedding_map(id_params[[id_idx]][.x, ])
+    ) %>%
+      reduce(rbind)
+    init_id_embeddings[[id_idx]] <- fit_weighted_spline(
+      id_x[[id_idx]] - id_preds,
+      id_params[[id_idx]],
+      id_weights[[id_idx]],
+      lambda,
+      id_folds
+    )
+    mean_x[[id_idx]] <- map(
+      1:nrow(id_params[[id_idx]]),
+      ~ init_id_embeddings[[id_idx]]$embedding_map(id_params[[id_idx]][.x, ])
+    ) %>%
[…]

### 2025-03-18 code/functions/fit_weighted_spline.R

@@ -1,0 +1,75 @@
+fit_weighted_spline <- function(x, params, weights, smoothing_vals, folds) {
+  k <- length(unique(folds))
+  coefs <- list()
+  d <- ncol(params)
+  smoothing_error <- vector()
+
+  for(smoothing_idx in 1:length(smoothing_vals)) {
+    coefs[[smoothing_idx]] <- list()
+    fold_errors <- vector()
+    for (k_idx in 1:k) {
+      train_params <- params[!(folds == k_idx), ] %>%
+        matrix(ncol = d)
+      fold_params <- params[folds == k_idx, ] %>%
+        matrix(ncol = d)
+      train_param_mat <- cbind(rep(1, nrow(train_params)), train_params)
+      train_x <- x[!(folds == k_idx), ]
+      fold_x <- x[folds == k_idx, ]
+      train_weights <- diag(weights[!(folds == k_idx)])
+      train_E <- calcE(train_params, 4 - ncol(train_params))
+      coefs[[smoothing_idx]][[k_idx]] <- solve_weighted_spline(
+        train_E,
+        train_weights,
+        train_param_mat,
+        train_x,
+        lambda[smoothing_idx],
+        ncol(train_params),
+        ncol(train_x)
+      )
+      fold_embedding <- function(parameters) {
+        as.vector(
+          (t(coefs[[smoothing_idx]][[k_idx]][1:nrow(train_x), ]) %*% etaFunc(parameters, train_params, 4 - d)) +
+            (t(coefs[[smoothing_idx]][[k_idx]][(nrow(train_x) + 1):(nrow(train_x) + d + 1), ]) %*% matrix(c(1, parameters), ncol = 1))
+        )
+      }
+      fold_errors[k_idx] <- map(
+        1:nrow(fold_x),
+        ~ weights[folds == k_idx][.x] * dist_euclidean(fold_x[.x, ], fold_embedding(fold_params[.x, ]))^2
+      ) %>%
+        reduce(c) %>%
+        mean()
+    }
+    smoothing_error[smoothing_idx] <- mean(fold_errors)
+    if (length(smoothing_error) >= 5) {
+      if (!is.unsorted(smoothing_error[(length(smoothing_error) - 4):length(smoothing_error)])) {
+        break
+      }
+    }
+  }
+  opt_run <- which.min(smoothing_error)
+
+  E <- calcE(params, 4 - ncol(params))
[…]

### 2025-03-18 code/functions/plot_id_pred.R

@@ -1,0 +1,32 @@
+plot_id_pred <- function(id, pop_embedding, id_embeddings, params, ids) {
+  # mnist_data <- mnist_x[mnist_id == id, ]
+  mnist_data <- input[id_vals == id, ]
+  id_params <- params[ids == id, ]
+
+  means <- list()
+  for (id_idx in 1:length(id_embeddings)) {
+    means[[id_idx]] <- map(
+      1:nrow(matrix(params[ids == id, ])),
+      ~ id_embeddings[[id_idx]]$embedding_map(matrix(params[ids == id, ])[.x, ])
+    ) %>%
+      reduce(rbind) %>%
+      colMeans()
+  }
+
+  sample_pop_params <- seq(min(params), max(params), length.out = 1000)
+  sample_id_params <- seq(min(id_params), max(id_params), length.out = 1000)
+  pop_out <- map(1:length(sample_pop_params), ~ pop_embedding$embedding_map(sample_pop_params[.x])) %>%
+    reduce(rbind)
+  id_out <- map(
+    1:length(sample_id_params),
+    ~ id_embeddings[[id]]$embedding_map(sample_id_params[.x]) - means[[id]]
+  ) %>%
+    reduce(rbind)
+  plt <- ggplot() +
+    geom_jitter(aes(x = mnist_data[, 1], y = mnist_data[, 2])) +
+    geom_point(aes(x = pop_out[, 1], y = pop_out[, 2]), color = "red") +
+    geom_point(aes(x = pop_out[, 1] + id_out[, 1], y = pop_out[, 2] + id_out[, 2]), color = "blue") +
+    xlab("X") +
+    ylab("Y")
+  print(plt)
+}

### 2025-03-18 code/functions/print_SSD.R

@@ -1,0 +1,13 @@
+print_SSD <- function(SSD_new, SSD_ratio, count) {
+  print(
+    paste0(
+      "For iteration # ",
+      as.character(count),
+      ", SSD = ",
+      as.character(round(SSD_new, 4)),
+      " and SSD_ratio = ",
+      as.character(round(SSD_ratio, 4)),
+      "."
+    )
+  )
+}

### 2025-03-18 code/mnist_pme_test.R

@@ -1,0 +1,126 @@
+library(doParallel)
+library(foreach)
+# library(pme)
+library(ggplot2)
+library(ggimage)
+library(imager)
+library(progress)
+library(Rfast)
+library(tidyverse)
+
+plot_img <- function(img_vec) {
+  img_mat <- matrix(img_vec, nrow = 28)[, 28:1]
+  image(img_mat)
+}
+
+mnist <- readRDS("~/Documents/brown/research/population-pme/data/mnist.RDS")
+mnist_train5 <- mnist$train$images[mnist$train$labels == 5, ]
+mnist_train5_ids <- 1:nrow(mnist_train5)
+
+mnist_train <- mnist$train$images
+mnist_train_ids <- 1:nrow(mnist_train)
+mnist_train_labels <- mnist$train$labels
+
+set.seed(58173)
+mnist_train_idx <- sample(1:nrow(mnist_train), 2500)
+mnist_train <- mnist_train[mnist_train_idx, ]
+mnist_train_labels <- mnist_train_labels[mnist_train_idx]
+
+mnist_pme <- pme(
+  mnist_train,
+  d = 3,
+  initialization_type = "subsample",
+  lambda = exp(-20:5),
+  epsilon = 0.01,
+  alpha = 0.01,
+  max_iter = 1000,
+  verbose = TRUE
+)
+
+params <- mnist_pme$parameterization[[which.min(mnist_pme$MSD)]]
+est_centers <- purrr::map(
+  1:nrow(params),
+  ~ mnist_pme$embedding_map(params[.x, ])
+) %>%
+  reduce(rbind)
+
+nearest_centers <- calc_nearest_x(mnist_train, est_centers)
+
+proj_params <- purrr::map(
+  1:nrow(mnist_train),
+  ~ projection_pme(mnist_train[.x, ], f = mnist_pme$embedding_map, initial_guess = params[nearest_centers[.x] + 1, ])
+) %>%
+  reduce(rbind)
+
+
+mnist_images <- list()
+for (idx in 1:nrow(mnist_train)) {
+  # mnist_mat <- matrix(mnist_train5[idx, ], nrow = 28)[, 28:1]
+  # mnist_mat <- matrix(mnist_train5[idx, ], nrow = 28)[, 28:1]
+  mnist_mat <- matrix(mnist_train[idx, ], nrow = 28)
+  mnist_images[[idx]] <- as.cimg(mnist_mat)
+}
+
+for (img_val in 1:length(mnist_images)) {
[…]

### 2025-03-18 code/mnist_test.R

@@ -1,0 +1,278 @@
+library(doParallel)
+library(foreach)
+# library(pme)
+library(progress)
+library(Rfast)
+library(tidyverse)
+
+source("functions/fit_weighted_spline.R")
+source("functions/print_SSD.R")
+source("functions/fit_mnist_additive_model.R")
+source("functions/plot_id_pred.R")
+
+SSD_ratio_threshold <- 5
+verbose <- TRUE
+
+
+
+mnist <- readRDS("~/Documents/brown/research/population-pme/data/mnist.RDS")
+mnist_train5 <- mnist$train$images[mnist$train$labels == 5, ]
+mnist_train5_ids <- 1:nrow(mnist_train5)
+
+mnist_train5_proc <- list()
+for (idx in mnist_train5_ids[1:100]) {
+  mnist_mat <- matrix(mnist_train5[idx, ], nrow = 28)[, 28:1]
+  digit_indices <- which(mnist_mat != 0, arr.ind = TRUE)
+  digit_indices <- map(
+    1:nrow(digit_indices),
+    ~ matrix(
+      rep(
+        digit_indices[.x, ],
+        mnist_mat[digit_indices[.x, 1], digit_indices[.x, 2]]
+      ),
+      ncol = 2,
+      byrow = TRUE
+    )
+  ) %>%
+    reduce(rbind)
+  digit_indices_range <- colMinsMaxs(digit_indices)
+  digit_indices[, 1] <- (digit_indices[, 1] - digit_indices_range[1, 1]) /
+    (digit_indices_range[2, 1] - digit_indices_range[1, 1])
+  digit_indices[, 2] <- (digit_indices[, 2] - digit_indices_range[1, 2]) /
+    (digit_indices_range[2, 2] - digit_indices_range[1, 2])
+  digit_indices <- cbind(idx, digit_indices)
+  mnist_train5_proc[[idx]] <- digit_indices
+}
+mnist_train5 <- reduce(mnist_train5_proc, rbind)
+
+mnist_train3 <- mnist$train$images[mnist$train$labels == 3, ]
+mnist_train3_ids <- 1:nrow(mnist_train3)
+
+mnist_train3_proc <- list()
+for (idx in mnist_train3_ids) {
+  mnist_mat <- matrix(mnist_train3[idx, ], nrow = 28)[, 28:1]
+  digit_indices <- which(mnist_mat != 0, arr.ind = TRUE)
[…]

### 2025-03-18 code/adni_mri_preprocess.R
fix typo

@@ -349,12 +349,12 @@ for (dir_idx in 1:length(processed_dirs)) {
         }
       }
       rhipp_temp <- cbind(
-        scan_id
+        scan_id,
         rhipp_idx %*% diag(pixdim(as.nifti(temp_rhipp))[2:4]),
         temp_rhipp[rhipp_idx]
       )
       rhipp_surface_temp <- cbind(
-        scan_id
+        scan_id,
         rhipp_idx[surface, ] %*% diag(pixdim(as.nifti(temp_rhipp))[2:4]),
         temp_rhipp[rhipp_idx[surface, ]]
       )

### 2025-03-18 code/adni_mri_preprocess.R
remove unneeded vectors

@@ -236,8 +236,6 @@ for (dir_idx in 1:length(processed_dirs)) {
     scan_info <- scan_info[scan_id, ]
 
     if (file.exists(paste0(scan_dirs[scan_idx], "_all_fast_origsegs.nii.gz"))) {
-      img_id <- c(img_id, scan_id)
-      img_date <- c(img_date, floor_date(scan_dates[scan_idx], unit = "day"))
       temp_segs <- readnii(paste0(scan_dirs[scan_idx], "_all_fast_origsegs.nii.gz"))
       seg_vols <- fslstats(
         paste0(scan_dirs[scan_idx], "_all_fast_origsegs.nii.gz"),

### 2025-03-18 code/adni_mri_preprocess.R
add scan_id to adni_info

@@ -502,6 +502,7 @@ for (dir_idx in 1:length(processed_dirs)) {
 
       scan_info <- scan_info |>
         mutate(
+          scan_id = scan_id,
           lhipp_vol = temp_lhipp_vol,
           rhipp_vol = temp_rhipp_vol,
           lthal_vol = temp_lthal_vol,

### 2025-03-20 code/adni_mri_preprocess.R
attempt to ignore errors

@@ -27,8 +27,8 @@ ncores <- parallel::detectCores() - 1
 cl <- parallel::makeCluster(ncores, type = "FORK")
 doParallel::registerDoParallel(cl = cl)
 
-# foreach (dir_idx = 1:length(sub_dirs)) %dopar% {
-foreach(dir_idx = 1:8) %dopar% {
+foreach (dir_idx = 1:length(sub_dirs), .errorhandling = "pass") %dopar% {
+# foreach(dir_idx = 1:64) %dopar% {
   img_dirs <- list.files(sub_dirs[dir_idx], recursive = TRUE, full.names = TRUE) %>%
     gsub(pattern = "([^/]+$)", replacement = "") %>%
     unique()

### 2025-03-20 code/adni_mri_preprocess.R
fix ouptut path

@@ -9,7 +9,7 @@ library(tidyverse)
 library(lubridate)
 # library(ANTsRCore)
 
-sub_dirs <- list.dirs("data/adni", recursive = FALSE)
+sub_dirs <- list.dirs("data/ADNI", recursive = FALSE)
 
 nii_scans <- list()
 
@@ -27,8 +27,7 @@ ncores <- parallel::detectCores() - 1
 cl <- parallel::makeCluster(ncores, type = "FORK")
 doParallel::registerDoParallel(cl = cl)
 
-# foreach (dir_idx = 1:length(sub_dirs)) %dopar% {
-foreach(dir_idx = 1:8) %dopar% {
+foreach (dir_idx = 1:length(sub_dirs)) %dopar% {
   img_dirs <- list.files(sub_dirs[dir_idx], recursive = TRUE, full.names = TRUE) %>%
     gsub(pattern = "([^/]+$)", replacement = "") %>%
     unique()
@@ -36,7 +35,7 @@ foreach(dir_idx = 1:8) %dopar% {
   nii_scans[[dir_idx]] <- list()
 
   for (img_idx in 1:length(img_dirs)) {
-    proc_dir <- paste0("data/adni_processed_fsl", gsub(pattern = "data/adni", replacement = "", img_dirs[img_idx]))
+    proc_dir <- paste0("data/adni_processed_fsl", gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx]))
     if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
       next
     } else {

### 2025-03-20 code/adni_mri_preprocess.R
fix changes

@@ -1,0 +1,528 @@
+library(oro.dicom)
+library(oro.nifti)
+library(neurobase)
+library(fslr)
+# library(extrantsr)
+library(stringr)
+library(foreach)
+library(tidyverse)
+library(lubridate)
+# library(ANTsRCore)
+
+sub_dirs <- list.dirs("data/adni", recursive = FALSE)
+
+nii_scans <- list()
+
+processed_scans <- list()
+
+hipp_mask <- readNIfTI(
+  paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm_Hipp_mask_dil8.nii.gz")
+)
+
+mni_nifti <- readNIfTI(
+  paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm.nii.gz")
+)
+
+ncores <- parallel::detectCores() - 1
+cl <- parallel::makeCluster(ncores, type = "FORK")
+doParallel::registerDoParallel(cl = cl)
+
+foreach (dir_idx = 1:length(sub_dirs), .errorhandling = "pass") %dopar% {
+# foreach(dir_idx = 1:64) %dopar% {
+  img_dirs <- list.files(sub_dirs[dir_idx], recursive = TRUE, full.names = TRUE) %>%
+    gsub(pattern = "([^/]+$)", replacement = "") %>%
+    unique()
+
+  nii_scans[[dir_idx]] <- list()
+
+  for (img_idx in 1:length(img_dirs)) {
+    proc_dir <- paste0("data/adni_processed_fsl", gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx]))
+    if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
+      next
+    } else {
+      all_slices <- readDICOM(img_dirs[img_idx])
+      nii <- dicom2nifti(all_slices)
+
+      run_first_all(
+        nii,
+        oprefix = proc_dir,
+        verbose = FALSE,
+        opts = "-d"
+      )
+    }
+  }
+}
+
+parallel::stopCluster(cl = cl)
+
+test_dicom <- readDICOM(
+  list.files(sub_dirs[1], recursive = TRUE, full.names = TRUE)[1]
+)
+
+adni_info <- read_csv("data/adni_info.csv")
+adni_info <- adni_info  |> 
+  rename(
+    image_id = "Image Data ID",
+    subid = "Subject",
+    date = "Acq Date"
[…]

### 2025-03-20 code/functions/fit_weighted_spline.R
add lambda parameter

@@ -1,4 +1,4 @@
-fit_weighted_spline <- function(x, params, weights, smoothing_vals, folds) {
+fit_weighted_spline <- function(x, params, weights, smoothing_vals, folds, lambda = exp(-20:5)) {
   k <- length(unique(folds))
   coefs <- list()
   d <- ncol(params)

### 2025-03-20 code/mnist_pme_test.R
remove ggimage dependency

@@ -2,7 +2,7 @@ library(doParallel)
 library(foreach)
 # library(pme)
 library(ggplot2)
-library(ggimage)
+# library(ggimage)
 library(imager)
 library(progress)
 library(Rfast)

### 2025-03-20 code/adni_test.R
attempt hierarchical model for adni

@@ -1,0 +1,460 @@
+library(doParallel)
+library(foreach)
+library(pme)
+library(progress)
+library(Rfast)
+library(tidyverse)
+
+source("code/functions/fit_weighted_spline.R")
+source("code/functions/print_SSD.R")
+source("code/functions/fit_adni_additive_model.R")
+source("code/functions/plot_id_pred.R")
+source("code/functions/calculate_pme_reconstructions.R")
+
+SSD_ratio_threshold <- 5
+verbose <- TRUE
+
+lhipp_surface <- read_csv("data/adni_fsl_lhipp_surface.csv")
+adni_info <- read_csv("data/adni_fsl_info.csv")
+
+lhipp_surface <- lhipp_surface |>
+  mutate(
+    date = decimal_date(date),
+    Group = gsub("EMCI", "MCI", Group),
+    Group = gsub("LMCI", "MCI", Group),
+  )
+
+lhipp_surface_centers <- lhipp_surface |>
+  group_by(subid, date, scan_id) |>
+  summarize(
+    mean_x = mean(x),
+    mean_y = mean(y),
+    mean_z = mean(z),
+    max_x = max(abs(x - mean_x)),
+    max_y = max(abs(y - mean_y)),
+    max_z = max(abs(z - mean_z))
+  )
+
+lhipp_bl <- lhipp_surface |>
+  group_by(subid) |>
+  arrange(date) |>
+  summarize(
+    date_bl = first(date),
+    max_date = max(date)
+  ) |>
+  mutate(duration = max_date - date_bl)
+
+lhipp_surface <- lhipp_surface |>
+  full_join(lhipp_surface_centers, by = c("subid", "date", "scan_id")) |>
+  full_join(lhipp_bl, by = "subid") |>
+  mutate(
+    x = (x - mean_x) / max_x,
+    y = (y - mean_y) / max_y,
+    z = (z - mean_z) / max_z,
+    time_from_bl = date - date_bl,
+    partition = ifelse(x > 0, 1, 2)
+  )
+
+lhipp_surface_inputs <- lhipp_surface |>
+  select(date, x, y, z) |>
+  as.matrix()
+
+lhipp_groups <- as.factor(lhipp_surface$Group)
+lhipp_ids <- as.factor(lhipp_surface$subid)
[…]

### 2025-03-20 code/functions/fit_adni_additive_model.R
specify adni-specific additive model

@@ -1,0 +1,424 @@
+fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, scans, times, epsilon = 0.05, max_iter = 100) {
+  D <- ncol(x)
+  d <- ncol(params)
+
+  groups <- as.factor(groups)
+  ids = as.factor(ids)
+  scans <- as.factor(scans)
+
+
+  fold_vec_full <- rep(1:k, ceiling(length(unique(scans)) / k))
+  scan_folds <- fold_vec_full[
+    sample(
+      1:length(unique(scans)),
+      length(unique(scans)),
+      replace = FALSE
+    )
+  ]
+
+  print("Initializing population model")
+  init_population_embedding <- fit_weighted_spline(
+    x,
+    params,
+    weights,
+    lambda,
+    scan_folds[scans]
+  )
+
+  print("Initializing Group models")
+  init_group_embeddings <- list()
+  group_x <- list()
+  group_params <- list()
+  group_weights <- list()
+  group_coefs <- list()
+
+  for (group_idx in seq_along(unique(groups))) {
+    group_set <- groups == unique(groups)[group_idx]
+    group_x[[group_idx]] <- x[group_set,]
+    group_params[[group_idx]] <- params[group_set,] 
+    group_weights[[group_idx]] <- weights[group_set]
+    group_ids <- droplevels(ids[group_set])
+    if (length(unique(group_ids)) < k) {
+      temp_k <- length(unique(group_ids))
+    } else {
+      temp_k <- k
+    }
+    id_fold_vec <- rep(1:temp_k, ceiling(length(unique(group_ids)) / temp_k))
+    id_folds <- id_fold_vec[
+      sample(
+        unique(group_ids),
+        length(unique(group_ids)),
+        replace = FALSE
+      )
+    ]
+    group_preds <- map(
+      1:nrow(group_params[[group_idx]]),
+      ~ init_population_embedding$embedding_map(group_params[[group_idx]][.x, ])
+    ) %>%
+      reduce(rbind)
+    init_group_embeddings[[group_idx]] <- fit_weighted_spline(
+      group_x[[group_idx]] - group_preds,
+      group_params[[group_idx]],
+      group_weights[[group_idx]],
+      lambda,
+      id_folds[group_ids]
+    )
+  }
+
+  print("Initializing ID models")
+  init_id_embeddings <- list()
[…]

### 2025-04-07 code/adni_mri_preprocess.R
change temp directory

@@ -1,3 +1,8 @@
+newtemp <- "~/tmp/rcall"
+dir.create(newtemp, recursive = TRUE)
+Sys.setenv(TMPDIR = tools::file_path_as_absolute(newtemp))
+unlink(tempdir(), recursive = TRUE)
+
 library(doSNOW)
 library(oro.dicom)
 library(oro.nifti)
@@ -28,7 +33,7 @@ ncores <- parallel::detectCores() / 2
 cl <- parallel::makeCluster(ncores)
 registerDoSNOW(cl = cl)
 
-foreach (
+error_list <- foreach (
   dir_idx = 1:length(sub_dirs), 
   .packages = c("oro.dicom", "oro.nifti", "neurobase", "fslr", "magrittr"),
   .export = c("sub_dirs", "nii_scans"),
@@ -60,3 +65,12 @@ foreach (
 }
 
 parallel::stopCluster(cl = cl)
+
+run_include <- vector(mode = "logical", length = length(error_list))
+for (i in seq_along(error_list)) {
+  if (class(error_list[[i]]) == "numeric") {
+    run_include[i] <- TRUE
+  } else {
+    run_include[i] <- FALSE
+  }
+}

### 2025-04-07 code/adni_mri_preprocess.R
simplified processing script

@@ -15,11 +15,9 @@ library(tidyverse)
 library(lubridate)
 # library(ANTsRCore)
 
-sub_dirs <- list.dirs("data/ADNI", recursive = FALSE)
-
-nii_scans <- list()
-
-processed_scans <- list()
+img_dirs <- list.files("data/ADNI", recursive = TRUE, full.names = TRUE) %>%
+  gsub(pattern = "([^/]+$)", replacement = "") %>%
+  unique()
 
 hipp_mask <- readNIfTI(
   paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm_Hipp_mask_dil8.nii.gz")
@@ -34,43 +32,36 @@ cl <- parallel::makeCluster(ncores)
 registerDoSNOW(cl = cl)
 
 error_list <- foreach (
-  dir_idx = 1:length(sub_dirs), 
+  img_idx = seq_along(img_dirs), 
   .packages = c("oro.dicom", "oro.nifti", "neurobase", "fslr", "magrittr"),
-  .export = c("sub_dirs", "nii_scans"),
+  .export = c("img_dirs"),
   .errorhandling = "pass"
 ) %dopar% {
 # foreach(dir_idx = 1:64) %dopar% {
-  img_dirs <- list.files(sub_dirs[dir_idx], recursive = TRUE, full.names = TRUE) %>%
-    gsub(pattern = "([^/]+$)", replacement = "") %>%
-    unique()
+  proc_dir <- paste0(
+    "data/adni_processed_fsl", 
+    gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx])
+  )
+  if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
+    seg_error <- FALSE
+  } else {
+    all_slices <- readDICOM(img_dirs[img_idx])
+    nii <- dicom2nifti(all_slices)
 
-  nii_scans[[dir_idx]] <- list()
-
-  for (img_idx in 1:length(img_dirs)) {
-    proc_dir <- paste0("data/adni_processed_fsl", gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx]))
-    if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
-      next
-    } else {
-      all_slices <- readDICOM(img_dirs[img_idx])
-      nii <- dicom2nifti(all_slices)
-
-      run_first_all(
-        nii,
-        oprefix = proc_dir,
-        verbose = FALSE,
[…]

### 2025-04-08 code/adni_mri_preprocess.R
skip problematic images

@@ -15,6 +15,8 @@ library(tidyverse)
 library(lubridate)
 # library(ANTsRCore)
 
+skip_imgs <- c("I436253")
+
 img_dirs <- list.files("data/ADNI", recursive = TRUE, full.names = TRUE) %>%
   gsub(pattern = "([^/]+$)", replacement = "") %>%
   unique()
@@ -27,8 +29,8 @@ mni_nifti <- readNIfTI(
   paste0(fsl_dir(), "/data/standard/MNI152_T1_1mm.nii.gz")
 )
 
-ncores <- parallel::detectCores() / 2
-cl <- parallel::makeCluster(ncores)
+ncores <- parallel::detectCores()
+cl <- parallel::makeCluster(ncores / 2)
 registerDoSNOW(cl = cl)
 
 error_list <- foreach (
@@ -38,12 +40,15 @@ error_list <- foreach (
   .errorhandling = "pass"
 ) %dopar% {
 # foreach(dir_idx = 1:64) %dopar% {
+  img_val <- str_split(img_dirs[img_idx], "/")[[1]][6]
   proc_dir <- paste0(
     "data/adni_processed_fsl", 
     gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx])
   )
   if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
     seg_error <- FALSE
+  } else if (img_val %in% skip_imgs) {
+    seg_error <- TRUE
   } else {
     all_slices <- readDICOM(img_dirs[img_idx])
     nii <- dicom2nifti(all_slices)

### 2025-04-09 code/adni_mri_preprocess.R
use doFuture for parallelization

@@ -3,17 +3,19 @@ dir.create(newtemp, recursive = TRUE)
 Sys.setenv(TMPDIR = tools::file_path_as_absolute(newtemp))
 unlink(tempdir(), recursive = TRUE)
 
-library(doSNOW)
+library(doFuture)
+library(foreach)
+library(fslr)
 library(oro.dicom)
 library(oro.nifti)
 library(neurobase)
-library(fslr)
+library(progressr)
 # library(extrantsr)
 library(stringr)
-library(foreach)
 library(tidyverse)
 library(lubridate)
 # library(ANTsRCore)
+handlers(global = TRUE)
 
 skip_imgs <- c("I436253")
 
@@ -30,41 +32,47 @@ mni_nifti <- readNIfTI(
 )
 
 ncores <- parallel::detectCores()
-cl <- parallel::makeCluster(ncores / 2)
-registerDoSNOW(cl = cl)
+plan(multisession, workers = ncores / 2)
 
-error_list <- foreach (
-  img_idx = seq_along(img_dirs), 
-  .packages = c("oro.dicom", "oro.nifti", "neurobase", "fslr", "magrittr"),
-  .export = c("img_dirs"),
-  .errorhandling = "pass"
-) %dopar% {
+segment_imgs <- function(img_dirs) {
+  options(warn = 1)
+  # p <- progressor(along = img_dirs)
+  error_list <- foreach (
+      img_idx = seq_along(img_dirs), 
+      # .packages = c("oro.dicom", "oro.nifti", "neurobase", "fslr", "magrittr"),
+      # .export = c("img_dirs"),
+      .errorhandling = "pass"
+    ) %dofuture% {
 # foreach(dir_idx = 1:64) %dopar% {
-  img_val <- str_split(img_dirs[img_idx], "/")[[1]][6]
-  proc_dir <- paste0(
-    "data/adni_processed_fsl", 
-    gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx])
-  )
-  if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
-    seg_error <- FALSE
-  } else if (img_val %in% skip_imgs) {
-    seg_error <- TRUE
-  } else {
-    all_slices <- readDICOM(img_dirs[img_idx])
-    nii <- dicom2nifti(all_slices)
+      img_val <- str_split(img_dirs[img_idx], "/")[[1]][6]
+      print(img_idx)
[…]

### 2025-04-09 code/adni_mri_preprocess.R
use divest package for dicom conversion

@@ -3,11 +3,12 @@ dir.create(newtemp, recursive = TRUE)
 Sys.setenv(TMPDIR = tools::file_path_as_absolute(newtemp))
 unlink(tempdir(), recursive = TRUE)
 
+library(divest)
 library(doFuture)
 library(foreach)
 library(fslr)
-library(oro.dicom)
-library(oro.nifti)
+# library(oro.dicom)
+# library(oro.nifti)
 library(neurobase)
 library(progressr)
 # library(extrantsr)
@@ -35,8 +36,7 @@ ncores <- parallel::detectCores()
 plan(multisession, workers = ncores / 2)
 
 segment_imgs <- function(img_dirs) {
-  options(warn = 1)
-  # p <- progressor(along = img_dirs)
+  p <- progressor(along = img_dirs)
   error_list <- foreach (
       img_idx = seq_along(img_dirs), 
       # .packages = c("oro.dicom", "oro.nifti", "neurobase", "fslr", "magrittr"),
@@ -45,7 +45,6 @@ segment_imgs <- function(img_dirs) {
     ) %dofuture% {
 # foreach(dir_idx = 1:64) %dopar% {
       img_val <- str_split(img_dirs[img_idx], "/")[[1]][6]
-      print(img_idx)
       proc_dir <- paste0(
         "data/adni_processed_fsl", 
         gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx])
@@ -55,8 +54,9 @@ segment_imgs <- function(img_dirs) {
       } else if (img_val %in% skip_imgs) {
         seg_error <- TRUE
       } else {
-        all_slices <- readDICOM(img_dirs[img_idx])
-        nii <- dicom2nifti(all_slices)
+        # all_slices <- readDICOM(img_dirs[img_idx])
+        # nii <- dicom2nifti(all_slices)
+        nii <- readDicom(img_dirs[img_idx], interactive = FALSE, verbosity = -1)
 
         first_out <- run_first_all(
           nii,
@@ -64,7 +64,7 @@ segment_imgs <- function(img_dirs) {
           verbose = FALSE,
           opts = "-d"
         )
-#         p(sprintf("img: %s", img_val))
+        p(sprintf("img: %s", img_val))
         seg_error <- is.null(first_out$segmentation)
       }
       seg_error

### 2025-04-09 paper/main.tex
Updates from Overleaf

@@
-%\documentclass[12pt]{amsart}
-\documentclass[11pt,reqno]{article}
-%\usepackage{cases}
-
-%\RequirePackage[numbers]{natbib}
-%\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
-\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
-\RequirePackage{graphicx}%% uncomment this for including figures
-
-\usepackage[
-  backend=biber,
-  style=authoryear
-]{biblatex}
-
-\addbibresource{references.bib}
-
-\usepackage{setspace}
+\documentclass[12pt]{article}
+\usepackage{algorithm2e}
+\usepackage{amscd}
+\usepackage{amsfonts}
 \usepackage{amsmath}
 \usepackage{amssymb}
-\usepackage{amscd}
 \usepackage{amsthm}
-\usepackage{amsfonts}
+\usepackage{authblk}
+\usepackage[english]{babel}
+\usepackage{bm}
+\usepackage{caption}
+\usepackage{color}
+\usepackage{float}
+\usepackage[perpage, symbol*]{footmisc}
+\usepackage{graphicx}
+\usepackage{hyperref}
+\usepackage{enumerate}
 \usepackage{mathrsfs}
-\usepackage{graphicx}
+\usepackage{natbib}
+\usepackage{setspace}
+\usepackage{subcaption}
+\usepackage{tikz}
+\usepackage{url}
+
+\newcommand{\blind}{1}
 \graphicspath{{figures/}}
-\usepackage[perpage,symbol*]{footmisc}
-\usepackage{float}
-\usepackage{hyperref}
-\usepackage{color}  
-\usepackage{tikz}
-\usepackage{caption}
-\usepackage{subcaption}
 
-\usepackage{algorithm2e}
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-1in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.7in}%
+\addtolength{\topmargin}{-1in}%
 
-%\usepackage{natbib}
-%\bibliographystyle{abbrvnat}
-%\setcitestyle{authoryear,open={(},close={)}}
-\usepackage{authblk}
-
-\usepackage{csquotes}
-\usepackage[english]{babel}
-
-\renewcommand{\baselinestretch}{1.0}
-\setlength{\oddsidemargin}{-0.5cm}
-\setlength{\evensidemargin}{-0.5cm}
-\renewcommand{\topmargin}{-2cm}
-\renewcommand{\oddsidemargin}{0mm}
-\renewcommand{\evensidemargin}{0mm}
-\renewcommand{\textwidth}{180mm}
-\renewcommand{\textheight}{240mm}
+\pdfminorversion=4
 
@@
 \newcommand{\T}{\intercal}
-\newcommand{\commentout}[1]{}
-
-\newcommand{\kmedit}[1]{{\color{purple}  #1}}
-\newcommand{\meng}[1]{{\color{purple} \sf $\clubsuit\clubsuit\clubsuit$ Kun Meng: [#1]}}
-\newcommand{\Meng}[1]{\margMa{(Kun Meng) #1}}
-
-\newcommand{\zielinski}[1]{{\color{blue} \sf $\spadesuit\spadesuit\spadesuit$ Rob Zielinski: [#1]}}
 
@@
 
+\setcitestyle{aysep={}}
+
 \begin{document}
 
-\title{Hierarchical Principal Manifold Estimation}
-\author[1]{Robert Zielinski}
-\author[2]{Kun Meng}
-\author[1]{Ani Eloyan}
-\affil[1]{Department of Biostatistics, Brown University}
-\affil[2]{Division of Applied Mathematics, Brown University}
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if1\blind
+{
+  \title{\bf A Principal Manifold-based Framework for Comparisons of Group-Level Manifold Estimates}
+  \author{
+    Robert Zielinski\hspace{.2cm}\\
+    Department of Biostatistics, Brown University\\
+    Kun Meng \\
+    Division of Applied Mathematics, Brown University\\
+    and \\
+    Ani Eloyan \\
+    Department of Biostatistics, Brown University\\
+    for the Alzheimer's Disease Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu).
+As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report.
+A complete listing of ADNI investigators can be found at \href{http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf}{\texttt{http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}}}
   \maketitle
+} \fi
 
-\doublespacing
+\if0\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf A Principal Manifold-based Framework for Comparisons of Group-Level Manifold Estimates}
+\end{center}
+  \medskip
+} \fi
 
-\section*{Abstract} \label{s:abstract}
+\bigskip
+\begin{abstract}
+Structural magnetic resonance imaging (MRI) data is frequently used to detect and monitor the progression of neurodegenerative conditions, such as Alzheimer's disease (AD). While 
+\end{abstract}
 
+\noindent%
+{\it Keywords:}  Alzheimer's disease, smoothing spline, magnetic resonance imaging
+\vfill
+
+\newpage
+\spacingset{1.9}
 \section{Introduction} \label{s:hpme_intro}
@@
 \begin{equation}
-  \mathbf{x}_{jq} = \psi(\mathbf{y}_{jq}, t_j) = f_p(\mathbf{y}_{jq}, t_{j}) + f_{g(j)}(\mathbf{y}_{jq}, t_{j}) + f_{i(j)}(\mathbf{y}_{jq}, t_{j}) + f_j(\mathbf{y}_{jq}) + \epsilon_{jq}. \label{eq:2}
+  \mathbf{x}_{jq} = \psi(\mathbf{y}_{jq}, t_j) = f_p(\mathbf{y}_{jq}, t_{j}) + f_{g(j)}(\mathbf{y}_{jq}, t_{j}) + f_{i(j)}(\mathbf{y}_{jq}, t_{j}) + f_j(\mathbf{y}_{jq}) + \epsilon_{jq}j. \label{eq:2}
 \end{equation}
@@
 
-\nocite{*}
-%\bibliographystyle{plain}
-%\bibliography{references}
-\printbibliography
+\bibliographystyle{jasa}
+\bibliography{references}
 

### 2025-04-09 code/adni_mri_preprocess.R
update images giving errors

@@ -18,7 +18,7 @@ library(lubridate)
 # library(ANTsRCore)
 handlers(global = TRUE)
 
-skip_imgs <- c("I436253")
+skip_imgs <- c("I436253", "I59843")
 
 img_dirs <- list.files("data/ADNI", recursive = TRUE, full.names = TRUE) %>%
   gsub(pattern = "([^/]+$)", replacement = "") %>%
@@ -33,7 +33,7 @@ mni_nifti <- readNIfTI(
 )
 
 ncores <- parallel::detectCores()
-plan(multisession, workers = ncores / 2)
+plan(cluster, workers = ncores / 2)
 
 segment_imgs <- function(img_dirs) {
   p <- progressor(along = img_dirs)

### 2025-04-11 code/adni_mri_preprocess.R
try to avoid memory isues

@@ -37,44 +37,56 @@ plan(multisession, workers = ncores / 2)
 
 segment_imgs <- function(img_dirs) {
   p <- progressor(along = img_dirs)
-  error_list <- foreach (
-      img_idx = seq_along(img_dirs), 
-      # .packages = c("oro.dicom", "oro.nifti", "neurobase", "fslr", "magrittr"),
-      # .export = c("img_dirs"),
-      .errorhandling = "pass"
-    ) %dofuture% {
+  foreach (
+    img_idx = seq_along(img_dirs), 
+    # .packages = c("oro.dicom", "oro.nifti", "neurobase", "fslr", "magrittr"),
+    # .export = c("img_dirs"),
+    .errorhandling = "pass"
+  ) %dofuture% {
 # foreach(dir_idx = 1:64) %dopar% {
-      img_val <- str_split(img_dirs[img_idx], "/")[[1]][6]
-      proc_dir <- paste0(
-        "data/adni_processed_fsl", 
-        gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx])
-      )
-      if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
-        seg_error <- FALSE
-      } else if (img_val %in% skip_imgs) {
-        seg_error <- TRUE
-      } else {
-        # all_slices <- readDICOM(img_dirs[img_idx])
-        # nii <- dicom2nifti(all_slices)
-        nii <- readDicom(img_dirs[img_idx], interactive = FALSE, verbosity = -1)
+    img_val <- str_split(img_dirs[img_idx], "/")[[1]][6]
+    proc_dir <- paste0(
+      "data/adni_processed_fsl", 
+      gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx])
+    )
+    if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
+    } else if (img_val %in% skip_imgs) {
+    } else {
+      # all_slices <- readDICOM(img_dirs[img_idx])
+      # nii <- dicom2nifti(all_slices)
+      nii <- readDicom(img_dirs[img_idx], interactive = FALSE, verbosity = -1)
 
-        first_out <- run_first_all(
-          nii,
-          oprefix = proc_dir,
-          verbose = FALSE,
-          opts = "-d"
-        )
-        p(sprintf("img: %s", img_val))
[…]

### 2025-04-11 code/functions/fit_weighted_spline.R
parallelize cross-validation

@@ -1,4 +1,5 @@
-fit_weighted_spline <- function(x, params, weights, smoothing_vals, folds) {
+fit_weighted_spline <- function(x, params, weights, smoothing_vals, folds, parallel = "sequential") {
+  require(future, quietly = TRUE)
   require(pme, quietly = TRUE)
   require(purrr, quietly = TRUE)
   k <- length(unique(folds))
@@ -7,43 +8,58 @@ fit_weighted_spline <- function(x, params, weights, smoothing_vals, folds) {
   smoothing_error <- vector()
   lambda <- smoothing_vals
 
+  cores <- availableCores()
+
   for (smoothing_idx in seq_along(smoothing_vals)) {
     coefs[[smoothing_idx]] <- list()
-    fold_errors <- vector()
-    for (k_idx in 1:k) {
-      train_params <- params[!(folds == k_idx), ] %>%
-        matrix(ncol = d)
-      fold_params <- params[folds == k_idx, ] %>%
-        matrix(ncol = d)
-      train_param_mat <- cbind(1, train_params)
-      train_x <- x[!(folds == k_idx), ]
-      fold_x <- x[folds == k_idx, ]
-      train_weights <- diag(weights[!(folds == k_idx)])
-      train_E <- calcE(train_params, 4 - ncol(train_params))
-      coefs[[smoothing_idx]][[k_idx]] <- solve_weighted_spline(
-        train_E,
-        train_weights,
-        train_param_mat,
-        train_x,
-        lambda[smoothing_idx],
-        ncol(train_params),
-        ncol(train_x)
-      )
-      fold_embedding <- function(parameters) {
-        as.vector(
-          (t(coefs[[smoothing_idx]][[k_idx]][seq_len(nrow(train_x)), ]) %*% etaFunc(parameters, train_params, 4 - d)) +
-            (t(coefs[[smoothing_idx]][[k_idx]][(nrow(train_x) + 1):(nrow(train_x) + d + 1), ]) %*% matrix(c(1, parameters), ncol = 1))
-        )
-      }
-      fold_errors[k_idx] <- map(
-        seq_len(nrow(fold_x)),
-        ~ weights[folds == k_idx][.x] * dist_euclidean(
-          fold_x[.x, ], fold_embedding(fold_params[.x, ])
-        )^2
-      ) %>%
-        reduce(c) %>%
-        mean()
+
+    if (parallel == "sequential") {
+      plan(sequential)
+    } else if (parallel == "multisession") {
[…]

### 2025-04-11 code/functions/fit_adni_additive_model.R
parallelize

@@ -1,4 +1,5 @@
-fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, scans, times, epsilon = 0.05, max_iter = 100) {
+fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, scans, times, epsilon = 0.05, max_iter = 100, cores = 1) {
+  require(future, quietly = TRUE)
   require(pme, quietly = TRUE)
   D <- ncol(x)
   d <- ncol(params)
@@ -6,7 +7,6 @@ fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, 
   groups <- as.factor(groups)
   ids = as.factor(ids)
   scans <- as.factor(scans)
-
 
   fold_vec_full <- rep(1:k, ceiling(length(unique(scans)) / k))
   scan_folds <- fold_vec_full[
@@ -17,13 +17,14 @@ fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, 
     )
   ]
 
-  print("Initializing population model")
+  print("Initializing Population model")
   init_population_embedding <- fit_weighted_spline(
     x,
     params,
     weights,
     lambda,
-    scan_folds[scans]
+    scan_folds[scans],
+    parallel = "multicore"
   )
 
   print("Initializing Group models")
@@ -31,39 +32,58 @@ fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, 
   group_x <- list()
   group_params <- list()
   group_weights <- list()
-  group_coefs <- list()
+
+  avail_cores <- min(length(unique(groups)), cores - k)
+  plan(multisession, workers = avail_cores)
+  group_out <- list()
+  for (group_idx in seq_along(unique(groups))) {
+    group_out[[group_idx]] <- future({
+      group_set <- groups == unique(groups)[group_idx]
+      group_x <- x[group_set, ]
+      group_params <- params[group_set, ]
+      group_weights <- weights[group_set]
+      group_ids <- droplevels(ids[group_set])
+      if (length(unique(group_ids)) < k) {
+        temp_k <- length(unique(group_ids))
+      } else {
+        temp_k <- k
+      }
+      id_fold_vec <- rep(1:temp_k, ceiling(length(unique(group_ids)) / temp_k))
+      id_folds <- id_fold_vec[
+        sample(
+          unique(group_ids),
[…]

### 2025-04-11 code/functions/calculate_lpme_reconstructions.R
update helper functions

@@ -1,0 +1,76 @@
+calculate_lpme_reconstructions <- function(lpme_result, data) {
+  #' Calculate LPME Reconstructions
+  #'
+  #' Uses a fitted LPME model to estimate reconstructed observations in D-dimensional space
+  #'
+  #' @param lpme_result The fitted LPME model
+  #' @param data A numeric matrix including a column of time values and d-dimensional parameters
+
+  # obtain lpme run with optimal tuning value and
+  # parameter values for the associated cluster centers
+  lpme_opt <- which.min(lpme_result$msd)
+  params <- lpme_result$parameterization_list[[lpme_opt]]
+
+  # convert the cluster center parameters to D dimensions
+  centers <- map(
+    1:nrow(params),
+    ~ lpme_result$embedding_map(params[.x, ])
+  ) |>
+    reduce(rbind)
+
+  # within the clusters calculated for each time point, find the nearest
+  # cluster center for each observation.
+  # adjustments are needed to compare to indices among full set of clusters.
+  nearest_center <- map(
+    1:nrow(data),
+    ~ which.min(
+      apply(
+        centers[centers[, 1] == data[.x, 1], ],
+        1,
+        dist_euclidean,
+        y = data[.x, ]
+      )
+    )
+  ) |>
+    reduce(c)
+  center_idx_adjustment <- map(
+    1:length(nearest_center),
+    ~ nrow(
+      centers[centers[, 1] < data[.x, 1], ]
+    )
+  ) |>
+    reduce(c)
+  nearest_center <- nearest_center + center_idx_adjustment
+
+  # obtain the parameters associated with each nearest_cluster
+  nearest_params <- map(
+    1:length(nearest_center),
+    ~ params[nearest_center[.x], ]
+  ) |>
+    reduce(rbind)
+
+  # find d-dimensional projections to the estimated manifold for each observation
+  lpme_projections <- map(
+    1:nrow(data),
+    ~ projection_lpme(
+      data[.x, ],
+      function(x) lpme_result$embedding_map(x),
+      nearest_params[.x, ]
+    )
+  ) |>
+    reduce(rbind)
+
+  # embed projections in D-dimensional space to calculate reconstructions
+  lpme_reconstructions <- map(
+    1:nrow(data),
+    ~ lpme_result$embedding_map(lpme_projections[.x, ])
+  ) |>
+    reduce(rbind)
+
+  return(
+    list(
+    reconstructions = lpme_reconstructions,
+    projections = lpme_projections
+  )
+  )
+}

### 2025-04-11 code/functions/calculate_pme_reconstructions.R

@@ -1,0 +1,67 @@
+calculate_pme_reconstructions <- function(pme_result, data) {
+  #' Calculate PME Reconstructions
+  #'
+  #' Uses a fitted PME model to estimate reconstructed observations in D-dimensional space
+  #'
+  #' @param lpme_result The fitted PME model
+  #' @param data A numeric matrix of d-dimensional parameters, with time values excluded
+
+  require(purrr)
+
+  # obtain pme run with optimal tuning value and
+  # parameter values for the associated cluster centers
+  pme_opt <- which.min(pme_result$MSD)
+  params <- pme_result$parameterization[[pme_opt]]
+
+  centers <- map(
+    seq_len(nrow(params)),
+    ~ pme_result$embedding_map(params[.x, ])
+  ) |>
+    reduce(rbind)
+
+  # calculate the cluster centers nearest to each observation
+  nearest_center <- map(
+    seq_len(nrow(data)),
+    ~ which.min(
+      apply(
+        centers,
+        1,
+        dist_euclidean,
+        y = data[.x, ]
+      )
+    )
+  ) |>
+    reduce(c)
+
+  # obtain the parameters associated with each nearest_cluster
+  nearest_params <- map(
+    seq_along(nearest_center),
+    ~ params[nearest_center[.x], ]
+  ) |>
+    reduce(rbind)
+
+  # find d-dimensional projections to the estimated manifold for each observation
+  pme_projections <- map(
+    seq_len(nrow(data)),
+    ~ projection_pme(
+      data[.x, ],
+      function(x) pme_result$embedding_map(x),
+      nearest_params[.x, ]
+    )
+  ) |>
+    reduce(rbind)
+
+  # embed projections in D-dimensional space to calculate reconstructions
+  pme_reconstructions <- map(
+    seq_len(nrow(data)),
+    ~ pme_result$embedding_map(pme_projections[.x, ])
+  ) |>
+    reduce(rbind)
+
+  return(
+    list(
+      reconstructions = pme_reconstructions, 
+      projections = pme_projections
+    )
+  )
+}

### 2025-04-11 code/functions/fit_adni_additive_model.R

@@ -24,7 +24,7 @@ fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, 
     weights,
     lambda,
     scan_folds[scans],
-    parallel = "multicore"
+    parallel = "multisession"
   )
 
   print("Initializing Group models")
@@ -67,7 +67,7 @@ fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, 
         group_weights,
         lambda,
         id_folds[group_ids],
-        parallel = "multicore"
+        parallel = "multisession"
       )
 
       list(
@@ -282,7 +282,7 @@ fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, 
       weights,
       lambda,
       scan_folds[scans],
-      parallel = "multicore"
+      parallel = "multisession"
     )
 
     avail_cores <- min(length(unique(groups)), cores - k)
@@ -319,7 +319,7 @@ fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, 
           group_weights,
           lambda,
           id_folds[group_ids],
-          parallel = "multicore"
+          parallel = "multisession"
         )
 
         list(
@@ -387,7 +387,7 @@ fit_adni_additive_model <- function(x, params, weights, lambda, k, groups, ids, 
           id_weights,
           lambda,
           scan_fold_vals,
-          parallel = "multicore"
+          parallel = "multisession"
         )
 
         list(

### 2025-04-11 code/adni_test.R
add parallelization

@@ -1,9 +1,13 @@
 library(doParallel)
+library(doFuture)
 library(foreach)
+library(listenv)
 library(pme)
 library(progress)
 library(Rfast)
 library(tidyverse)
+
+cores <- detectCores()
 
 source("code/functions/fit_weighted_spline.R")
 source("code/functions/print_SSD.R")
@@ -64,7 +68,6 @@ lhipp_ids <- as.factor(lhipp_surface$subid)
 lhipp_scans <- as.factor(lhipp_surface$image_id)
 lhipp_partition <- lhipp_surface$partition
 
-lhipp_init_pme <- list()
 lhipp_init_data <- list()
 lhipp_init_data[[1]] <- lhipp_surface_inputs[
   (lhipp_groups == lhipp_groups[1]) & 
@@ -76,93 +79,112 @@ lhipp_init_data[[2]] <- lhipp_surface_inputs[
   (lhipp_ids == lhipp_ids[1]) & 
   (lhipp_scans == lhipp_scans[1]) & 
   (lhipp_partition == 2), ]
- 
-lhipp_init_pme[[1]] <- pme(
-  lhipp_init_data[[1]][, -1],
-  d = 2, 
-  print_plots = FALSE, 
-  verbose = TRUE
-)
-lhipp_init_pme[[2]] <- pme(
-  lhipp_init_data[[2]][, -1],
-  d = 2, 
-  print_plots = FALSE, 
-  verbose = TRUE
+
+plan(multicore, workers = 2)
+init_pme_wrapper <- list()
+for (i in seq_len(2)) {
+  init_pme_wrapper[[i]] <- future({
+    pme(
+      lhipp_init_data[[i]][, -1],
+      d = 2, 
+      print_plots = FALSE, 
+      verbose = FALSE 
+    )
+  }, seed = TRUE)
+}
+
+lhipp_init_pme <- map(
+  init_pme_wrapper,
+  ~ value(.x)
 )
 
-group_vals <- unique(lhipp_groups)
-id_vals <- unique(lhipp_ids)
+plan(multicore, workers = cores - 1)
+lhipp_reduced <- list()
+scan_list <- unique(lhipp_scans)
+for (scan_idx in seq_along(scan_list)) {
+  lhipp_reduced[[scan_idx]] <- future({
+    scan_indices <- lhipp_scans == scan_list[scan_idx]
[…]

### 2025-04-11 code/adni_test.R
remove extra )

@@ -487,9 +487,8 @@ lhipp_projections_pop <- map(
   seq_len(nrow(lhipp_surface_inputs)),
   ~ additive_model[[lhipp_partition[.x]]]$population_embedding$embedding_map(lhipp_params[.x, ]),
   .progress = TRUE
-  ) |>
+) |>
   reduce(rbind)
-)
 
 hpme_msd <- map(
   seq_len(nrow(lhipp_surface_inputs)),

### 2025-04-11 code/mesh_plotting_example.qmd
add mesh example

---
title: "plotly mesh plotting example"
author: "Rob Zielinski"
date: "04/07/2025"
---

```{r}
library(alphashape3d)
library(doParallel)
library(foreach)
library(Morpho)
library(plotly)
library(pme)
library(progress)
library(rgl)
library(Rfast)
library(tidyverse)

source("code/functions/fit_weighted_spline.R")
source("code/functions/print_SSD.R")
source("code/functions/fit_adni_additive_model.R")
source("code/functions/plot_id_pred.R")
source("code/functions/calculate_pme_reconstructions.R")

SSD_ratio_threshold <- 5
verbose <- TRUE

lhipp_surface <- read_csv("data_old/adni_fsl_lhipp_surface.csv")
adni_info <- read_csv("data_old/adni_fsl_info.csv")

lhipp_surface <- lhipp_surface |>
  mutate(
    date = decimal_date(date),
    Group = gsub("EMCI", "MCI", Group),
    Group = gsub("LMCI", "MCI", Group),
  )

lhipp_surface_centers <- lhipp_surface |>
  group_by(subid, date, scan_id) |>
  summarize(
    mean_x = mean(x),
    mean_y = mean(y),
    mean_z = mean(z),
    max_x = max(abs(x - mean_x)),
    max_y = max(abs(y - mean_y)),
    max_z = max(abs(z - mean_z))
  )

lhipp_bl <- lhipp_surface |>
  group_by(subid) |>
  arrange(date) |>
  summarize(
    date_bl = first(date),
    max_date = max(date)
  ) |>
  mutate(duration = max_date - date_bl)

lhipp_surface <- lhipp_surface |>
  full_join(lhipp_surface_centers, by = c("subid", "date", "scan_id")) |>
  full_join(lhipp_bl, by = "subid") |>
  mutate(
    x = (x - mean_x) / max_x,
    y = (y - mean_y) / max_y,
    z = (z - mean_z) / max_z,
    time_from_bl = date - date_bl,
    partition = ifelse(x > 0, 1, 2)
  )

lhipp_test <- lhipp_surface |>
  filter(image_id == "I13721") |>
  select(x, y, z) |>
  as.matrix()

ash <- ashape3d(
  lhipp_test,
  alpha = 0.25,
  pert = TRUE
)
test_mesh <- as.mesh3d(ash)
```

```{r}
plot_ly(
  x = test_mesh$vb[1, ],
  y = test_mesh$vb[2, ],
  z = test_mesh$vb[3, ],
  i = test_mesh$it[1, ] - 1,
  j = test_mesh$it[2, ] - 1,
  k = test_mesh$it[3, ] - 1,
  type = "mesh3d"
) |>
  layout(scene = list(aspectmode = "data"))
```

```{r}
plot_ly(
  x = lhipp_test[, 1],
  y = lhipp_test[, 2],
  z = lhipp_test[, 3],
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3)
)
```

```{r}
data(nose)##load data
##warp a mesh onto another landmark configuration:
longnose.mesh <- tps3d(shortnose.mesh, shortnose.lm, longnose.lm,threads=1)
## Not run: 
mD <- meshDist(longnose.mesh, shortnose.mesh)
##now change the color ramp
render(mD,rampcolors = c("white","red"))

## End(Not run)
#use unsigned distances and a ramp from blue to red
#color distances < 0.01 green:
## Not run: 
meshDist(
  longnose.mesh, 
  shortnose.mesh, 
  rampcolors = c("blue", "red"),
  sign=FALSE, 
  tol=0.5
)
```

### 2025-04-11 code/lpme_adni_application.R
add lpme example

@@ -1,0 +1,1899 @@
+library(foreach)
+library(lubridate)
+library(plotly)
+library(plot3D)
+library(pme)
+library(pracma)
+library(RColorBrewer)
+library(Rfast)
+library(tidyverse)
+
+source("code/functions/calc_pme_est.R")
+source("code/functions/calc_lpme_est_params.R")
+source("code/functions/estimate_volume.R")
+source("code/functions/interior_identification.R")
+source("code/functions/create_cross_section_matrix.R")
+
+
+
+lhipp_surface <- read_csv("data/adni_fsl_lhipp_surface.csv")
+lthal_surface <- read_csv("data/adni_fsl_lthal_surface.csv")
+rhipp_surface <- read_csv("data/adni_fsl_rhipp_surface.csv")
+rthal_surface <- read_csv("data/adni_fsl_rthal_surface.csv")
+
+hipp_info <- read_csv("data/adni_fsl_hipp_info.csv")
+thal_info <- read_csv("data/adni_fsl_thal_info.csv")
+
+est_hipp_info <- data.frame(
+  patno = character(),
+  date = numeric(),
+  lhipp_data_vol2 = numeric(),
+  lhipp_vol_lpme1 = numeric(),
+  lhipp_vol_lpme2 = numeric(),
+  lhipp_vol_pme1 = numeric(),
+  lhipp_vol_pme2 = numeric(),
+  rhipp_data_vol2 = numeric(),
+  rhipp_vol_lpme1 = numeric(),
+  rhipp_vol_lpme2 = numeric(),
+  rhipp_vol_pme1 = numeric(),
+  rhipp_vol_pme2 = numeric()
+)
+
+est_thal_info <- data.frame(
+  patno = character(),
+  date = numeric(),
+  lthal_data_vol2 = numeric(),
+  lthal_vol_lpme1 = numeric(),
+  lthal_vol_lpme2 = numeric(),
+  lthal_vol_pme1 = numeric(),
+  lthal_vol_pme2 = numeric(),
+  rthal_data_vol2 = numeric(),
+  rthal_vol_lpme1 = numeric(),
+  rthal_vol_lpme2 = numeric(),
+  rthal_vol_pme1 = numeric(),
+  rthal_vol_pme2 = numeric()
+)
+
+# process scan dates
+lhipp_surface <- lhipp_surface %>%
+  mutate(
+    scan_date = gsub("\\.0", "", scan_date),
+    scan_date = ymd_hms(scan_date),
[…]

### 2025-04-14 code/adni_test.R
udpate parallelization export limit

@@ -7,7 +7,8 @@ library(progress)
 library(Rfast)
 library(tidyverse)
 
-cores <- detectCores()
+options(future.globals.maxSize = 32 * 1024^3)
+cores <- detectCores() / 2
 
 source("code/functions/fit_weighted_spline.R")
 source("code/functions/print_SSD.R")

### 2025-04-14 code/adni_extract_surfaces.R
debug cases with multiple scans per visit

@@ -3,6 +3,7 @@ library(oro.nifti)
 library(neurobase)
 library(fslr)
 # library(extrantsr)
+library(progressr)
 library(stringr)
 library(foreach)
 library(tidyverse)
@@ -29,8 +30,8 @@ adni_info <- adni_info  |>
   ) |>
   arrange(subid, date)
 
-patnos <- list.dirs("data/adni_processed_fsl", recursive = FALSE, full.names = FALSE)
-processed_dirs <- list.dirs("data/adni_processed_fsl", recursive = FALSE)
+patnos <- list.dirs("data/adni_processed_fsl_pt", recursive = FALSE, full.names = FALSE)
+processed_dirs <- list.dirs("data/adni_processed_fsl_pt", recursive = FALSE)
 
 lhipp <- tibble(
   subid = character(),
@@ -155,16 +156,11 @@ image_info <- tibble(
 )
 
 
-mprage_names <- c("MP-RAGE", "MPRAGE", "MP-RAGE_REPEAT", "MPRAGE_REPEAT")
-for (dir_idx in 1:length(processed_dirs)) {
+mprage_names <- c("MP-RAGE", "MPRAGE", "MP-RAGE_REPEAT", "MPRAGE_REPEAT", "MPRAGE_Repeat")
+with_progress({
+for (dir_idx in seq_along(processed_dirs)) {
   print(patnos[dir_idx])
-  scan_dates <- list.dirs(
-    paste0(processed_dirs[dir_idx], "/", mprage_names),
-    full.names = FALSE,
-    recursive = FALSE
-  )
-  scan_dates <- ymd_hms(scan_dates)
-  # scan_dates <- floor_date(scan_dates, unit = "day")
+    # scan_dates <- floor_date(scan_dates, unit = "day")
 
   scan_dirs <- list.files(
     paste0(processed_dirs[dir_idx], "/", mprage_names),
@@ -173,17 +169,27 @@ for (dir_idx in 1:length(processed_dirs)) {
   ) %>%
     gsub(pattern = "([^/]+$)", replacement = "") %>%
     unique()
+
+  scan_dates <- map(
+    scan_dirs,
+    ~ str_split(.x, "/")[[1]][5]
+  ) |>
+    reduce(c)
+  scan_ids <- map(
+    scan_dirs,
+    ~ str_split(.x, "/")[[1]][6]
+  ) |>
+    reduce(c)
+  scan_dates <- ymd_hms(scan_dates)
+
   for (scan_idx in 1:length(scan_dirs)) {
[…]

### 2025-04-14 code/adni_mri_preprocess.R
add erroneous images

@@ -18,7 +18,7 @@ library(lubridate)
 # library(ANTsRCore)
 handlers(global = TRUE)
 
-skip_imgs <- c("I436253", "I59843")
+skip_imgs <- c("I436253", "I59843", "I272379", "I135611")
 
 img_dirs <- list.files("data/ADNI", recursive = TRUE, full.names = TRUE) %>%
   gsub(pattern = "([^/]+$)", replacement = "") %>%
@@ -50,7 +50,9 @@ segment_imgs <- function(img_dirs) {
       gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx])
     )
     if (file.exists(paste0(proc_dir, "/_all_fast_origsegs.nii.gz"))) {
+      p(sprintf("img: %s", img_val))
     } else if (img_val %in% skip_imgs) {
+      p(sprintf("img: %s", img_val))
     } else {
       # all_slices <- readDICOM(img_dirs[img_idx])
       # nii <- dicom2nifti(all_slices)
